### 2019-09-29 docs/paper/paper.tex
docs: Write introduction of paper

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\title{RL-Musician: A Tool for Music Composition with Deep Reinforcement Learning}

\date{September 30, 2019}

\author{
  Nikolay~Lysenko\\
  \texttt{nikolay.lysenko.1992@gmail.com} \\
}

\renewcommand{\headeright}{A draft}
\renewcommand{\undertitle}{A draft}

\begin{document}
\maketitle

\begin{abstract}
The notion of creativity is wider than generating something inspired by reference pieces of art. It also includes generation of something that meets criteria of being an art regardless of similar pieces existence. However, supervised machine learning is limited by the former scope, because it requires a dataset. Reinforcement learning can be used instead if the goal is to reveal more aspects of creativity. In this paper, a cross-entropy agent is trained to compose musical pieces by interacting with a piano roll environment. The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory.
\end{abstract}

\keywords{algorithmc composition \and music generation \and reinforcement learning}

\section{Introduction}
\label{sec:introduction}

Algorithmic music composition is automatic generation of outputs representing musical pieces and written in some formal notation. To name a few of common notations, there are sheet music, tablature, and MIDI standard. It is not required from output representation to unambiguously define sound waveform. For example, sheet music leaves exact loudness of played notes to discretion of performer and may include only imprecise hints like pianissimo (very quiet). Anyway, there are parameters of sounds that must be determined by their representation. Usually, such parameters are pitch, start time, and duration.

%Human decisions are replaced with randomness or pseudorandomness and an algorithm must turn random inputs into representation of a musical piece.

Plenty of various approaches for composing music automatically exist. In particular, there are approaches based on machine learning. Both supervised learning and reinforcement learning are applicable here. Moreover, some researchers combine them \cite{jaques2016generating}.

In this paper, a pure reinforcement learning approach to music generation is studied. The reasons for not involving supervised learning at all are as follows:
\begin{itemize}
	\item Finding new ways of music creation is a more challenging task than imitation of famous pieces. If no known pieces are used, chances are that the harder problem is considered and it is not replaced with the simpler problem of imitation.
	\item There are tuning systems other than equal temperament (for instance, in microtonal music). For some of them it may be impossible to collect dataset large enough to allow training models in a supervised fashion. However, developers of a tuning system should know some underlying principles and so (at least, in theory) it is possible to create evaluational rules and train an agent based on them.
\end{itemize}

Results reported at this draft version are far from using above advantages at full scale. The current study is rather a proof-of-concept for composition of music with reinforcement learning solely.

\section{Background and Related Work}
\label{sec:literature}

\subsection{Algorithmic Composition}
\label{subsec:composition}

Along with other options, algorithmic composition can be framed as sampling from a model that returns probabilities of next sound events (say, played notes) given current state. Sampling can be done greedily or with more sophisticated techniques like beam search. Although finding optimal sampling procedure from a given model is an important task, let us focus on creation of such models.

First of all, to create a model, its type and structure must be defined. Then it must be trained and so loss function, optimizer, and dataset must be introduced. In music generation, neural networks are often used as models, because they are universal approximators that can be trained with gradient descent. What is also common, training problem is usually set up as maximum likelihood estimation, i.e., the loss is cross-entropy between predicted distribution and distribution concentrated at target. Other details vary from paper to paper.

\subsection{Cross-Entropy Method}
\label{subsec:crossentropy}

\section{Methodology}
\label{sec:approach}

\section{Experimental Results}
\label{sec:results}

A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}. The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.1.1/}}.

%The implementation is non-demanding and transparent.

The implementation relies on some open-source tools \cite{brockman2016openai,chollet2015keras,oliphant2006guide,raffel2014intuitive,dong2018pypianoroll}.

\section{Further Improvements}
\label{sec:improvements}
% start from supervisedly pre-trained weigths;

\section{Conclusion}
\label{sec:conclusion}

\bibliographystyle{unsrt}  
\bibliography{references}

\end{document}

### 2019-09-30 docs/paper/paper.tex
docs: Revise abstract and introduction

@@
 The notion of creativity is wider than generating something inspired by reference pieces of art.
-It also includes generation of something that meets criteria of being an art regardless of similar pieces existence.
-However, supervised machine learning is limited by the former scope, because it requires a dataset.
-Reinforcement learning can be used instead if the goal is to reveal more aspects of creativity.
-In this paper, a cross-entropy agent is trained to compose musical pieces by interacting with a piano roll environment.
+It also includes generation of something that meets criteria of being an art, but is not backed by existing pieces.
+On the one hand, supervised machine learning is limited by the former scope, because it requires a dataset.
+Reinforcement learning can be applied instead if the goal is to reveal more aspects of creativity.
+On the other hand, numerous recent advances in music generation are achieved with supervised deep learning.
+In this paper, a method that uses both reinforcement learning and deep learning is developed for automated music composition.
+An agent with neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment.
 The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory.
+No dataset is involved, because cross-entropy method is used instead of gradient-based methods.
 \end{abstract}
@@
 It is not required from output representation to unambiguously define sound waveform.
-For example, sheet music leaves exact loudness of played notes to discretion of performer and may include only imprecise hints like pianissimo (very quiet).
+For example, sheet music leaves exact loudness of played notes to discretion of a performer and may include only imprecise hints like pianissimo (very quiet).
 Anyway, there are parameters of sounds that must be determined by their representation.
@@
 
-Plenty of various approaches for composing music automatically exist.
-In particular, there are approaches based on machine learning.
-Both supervised learning and reinforcement learning are applicable here.
+Currently, there are no perfect tools for algorithmic composition.
+Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that can not be distinguished from works of a talented human composer.
+Thus, there is an open research problem.
+Some recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating} and reinforcement learning \cite{smith2012reinforcement} are applicable.
 Moreover, some researchers combine them \cite{jaques2016generating}.
 
-In this paper, a pure reinforcement learning approach to music generation is studied. The reasons for not involving supervised learning at all are as follows:
+In this paper, a new approach to music composition is suggested.
+There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps.
+An agent observes current time step and some previous time steps.
+An action of the agent can be either playing a note at the current time step or moving to the next time step.
+When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition.
+Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions.
+These parameters are trained with cross-entropy method.
+
+Obviously, the approach belongs to reinforcement learning, because there is an environment, but a dataset is absent.
+Nevertheless, there are strong connections with supervised deep learning.
+From generation of new pieces point of view, actor model returns probabilities of next notes given current state.
+This is similar to sampling new sequences from a model trained with maximum likelihood method on a dataset of existing sequences.
+
+More details on the methodology are provided in Section \ref{sec:methodology}, but here it is appropriate to discuss its advantages.
+
+The reasons for not involving supervised training at all are as follows:
 \begin{itemize}
@@
 
-Results reported at this draft version are far from using above advantages at full scale. The current study is rather a proof-of-concept for composition of music with reinforcement learning solely.
+Further, sequential generation of composition by neural network implies that any well-studied architectures from sequence generation tasks can be plugged in and tested.
+The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or other methods (genetic algorithms, ant colony algorithms, evolutionary strategies \cite{salimans2017evolution}).
+
+Actually, results reported at this draft version are far from using above advantages at full scale.
+The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:improvements}.
+Also it is worth mentioning that the suggested approach is quite straightforward, but to the best of the author's knowledge it is not described anywhere.
 
@@
 \section{Methodology}
-\label{sec:approach}
+\label{sec:methodology}
+% dataset -> rules; backpropagation -> cross-entropy; learning rate -> weights std.
 
@@
 \label{sec:improvements}
+% better evaluational rules;
+% randomized initial observations;
+% evolutionary strategies instead of cross-entropy;
 % start from supervisedly pre-trained weigths;

### 2019-10-01 docs/paper/paper.tex
docs: Start writing about related work

@@
 
-%Human decisions are replaced with randomness or pseudorandomness and an algorithm must turn random inputs into representation of a musical piece.
-
 Currently, there are no perfect tools for algorithmic composition.
@@
 Thus, there is an open research problem.
-Some recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating} and reinforcement learning \cite{smith2012reinforcement} are applicable.
+Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating} and reinforcement learning \cite{smith2012reinforcement} are applicable.
 Moreover, some researchers combine them \cite{jaques2016generating}.
@@
 In this paper, a new approach to music composition is suggested.
+Although it is quite straightforward, to the best of the author's knowledge it is not described anywhere, so the first contribution is its rigorous definition.
+The second contribution\footnote{As of now, it is in progress.} is proper setup and tuning of the parameters.
+It is often the case that accurately tuned simple methods outperform complex methods \cite{dacrema2019are} and, in addition, they are more transparent and less demanding.
+
+Brief outline of the approach is as follows.
 There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps.
@@
 From generation of new pieces point of view, actor model returns probabilities of next notes given current state.
-This is similar to sampling new sequences from a model trained with maximum likelihood method on a dataset of existing sequences.
+This is similar to sampling new sequences from a next-step prediction model trained with maximum likelihood method on a dataset of existing sequences.
 
 More details on the methodology are provided in Section \ref{sec:methodology}, but here it is appropriate to discuss its advantages.
-
 The reasons for not involving supervised training at all are as follows:
@@
 The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:improvements}.
-Also it is worth mentioning that the suggested approach is quite straightforward, but to the best of the author's knowledge it is not described anywhere.
 
@@
 
-Along with other options, algorithmic composition can be framed as sampling from a model that returns probabilities of next sound events (say, played notes) given current state.
-Sampling can be done greedily or with more sophisticated techniques like beam search.
-Although finding optimal sampling procedure from a given model is an important task, let us focus on creation of such models.
+In algorithmic composition domain, numerous links between reinforcement learning and deep learning are established.
 
-First of all, to create a model, its type and structure must be defined.
-Then it must be trained and so loss function, optimizer, and dataset must be introduced.
-In music generation, neural networks are often used as models, because they are universal approximators that can be trained with gradient descent.
-What is also common, training problem is usually set up as maximum likelihood estimation, i.e., the loss is cross-entropy between predicted distribution and distribution concentrated at target.
-Other details vary from paper to paper.
+%Reinforcement learning can be used for
+
+Since algorithmic composition can be framed as training of a generative model, it sounds natural to try one of the most salient examples of generative models -- generative adversarial networks (GAN) \cite{goodfellow2014gan}.
+However, widespread notations for music assume sequences of discrete values but classical GAN work well only with continuous data, because gradient of discrete-valued functions is uninformative.
+Techniques originating from RL can be used to overcome this obstacle \cite{yu2016seqgan,hjelm2017boundary}.
+Namely, generator is trained with policy gradient method \cite{williams1992simple}.
+Such methodology is applied to various tasks including music composition \cite{limaguimaraes2017objective}.
 
@@
 
-%The implementation is non-demanding and transparent.
-
 The implementation relies on some open-source tools \cite{brockman2016openai,chollet2015keras,oliphant2006guide,raffel2014intuitive,dong2018pypianoroll}.
@@
 \label{sec:improvements}
-% better evaluational rules;
+% better evaluational rules (tonality);
 % randomized initial observations;
+% LSTM as actor model;
 % evolutionary strategies instead of cross-entropy;

### 2019-10-03 docs/paper/paper.tex
docs: Write about related work

@@
 Thus, there is an open research problem.
-Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating} and reinforcement learning \cite{smith2012reinforcement} are applicable.
-Moreover, some researchers combine them \cite{jaques2016generating}.
+Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{smith2012reinforcement} are applicable.
+Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}.
 
@@
 Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions.
-These parameters are trained with cross-entropy method.
+These parameters are trained with cross-entropy method (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).
 
@@
 Further, sequential generation of composition by neural network implies that any well-studied architectures from sequence generation tasks can be plugged in and tested.
-The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or other methods (genetic algorithms, ant colony algorithms, evolutionary strategies \cite{salimans2017evolution}).
+The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or alternative methods.
 
@@
 
-%Reinforcement learning can be used for
+Reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}.
+The goal is to make generated pieces more structured and conformed with music theory rules.
+To define an environment, let its state be composed of recurrent neural networks states and previously played notes, let an action be an output for current time step and let reward depend on both evaluational rules and probability of output according to initial RNN.
+Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred over cross-entropy method as a training algorithm.
+
+Usually, softmax activation function is used in the last layer of generative RNN.
+However, it is possible to consider a family of activation functions parametrized by one parameter $t$ such that output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$).
+Also, input vector can be extended by introducing an additional part indicating origination of initial input (so called plan).
+Then an environment can be defined so that state is a pair of $t$ and plan, actions are changes in either $t$ or plan, and reward depends on evaluation of produced with these settings piece \cite{kumar2019polyphonic}.
 
@@
 Namely, generator is trained with policy gradient method \cite{williams1992simple}.
-Such methodology is applied to various tasks including music composition \cite{limaguimaraes2017objective}.
+Such methodology is applied to various tasks and music composition is amongst them \cite{limaguimaraes2017objective}.
+
+Nevertheless, not all methods for automated music generation use deep learning.
+Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} where selection is based on scores returned by hand-written evaluation rules.
 
@@
 \label{subsec:crossentropy}
+
+Initially, cross-entropy method was developed for estimation of rare events probability \cite{rubinstein1997optimization}.
+However, it was found that it is also appropriate for solving optimization problems.
+More detailed discussion of cross-entropy method can be found in \cite{boer2005tutorial}.
 
@@
 \label{sec:improvements}
-% better evaluational rules (tonality);
+% better evaluational rules (tonality, rhythm);
 % randomized initial observations;

### 2019-10-06 docs/paper/paper.tex
docs: Continue writing paper draft

@@
 \usepackage{microtype}      % microtypography
+\usepackage{algorithm}      % environment for algorithms
+\usepackage{algorithmic}    % pseudocode
 
@@
 Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions.
-These parameters are trained with cross-entropy method (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).
+These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).
 
@@
 
+\begin{algorithm}
+	\caption{Cross-entropy method for optimization} \label{alg:crossentropy}
+	\textbf{Input:} $X$ -- set of elements, $f: X \to \mathbb{R}$ -- target function, $u(\cdot, w)$ -- probabilistic distribution over $X$ parametrized by vector $w$. \\
+	\textbf{Output:} $\hat{w}$ -- approximate solution to the problem $\max_w \mathbb{E}_{x \sim u(\cdot, w)} f(x)$. \\
+	\textbf{Hyperparameters:} $w_0$ -- initial value of $w$; $N$ -- number of iterations, $n$ -- number of vectors to draw at each iteration, $\sigma$ -- standard deviation for vectors generation, $m$ -- number of trials for each vector; $\rho$ -- fraction of best vectors to use for update; $\alpha$ -- smoothing coefficient of updates.
+	\begin{algorithmic}[1]
+		\FORALL{$i \in \{1, \dots, N\}$}
+		    \FORALL{$j \in \{1, \dots, n\}$}
+		        \STATE{draw $w_{ij} \sim \mathcal{N}(\cdot \vert w_{i-1}, \sigma)$}
+		        \STATE{$r_j \gets \sum_{k = 1}^{m} f(x_k)$ where $x_k \sim u(\cdot, w_{ij})$}
+		    \ENDFOR
+		    \STATE{$r_{\mathrm{threshold}} \gets$ $[\rho n]$-th highest value of $\{r_j: j \in \{1, \dots, n\}\}$}
+		    \STATE{$J \gets \{j: r_j \ge r_{\mathrm{threshold}}\}$}
+		    \STATE{$w_i \gets \alpha w_{i-1} + (1 - \alpha)(\sum_{j \in J} w_{ij}) / [\rho n]$}
+		\ENDFOR
+	    \STATE{$\hat{w} \gets w_N$}
+	\end{algorithmic}
+\end{algorithm}
+
+Algorithm \ref{alg:crossentropy} defines a variant of cross-entropy method for optimization.
+Sometimes, hyperparameter $m$ is omitted and intermediate results are not aggregated over multiple trials.
+In case of $u(\cdot, w)$ that acts like a deterministic function of $w$, $m$ is redundant, but, in general case, terminal result can be improved by setting $m > 1$.
+
 \section{Methodology}
 \label{sec:methodology}
-% dataset -> rules; backpropagation -> cross-entropy; learning rate -> weights std.
+
+\subsection{Setup}
+\label{subsec:setup}
+
+For this study, piano roll is chosen as music representation format, but there are some subtleties.
+
+Here, each row of a table does not correspond to a particular note or pitch.
+Relative axis is used instead of absolute axis.
+All that is important is that each row is one semitone higher than the row right below it.
+Thus, a musical piece is given only up to transposition and user-defined value for a pivot note is required to define absolute pitches.
+
+\subsection{Evaluational Rules}
+\label{subsec:setup}
 

### 2019-10-07 rlmusician/environment/environment.py
feat: Randomize observation padding

@@ -42,6 +42,7 @@ class PianoRollEnv(gym.Env):
             max_n_stalled_episode_steps: int,
             scoring_coefs: Dict[str, float],
             scoring_fn_params: Dict[str, Dict[str, Any]],
+            padding_mean: float,
             rendering_params: Dict[str, Any]
     ):
         """
@@ -61,6 +62,9 @@ class PianoRollEnv(gym.Env):
             mapping from scoring function names to their weights in final score
         :param scoring_fn_params:
             mapping from scoring function names to their parameters
+        :param padding_mean:
+            mean of binomial distribution for observation padding at
+            early piano roll's time steps
         :param rendering_params:
             settings of environment rendering
         """
@@ -70,6 +74,7 @@ class PianoRollEnv(gym.Env):
         self.max_n_stalled_episode_steps = max_n_stalled_episode_steps
         self.scoring_coefs = scoring_coefs
         self.scoring_fn_params = scoring_fn_params
+        self.padding_mean = padding_mean
         self.rendering_params = rendering_params
 
         self.piano_roll = None
@@ -134,10 +139,10 @@ class PianoRollEnv(gym.Env):
         if steps_to_see[0] >= 0:
             observation = self.piano_roll[:, steps_to_see[0]:steps_to_see[1]]
         else:
-            observation = np.hstack((
-                np.zeros((self.n_semitones, -steps_to_see[0]), dtype=np.int32),
-                self.piano_roll[:, 0:steps_to_see[1]]
-            ))
+            padding_size = (self.n_semitones, -steps_to_see[0])
+            padding = np.random.binomial(1, self.padding_mean, padding_size)
+            valid_roll_part = self.piano_roll[:, 0:steps_to_see[1]]
+            observation = np.hstack((padding, valid_roll_part))
         done = self.n_piano_roll_steps_passed == self.n_roll_steps - 1
         reward = self.__evaluate() if done else 0
         info = {}

### 2019-10-07 setup.py

@@ -17,7 +17,7 @@ with open(os.path.join(current_dir, 'README.md'), encoding='utf-8') as f:
 
 setup(
     name='rl-musician',
-    version='0.1.1',
+    version='0.1.2',
     description=description,
     long_description=long_description,
     long_description_content_type='text/markdown',

### 2019-10-07 tests/environment/test_environment.py

@@ -28,6 +28,7 @@ class TestPianoRollEnv:
                     max_n_stalled_episode_steps=2,
                     scoring_coefs={'horizontal_variance': 1},
                     scoring_fn_params={},
+                    padding_mean=0.1,
                     rendering_params={}
                 ),
                 # `actions`
@@ -59,6 +60,7 @@ class TestPianoRollEnv:
                     max_n_stalled_episode_steps=2,
                     scoring_coefs={'horizontal_variance': 1},
                     scoring_fn_params={},
+                    padding_mean=0,
                     rendering_params={}
                 ),
                 # `expected`

### 2019-10-07 rlmusician/environment/scoring.py
feat: Score number of notes in wrong key

@@ -213,6 +213,36 @@ def score_conjunct_motion(
     return score
 
 
+def score_tonality(
+        roll: np.ndarray,
+        scale: str = 'major',
+        tonic_position: int = 4
+) -> float:
+    """
+    Score composition based on absence of notes not from specified scale.
+
+    :param roll:
+        piano roll
+    :param scale:
+        name of scale
+    :param tonic_position:
+        number of the row (from bottom) that corresponds to the tonic
+    :return:
+        number of played notes not from the specified scale
+        (with negative sign)
+    """
+    scale_to_wrong_notes = {
+        'major': [0, 1, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0],
+        'minor': [0, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1]
+    }
+    mask = np.array(scale_to_wrong_notes[scale][::-1]).reshape((-1, 1))
+    mask = np.tile(mask, (np.ceil(roll.shape[0] / mask.shape[0]), 1))
+    mask = mask[-roll.shape[0]:, :]
+    mask = np.roll(mask, -tonic_position)
+    score = -np.sum(mask * roll).item()
+    return score
+
+
 def get_scoring_functions_registry() -> Dict[str, Callable]:
     """
     Get mapping from names of scoring functions to scoring functions.
@@ -226,6 +256,7 @@ def get_scoring_functions_registry() -> Dict[str, Callable]:
         'absence_of_long_sounds': score_absence_of_long_sounds,
         'noncyclicity': score_noncyclicity,
         'consonances': score_consonances,
-        'conjunct_motion': score_conjunct_motion
+        'conjunct_motion': score_conjunct_motion,
+        'tonality': score_tonality
     }
     return registry

### 2019-10-07 tests/environment/test_scoring.py

@@ -16,7 +16,8 @@ from rlmusician.environment.scoring import (
     score_absence_of_long_sounds,
     score_noncyclicity,
     score_consonances,
-    score_conjunct_motion
+    score_conjunct_motion,
+    score_tonality
 )
 
 
@@ -358,4 +359,63 @@ def test_score_conjunct_motion(
 ) -> None:
     """Test `score_conjunct_motion` function."""
     result = score_conjunct_motion(roll, max_n_semitones, max_n_time_steps)
+    assert result == expected
+
+
+@pytest.mark.parametrize(
+    "roll, scale, tonic_position, expected",
+    [
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0],
+                [0, 0, 1],
+                [1, 0, 0],
+                [0, 1, 0],
+                [0, 0, 0],
+                [0, 1, 0],
+                [1, 0, 1],
+                [0, 0, 0],
+                [1, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+            ]),
+            # `scale`
+            'major',
+            # `tonic_position`
+            4,
+            # `expected`
+            -3
+        ),
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0],
+                [0, 0, 1],
+                [1, 0, 0],
+                [0, 1, 0],
+                [0, 0, 0],
+                [0, 1, 0],
+                [1, 0, 1],
+                [0, 0, 0],
+                [1, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+            ]),
+            # `scale`
+            'minor',
+            # `tonic_position`
+            0,
+            # `expected`
+            -2
+        )
+    ]
+)
+def test_score_tonality(
+        roll: np.ndarray, scale: str, tonic_position: int, expected: float
+) -> None:
+    """Test `score_tonality` function."""
+    result = score_tonality(roll, scale, tonic_position)
     assert result == expected

### 2019-10-07 rlmusician/environment/scoring.py
fix: Cast np.float to int

@@ -236,7 +236,7 @@ def score_tonality(
         'minor': [0, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1]
     }
     mask = np.array(scale_to_wrong_notes[scale][::-1]).reshape((-1, 1))
-    mask = np.tile(mask, (np.ceil(roll.shape[0] / mask.shape[0]), 1))
+    mask = np.tile(mask, (int(np.ceil(roll.shape[0] / mask.shape[0])), 1))
     mask = mask[-roll.shape[0]:, :]
     mask = np.roll(mask, -tonic_position)
     score = -np.sum(mask * roll).item()

### 2019-10-12 rlmusician/environment/environment.py
refac: Drop the action that moves to the next time step

@@ -49,10 +49,11 @@ class PianoRollEnv(gym.Env):
         Initialize instance.
 
         :param n_semitones:
-            number of consecutive semitones (piano keys) available to an agent
+            size of available pitch range in semitones; in other words,
+            number of consecutive piano keys available to an agent
         :param n_roll_steps:
-            total duration of composition in piano roll's time steps
-            (in other words, number of columns of piano roll)
+            total duration of composition in piano roll's time steps;
+            in other words, number of columns of piano roll
         :param n_observed_roll_steps:
             number of previous piano roll's time steps available for observing
         :param max_n_stalled_episode_steps:
@@ -82,9 +83,7 @@ class PianoRollEnv(gym.Env):
         self.n_episode_steps_passed = None
         self.n_stalled_episode_steps = None
 
-        self.action_space = gym.spaces.Discrete(
-            n_semitones + 1  # The last action stands for step forward on roll.
-        )
+        self.action_space = gym.spaces.Discrete(n_semitones)
         self.observation_space = gym.spaces.Box(
             low=0,
             high=1,
@@ -120,13 +119,9 @@ class PianoRollEnv(gym.Env):
                     (helpful for debugging and sometimes learning).
         """
         # Act.
-        if action != self.n_semitones:
-            self.piano_roll[action, self.n_piano_roll_steps_passed] = 1
-            self.n_stalled_episode_steps += 1
-        force_movement = (
-            self.n_stalled_episode_steps == self.max_n_stalled_episode_steps
-        )
-        if force_movement or action == self.n_semitones:
+        self.piano_roll[action, self.n_piano_roll_steps_passed] = 1
+        self.n_stalled_episode_steps += 1
+        if self.n_stalled_episode_steps == self.max_n_stalled_episode_steps:
             self.n_piano_roll_steps_passed += 1
             self.n_stalled_episode_steps = 0
         self.n_episode_steps_passed += 1

### 2019-10-12 rlmusician/environment/scoring.py
refac: Drop unnecessary scoring functions

@@ -24,41 +24,13 @@ from rlmusician.utils import (
 N_SEMITONES_PER_OCTAVE = 12
 
 
-def score_horizontal_variance(roll: np.ndarray) -> float:
-    """
-    Score composition based on variance of usage of each note.
-
-    It is a proxy of how non-trivial melody and harmony are.
-
-    :param roll:
-        piano roll
-    :return:
-        averaged over all notes variance of their usage in time
-    """
-    return np.mean(np.var(roll, axis=1)).item()
-
-
-def score_vertical_variance(roll: np.ndarray) -> float:
-    """
-    Score composition based on variance of notes played at each time step.
-
-    It is a proxy of how non-trivial harmony is.
-
-    :param roll:
-        piano roll
-    :return:
-        averaged over all time steps variance of notes played there
-    """
-    return np.mean(np.var(roll, axis=0)).item()
-
-
-def score_absence_of_long_sounds(
+def score_absence_of_constant_notes(
         roll: np.ndarray, max_n_time_steps: int = 4
 ) -> float:
     """
-    Score composition based on absence of too long sounds.
+    Score composition based on absence of notes played for too long time.
 
-    It is a proxy of how non-trivial melody and harmony are.
+    It is a proxy of how non-trivial composition is.
 
     :param roll:
         piano roll
@@ -76,33 +48,36 @@ def score_absence_of_long_sounds(
     return score
 
 
-def score_noncyclicity(
-        roll: np.ndarray, max_n_time_steps: int = 4, max_share: float = 0.2
+def score_conjunct_motion(
+        roll: np.ndarray,
+        max_n_semitones: int = 2,
+        max_n_time_steps: int = 1
 ) -> float:
     """
-    Score composition based on absence of cyclically repeated parts.
-
-    It is a proxy of how non-trivial composition is.
+    Score composition based on presence of small changes in pitch over time.
 
     :param roll:
         piano roll
+    :param max_n_semitones:
+        maximum number of semitones to consider change in pitch as small
     :param max_n_time_steps:
-        maximum duration of a fragment to be tested on cyclical repetitiveness
-    :param max_share:
-        upper limit on share of filled non-cyclically cells;
-        values that are higher, are clipped
+        maximum number of time steps for finding previous notes of close pitch
     :return:
[…]

### 2019-10-12 tests/environment/test_environment.py

@@ -26,15 +26,15 @@ class TestPianoRollEnv:
                     n_roll_steps=5,
                     n_observed_roll_steps=3,
                     max_n_stalled_episode_steps=2,
-                    scoring_coefs={'horizontal_variance': 1},
+                    scoring_coefs={'tonality': 1},
                     scoring_fn_params={},
                     padding_mean=0.1,
                     rendering_params={}
                 ),
                 # `actions`
-                [2] + [1 for _ in range(7)],
+                [2] + [0 for _ in range(7)],
                 # `expected`
-                0.064
+                -1
             )
         ]
     )
@@ -46,7 +46,7 @@ class TestPianoRollEnv:
         for action in actions:
             observation, reward, done, info = env.step(action)
         assert done
-        assert round(reward, 8) == expected
+        assert reward == expected
 
     @pytest.mark.parametrize(
         "env, expected",
@@ -58,7 +58,7 @@ class TestPianoRollEnv:
                     n_roll_steps=5,
                     n_observed_roll_steps=3,
                     max_n_stalled_episode_steps=2,
-                    scoring_coefs={'horizontal_variance': 1},
+                    scoring_coefs={'tonality': 1},
                     scoring_fn_params={},
                     padding_mean=0,
                     rendering_params={}

### 2019-10-12 tests/environment/test_scoring.py

@@ -11,72 +11,12 @@ import numpy as np
 import pytest
 
 from rlmusician.environment.scoring import (
-    score_horizontal_variance,
-    score_vertical_variance,
-    score_absence_of_long_sounds,
+    score_absence_of_constant_notes,
     score_noncyclicity,
     score_consonances,
     score_conjunct_motion,
     score_tonality
 )
-
-
-@pytest.mark.parametrize(
-    "roll, expected",
-    [
-        (
-            # `roll`
-            np.array([
-                [0, 0],
-                [1, 1]
-            ]),
-            # `expected`
-            0
-        ),
-        (
-            # `roll`
-            np.array([
-                [1, 0],
-                [0, 1]
-            ]),
-            # `expected`
-            0.25
-        ),
-    ]
-)
-def test_score_horizontal_variance(roll: np.ndarray, expected: float) -> None:
-    """Test `score_horizontal_variance` function."""
-    result = score_horizontal_variance(roll)
-    assert result == expected
-
-
-@pytest.mark.parametrize(
-    "roll, expected",
-    [
-        (
-            # `roll`
-            np.array([
-                [1, 0],
-                [1, 0]
-            ]),
-            # `expected`
-            0
-        ),
-        (
-            # `roll`
-            np.array([
-                [1, 0],
-                [0, 1]
-            ]),
-            # `expected`
-            0.25
-        ),
-    ]
-)
-def test_score_vertical_variance(roll: np.ndarray, expected: float) -> None:
-    """Test `score_vertical_variance` function."""
-    result = score_vertical_variance(roll)
-    assert result == expected
 
 
 @pytest.mark.parametrize(
@@ -120,11 +60,11 @@ def test_score_vertical_variance(roll: np.ndarray, expected: float) -> None:
         ),
     ]
 )
-def test_score_absence_of_long_sounds(
+def test_score_absence_of_constant_notes(
         roll: np.ndarray, max_n_time_steps: int, expected: int
 ) -> None:
-    """Test `score_absence_of_long_sounds` function."""
-    result = score_absence_of_long_sounds(roll, max_n_time_steps)
+    """Test `score_absence_of_constant_notes` function."""
+    result = score_absence_of_constant_notes(roll, max_n_time_steps)
     assert result == expected
 
 
@@ -410,6 +350,24 @@ def test_score_conjunct_motion(
             0,
             # `expected`
             -2
+        ),
+        (
+            # `roll`
+            np.array([
[…]

### 2019-10-12 rlmusician/environment/scoring.py
feat: Score composition based on desired number of melodic lines

@@ -185,6 +185,26 @@ def score_noncyclicity(
     return score
 
 
+def score_number_of_simultaneously_played_notes(
+        roll: np.ndarray, n_lines: int = 3
+) -> float:
+    """
+    Score composition based on number of simultaneously played notes.
+
+    It is a proxy for presence of desired number of melodic lines.
+
+    :param roll:
+        piano roll
+    :param n_lines:
+        desired number of melodic lines
+    :return:
+        negative sum over all time steps of absolute deviations
+        of number of simultaneously played notes from the desired number
+    """
+    score = -np.sum(np.abs((np.sum(roll, axis=0) - n_lines))).item()
+    return score
+
+
 def score_tonality(
         roll: np.ndarray,
         scale: str = 'major',
@@ -226,6 +246,7 @@ def get_scoring_functions_registry() -> Dict[str, Callable]:
         'absence_of_constant_notes': score_absence_of_constant_notes,
         'conjunct_motion': score_conjunct_motion,
         'consonances': score_consonances,
+        'lines': score_number_of_simultaneously_played_notes,
         'noncyclicity': score_noncyclicity,
         'tonality': score_tonality
     }

### 2019-10-12 tests/environment/test_scoring.py

@@ -12,9 +12,10 @@ import pytest
 
 from rlmusician.environment.scoring import (
     score_absence_of_constant_notes,
-    score_noncyclicity,
-    score_consonances,
     score_conjunct_motion,
+    score_consonances,
+    score_noncyclicity,
+    score_number_of_simultaneously_played_notes,
     score_tonality
 )
 
@@ -65,134 +66,6 @@ def test_score_absence_of_constant_notes(
 ) -> None:
     """Test `score_absence_of_constant_notes` function."""
     result = score_absence_of_constant_notes(roll, max_n_time_steps)
-    assert result == expected
-
-
-@pytest.mark.parametrize(
-    "roll, max_n_time_steps, max_share, expected",
-    [
-        (
-            # `roll`
-            np.array([
-                [1, 0, 1, 0, 1, 0],
-                [0, 1, 0, 1, 0, 1],
-            ]),
-            # `max_n_time_steps`
-            1,
-            # `max_share`
-            0.5,
-            # `expected`
-            1
-        ),
-        (
-            # `roll`
-            np.array([
-                [1, 0, 1, 0, 1, 0],
-                [0, 1, 0, 1, 0, 1],
-            ]),
-            # `max_n_time_steps`
-            2,
-            # `max_share`
-            0.5,
-            # `expected`
-            1 / 3
-        ),
-        (
-            # `roll`
-            np.array([
-                [1, 0, 1, 0, 1, 0],
-                [0, 1, 0, 1, 0, 1],
-                [0, 1, 1, 0, 0, 1],
-            ]),
-            # `max_n_time_steps`
-            3,
-            # `max_share`
-            1,
-            # `expected`
-            5 / 18
-        ),
-    ]
-)
-def test_score_noncyclicity(
-        roll: np.ndarray, max_n_time_steps: int, max_share: float,
-        expected: float
-) -> None:
-    """Test `score_noncyclicity` function."""
-    result = score_noncyclicity(roll, max_n_time_steps, max_share)
-    assert result == expected
-
-
-@pytest.mark.parametrize(
-    "roll, interval_consonances, distance_weights, expected",
-    [
-        (
-            # `roll`
-            np.array([
-                [0, 0, 0, 0, 0],
-                [1, 1, 0, 0, 0],
-                [0, 0, 0, 0, 1],
-                [1, 1, 1, 0, 0]
-            ]),
[…]

### 2019-10-12 rlmusician/environment/scoring.py
feat: Score contrary motion

@@ -155,6 +155,27 @@ def score_consonances(
     return score
 
 
+def score_contrary_motion(roll: np.ndarray) -> float:
+    """
+    Score composition based on number of time steps with contrary motion.
+
+    Here, a time step has contrary motion if there are at least two lines
+    moving in opposite directions.
+
+    :param roll:
+        piano roll
+    :return:
+        number of time steps with contrary motion
+    """
+    diff = np.diff(roll, axis=1, prepend=0)[:, 1:]
+    cum_sums = np.cumsum(diff, axis=0)
+    moves_down_penalty = np.abs(np.max(cum_sums, axis=0) - 1)
+    moves_up_penalty = np.abs(np.min(cum_sums, axis=0) + 1)
+    score = -np.sum((moves_down_penalty + moves_up_penalty) > 0).item()
+    score += roll.shape[1]  # For the sake of interpretation.
+    return score
+
+
 def score_noncyclicity(
         roll: np.ndarray, max_n_time_steps: int = 4, max_share: float = 0.2
 ) -> float:
@@ -246,6 +267,7 @@ def get_scoring_functions_registry() -> Dict[str, Callable]:
         'absence_of_constant_notes': score_absence_of_constant_notes,
         'conjunct_motion': score_conjunct_motion,
         'consonances': score_consonances,
+        'contrary_motion': score_contrary_motion,
         'lines': score_number_of_simultaneously_played_notes,
         'noncyclicity': score_noncyclicity,
         'tonality': score_tonality

### 2019-10-12 tests/environment/test_scoring.py

@@ -14,6 +14,7 @@ from rlmusician.environment.scoring import (
     score_absence_of_constant_notes,
     score_conjunct_motion,
     score_consonances,
+    score_contrary_motion,
     score_noncyclicity,
     score_number_of_simultaneously_played_notes,
     score_tonality
@@ -246,6 +247,55 @@ def test_score_consonances(
 ) -> None:
     """Test `score_consonances` function."""
     result = score_consonances(roll, interval_consonances, distance_weights)
+    assert result == expected
+
+
+@pytest.mark.parametrize(
+    "roll, expected",
+    [
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0, 0, 0],
+                [0, 1, 0, 0, 0],
+                [1, 0, 1, 0, 1],
+                [0, 0, 0, 1, 0],
+                [0, 0, 0, 0, 0],
+                [0, 0, 0, 0, 1],
+                [0, 0, 0, 1, 0],
+                [1, 0, 0, 0, 0],
+                [0, 1, 1, 0, 0],
+                [0, 0, 0, 0, 0],
+                [0, 0, 0, 0, 0],
+                [0, 0, 0, 0, 0],
+            ]),
+            # `expected`
+            3
+        ),
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0, 0, 0],
+                [0, 0, 0, 0, 0],
+                [0, 0, 0, 0, 0],
+                [1, 0, 0, 0, 1],
+                [0, 1, 0, 1, 0],
+                [0, 0, 1, 0, 0],
+                [1, 0, 1, 0, 1],
+                [0, 1, 0, 1, 0],
+                [0, 0, 0, 1, 0],
+                [0, 0, 1, 0, 1],
+                [0, 1, 0, 0, 0],
+                [1, 0, 0, 0, 0],
+            ]),
+            # `expected`
+            5
+        )
+    ]
+)
+def test_score_contrary_motion(roll: np.ndarray, expected: float) -> None:
+    """Test `score_contrary_motion` function."""
+    result = score_contrary_motion(roll)
     assert result == expected
 
 

### 2019-10-12 rlmusician/__main__.py
feat: Update default values

@@ -29,7 +29,7 @@ def parse_cli_args() -> argparse.Namespace:
         help='path to configuration file'
     )
     parser.add_argument(
-        '-p', '--populations', type=int, default=25,
+        '-p', '--populations', type=int, default=100,
         help='number of populations for agent training'
     )
     parser.add_argument(

### 2019-10-12 docs/paper/paper.tex
docs: Update some paragraphs of paper

@@
 Brief outline of the approach is as follows.
-There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps.
+There is an environment with 2D table representing piano roll which means that rows of the table correspond to notes and columns of the table correspond to time steps.
 An agent observes current time step and some previous time steps.
-An action of the agent can be either playing a note at the current time step or moving to the next time step.
+An action of the agent is playing a note at the current time step.
+After certain number of actions are done, a movement to the next time step happens.
 When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition.
@@
 \end{itemize}
-
-Further, sequential generation of composition by neural network implies that any well-studied architectures from sequence generation tasks can be plugged in and tested.
-The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or alternative methods.
 
@@
 Nevertheless, not all methods for automated music generation use deep learning.
-Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} where selection is based on scores returned by hand-written evaluation rules.
+Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} or Markov chains \cite{farbood2001analysis}.
+Comprehensive review that covers plenty of approaches beyond deep learning can be found in \cite{fernandez2013ai}.
 
@@
 In case of $u(\cdot, w)$ that acts like a deterministic function of $w$, $m$ is redundant, but, in general case, terminal result can be improved by setting $m > 1$.
+
+In context of reinforcement learning, cross-entropy method is used for searching parameters of agent that result in maximum expected reward.
+Such approach is best suitable for environments where reward becomes known only after episode is finished.
+If there are rewards granted after intermediate steps, methods based on Q-values usually outperform it.
 

### 2019-10-13 rlmusician/agent/actor_model.py
refac: Decrease probability of cycling

@@ -60,9 +60,9 @@ def create_actor_model(
 
     roll_input = Input(shape=observation_shape, name='piano_roll')
     reshaped_input = Reshape(observation_shape + (1,))(roll_input)
-    roll_hidden = Conv2D(4, (6, 6), activation='relu')(reshaped_input)
-    roll_embedded = Flatten()(roll_hidden)
-    output = Dense(n_actions, activation='softmax')(roll_embedded)
+    conv_hidden = Conv2D(2, (3, 3), activation='relu')(reshaped_input)
+    embedding = Flatten()(conv_hidden)
+    output = Dense(n_actions, activation='softmax')(embedding)
     model = Model(inputs=roll_input, outputs=output)
     model.compile(optimizer='sgd', loss='mse')  # Arbitrary unused values.
     return model

### 2019-10-13 rlmusician/environment/scoring.py

@@ -74,7 +74,7 @@ def score_conjunct_motion(
         if i != 0
     ]
     altered_roll = np.concatenate(altered_rolls, axis=2)
-    close_notes_roll = altered_roll.max(axis=2)
+    close_notes_roll = np.max(altered_roll, axis=2)
     starting_notes_roll = np.clip(roll - shift_horizontally(roll, 1), 0, None)
     matches = np.minimum(starting_notes_roll, close_notes_roll)
     score = np.sum(matches).item()
@@ -177,7 +177,7 @@ def score_contrary_motion(roll: np.ndarray) -> float:
 
 
 def score_noncyclicity(
-        roll: np.ndarray, max_n_time_steps: int = 4, max_share: float = 0.2
+        roll: np.ndarray, max_n_time_steps: int = 8, max_share: float = 0.2
 ) -> float:
     """
     Score composition based on absence of cyclically repeated parts.

### 2019-10-13 rlmusician/environment/scoring.py
feat: Score share of triads built on scale tonic

@@ -12,7 +12,7 @@ Author: Nikolay Lysenko
 """
 
 
-from typing import Callable, Dict
+from typing import Callable, Dict, Optional
 
 import numpy as np
 
@@ -22,9 +22,42 @@ from rlmusician.utils import (
 
 
 N_SEMITONES_PER_OCTAVE = 12
+DEFAULT_TONIC_POSITION = 4
 
 
-def score_absence_of_constant_notes(
+def score_absence_of_outer_notes(
+        roll: np.ndarray,
+        scale: str = 'major',
+        tonic_position: Optional[int] = None
+) -> float:
+    """
+    Score composition based on absence of notes not from specified scale.
+
+    :param roll:
+        piano roll
+    :param scale:
+        name of scale
+    :param tonic_position:
+        number of the row (from bottom) that corresponds to the tonic
+    :return:
+        number of played notes not from the specified scale
+        (with negative sign)
+    """
+    if tonic_position is None:
+        tonic_position = DEFAULT_TONIC_POSITION
+    scale_to_wrong_notes = {
+        'major': [0, 1, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0],
+        'minor': [0, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1]
+    }
+    mask = np.array(scale_to_wrong_notes[scale][::-1]).reshape((-1, 1))
+    mask = np.tile(mask, (int(np.ceil(roll.shape[0] / mask.shape[0])), 1))
+    mask = np.roll(mask, -tonic_position)
+    mask = mask[-roll.shape[0]:, :]
+    score = -np.sum(mask * roll).item()
+    return score
+
+
+def score_absence_of_stalled_notes(
         roll: np.ndarray, max_n_time_steps: int = 4
 ) -> float:
     """
@@ -177,7 +210,7 @@ def score_contrary_motion(roll: np.ndarray) -> float:
 
 
 def score_noncyclicity(
-        roll: np.ndarray, max_n_time_steps: int = 8, max_share: float = 0.2
+        roll: np.ndarray, max_n_time_steps: int = 8, max_share: float = 0.125
 ) -> float:
     """
     Score composition based on absence of cyclically repeated parts.
@@ -226,33 +259,44 @@ def score_number_of_simultaneously_played_notes(
     return score
 
 
-def score_tonality(
+def score_usage_of_tonic(
         roll: np.ndarray,
-        scale: str = 'major',
[…]

### 2019-10-13 tests/environment/test_environment.py

@@ -26,7 +26,7 @@ class TestPianoRollEnv:
                     n_roll_steps=5,
                     n_observed_roll_steps=3,
                     max_n_stalled_episode_steps=2,
-                    scoring_coefs={'tonality': 1},
+                    scoring_coefs={'absence_of_outer_notes': 1},
                     scoring_fn_params={},
                     padding_mean=0.1,
                     rendering_params={}
@@ -58,7 +58,7 @@ class TestPianoRollEnv:
                     n_roll_steps=5,
                     n_observed_roll_steps=3,
                     max_n_stalled_episode_steps=2,
-                    scoring_coefs={'tonality': 1},
+                    scoring_coefs={'absence_of_outer_notes': 1},
                     scoring_fn_params={},
                     padding_mean=0,
                     rendering_params={}

### 2019-10-13 tests/environment/test_scoring.py

@@ -11,14 +11,92 @@ import numpy as np
 import pytest
 
 from rlmusician.environment.scoring import (
-    score_absence_of_constant_notes,
+    score_absence_of_outer_notes,
+    score_absence_of_stalled_notes,
     score_conjunct_motion,
     score_consonances,
     score_contrary_motion,
     score_noncyclicity,
     score_number_of_simultaneously_played_notes,
-    score_tonality
+    score_usage_of_tonic
 )
+
+
+@pytest.mark.parametrize(
+    "roll, scale, tonic_position, expected",
+    [
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0],
+                [0, 0, 1],
+                [1, 0, 0],
+                [0, 1, 0],
+                [0, 0, 0],
+                [0, 1, 0],
+                [1, 0, 1],
+                [0, 0, 0],
+                [1, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+            ]),
+            # `scale`
+            'major',
+            # `tonic_position`
+            4,
+            # `expected`
+            -3
+        ),
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0],
+                [0, 0, 1],
+                [1, 0, 0],
+                [0, 1, 0],
+                [0, 0, 0],
+                [0, 1, 0],
+                [1, 0, 1],
+                [0, 0, 0],
+                [1, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+                [0, 0, 0],
+            ]),
+            # `scale`
+            'minor',
+            # `tonic_position`
+            0,
+            # `expected`
+            -2
+        ),
+        (
+            # `roll`
+            np.array([
+                [0, 0, 0],
+                [0, 0, 1],
+                [1, 0, 0],
+                [0, 1, 0],
+                [0, 0, 0],
+                [0, 1, 0],
+                [1, 0, 1],
+            ]),
+            # `scale`
+            'major',
+            # `tonic_position`
+            2,
+            # `expected`
+            -4
+        )
+    ]
+)
+def test_score_absence_of_outer_notes(
+        roll: np.ndarray, scale: str, tonic_position: int, expected: float
+) -> None:
+    """Test `score_absence_of_outer_notes` function."""
+    result = score_absence_of_outer_notes(roll, scale, tonic_position)
+    assert result == expected
 
 
 @pytest.mark.parametrize(
[…]

### 2019-10-13 rlmusician/agent/actor_model.py
feat: Add more past info in observation and simplify actor model

@@ -43,7 +43,7 @@ def create_actor_model(
     Create simple actor network for `PianoRollEnv`.
 
     :param observation_shape:
-        shape of observed part of piano roll
+        shape of observation
     :param n_actions:
         number of actions available to the agent
     :return:
@@ -55,14 +55,11 @@ def create_actor_model(
         warnings.simplefilter('ignore')
         import_keras_silently()
         from keras.models import Model
-        from keras.layers import Conv2D, Dense, Flatten, Input, Reshape
+        from keras.layers import Dense, Input
         mute_tensorflow()
 
-    roll_input = Input(shape=observation_shape, name='piano_roll')
-    reshaped_input = Reshape(observation_shape + (1,))(roll_input)
-    conv_hidden = Conv2D(2, (3, 3), activation='relu')(reshaped_input)
-    embedding = Flatten()(conv_hidden)
-    output = Dense(n_actions, activation='softmax')(embedding)
-    model = Model(inputs=roll_input, outputs=output)
+    observed_input = Input(shape=observation_shape, name='piano_roll')
+    output = Dense(n_actions, activation='softmax')(observed_input)
+    model = Model(inputs=observed_input, outputs=output)
     model.compile(optimizer='sgd', loss='mse')  # Arbitrary unused values.
     return model

### 2019-10-13 rlmusician/environment/environment.py

@@ -38,11 +38,10 @@ class PianoRollEnv(gym.Env):
             self,
             n_semitones: int,
             n_roll_steps: int,
-            n_observed_roll_steps: int,
-            max_n_stalled_episode_steps: int,
+            observation_decay: float,
+            n_draws_per_roll_step: int,
             scoring_coefs: Dict[str, float],
             scoring_fn_params: Dict[str, Dict[str, Any]],
-            padding_mean: float,
             rendering_params: Dict[str, Any]
     ):
         """
@@ -54,41 +53,38 @@ class PianoRollEnv(gym.Env):
         :param n_roll_steps:
             total duration of composition in piano roll's time steps;
             in other words, number of columns of piano roll
-        :param n_observed_roll_steps:
-            number of previous piano roll's time steps available for observing
-        :param max_n_stalled_episode_steps:
-            number of episode steps after which forced movement forward occurs
-            on piano roll
+        :param observation_decay:
+            coefficient of exponential decay for previous piano roll's
+            time steps
+        :param n_draws_per_roll_step:
+            number of episode steps after which movement to the next
+            piano roll's time step occurs
         :param scoring_coefs:
             mapping from scoring function names to their weights in final score
         :param scoring_fn_params:
             mapping from scoring function names to their parameters
-        :param padding_mean:
-            mean of binomial distribution for observation padding at
-            early piano roll's time steps
         :param rendering_params:
             settings of environment rendering
         """
         self.n_semitones = n_semitones
         self.n_roll_steps = n_roll_steps
-        self.n_observed_roll_steps = n_observed_roll_steps
-        self.max_n_stalled_episode_steps = max_n_stalled_episode_steps
+        self.observation_decay = observation_decay
+        self.n_draws_per_roll_step = n_draws_per_roll_step
         self.scoring_coefs = scoring_coefs
         self.scoring_fn_params = scoring_fn_params
-        self.padding_mean = padding_mean
         self.rendering_params = rendering_params
 
         self.piano_roll = None
-        self.n_piano_roll_steps_passed = None
-        self.n_episode_steps_passed = None
-        self.n_stalled_episode_steps = None
+        self.current_episode_step = None
[…]

### 2019-10-13 rlmusician/environment/scoring.py

@@ -44,7 +44,7 @@ def score_absence_of_outer_notes(
         (with negative sign)
     """
     if tonic_position is None:
-        tonic_position = DEFAULT_TONIC_POSITION
+        tonic_position = DEFAULT_TONIC_POSITION  # pragma: no cover
     scale_to_wrong_notes = {
         'major': [0, 1, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0],
         'minor': [0, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1]
@@ -283,7 +283,7 @@ def score_usage_of_tonic(
         in the lowest line
     """
     if tonic_position is None:
-        tonic_position = DEFAULT_TONIC_POSITION
+        tonic_position = DEFAULT_TONIC_POSITION  # pragma: no cover
 
     cum_summed = np.cumsum(roll, axis=0)
     summed = np.sum(roll, axis=0).reshape((1, -1))

### 2019-10-13 tests/environment/test_environment.py

@@ -20,15 +20,45 @@ class TestPianoRollEnv:
         "env, actions, expected",
         [
             (
+                    # `env`
+                    PianoRollEnv(
+                        n_semitones=5,
+                        n_roll_steps=5,
+                        observation_decay=0.5,
+                        n_draws_per_roll_step=2,
+                        scoring_coefs={'absence_of_outer_notes': 1},
+                        scoring_fn_params={},
+                        rendering_params={}
+                    ),
+                    # `actions`
+                    [2, 2, 1, 1, 1, 3, 3],
+                    # `expected`
+                    np.array([0, 0.75, 0.125, 1.5, 0])
+            )
+        ]
+    )
+    def test_observation(
+            self, env: PianoRollEnv, actions: List[int], expected: np.ndarray
+    ) -> None:
+        """Test that `step` method returns proper observation."""
+        env.reset()
+        for action in actions:
+            observation, reward, done, info = env.step(action)
+        assert not done
+        np.testing.assert_equal(observation, expected)
+
+    @pytest.mark.parametrize(
+        "env, actions, expected",
+        [
+            (
                 # `env`
                 PianoRollEnv(
                     n_semitones=5,
                     n_roll_steps=5,
-                    n_observed_roll_steps=3,
-                    max_n_stalled_episode_steps=2,
+                    observation_decay=0.5,
+                    n_draws_per_roll_step=2,
                     scoring_coefs={'absence_of_outer_notes': 1},
                     scoring_fn_params={},
-                    padding_mean=0.1,
                     rendering_params={}
                 ),
                 # `actions`
@@ -38,10 +68,10 @@ class TestPianoRollEnv:
             )
         ]
     )
-    def test_step(
+    def test_reward(
             self, env: PianoRollEnv, actions: List[int], expected: float
     ) -> None:
-        """Test `step` method."""
+        """Test that `step` method returns proper reward."""
         env.reset()
         for action in actions:
             observation, reward, done, info = env.step(action)
@@ -56,15 +86,14 @@ class TestPianoRollEnv:
                 PianoRollEnv(
                     n_semitones=5,
                     n_roll_steps=5,
-                    n_observed_roll_steps=3,
-                    max_n_stalled_episode_steps=2,
+                    observation_decay=0.5,
+                    n_draws_per_roll_step=2,
                     scoring_coefs={'absence_of_outer_notes': 1},
                     scoring_fn_params={},
[…]

### 2019-10-14 rlmusician/agent/crossentropy.py
refac: Prettify printing of training progress

@@ -210,6 +210,7 @@ class CrossEntropyAgent:
         :return:
             None
         """
+        print("population | avg_score_over_current_top |   global_best_score")
         for i_population in range(n_populations):
             args = [(self, deepcopy(env)) for _ in range(self.population_size)]
             entries = map_in_parallel(
@@ -230,9 +231,9 @@ class CrossEntropyAgent:
             top_scores = [x['score'] for x in top_entries]
             avg_top_score = np.mean(top_scores)
             print(
-                f"Population {i_population}: "
-                f"mean score over top candidates is {avg_top_score}, "
-                f"global best score is {self.best['score']}."
+                f"{i_population:>10} | "
+                f"{avg_top_score:>26} | "
+                f"{self.best['score']:>19}"
             )
         best_flat_weights = self.best['flat_weights']
         self.model = self.create_model(best_flat_weights)

### 2019-10-15 rlmusician/agent/actor_model.py
refac: Rename input

@@ -58,7 +58,7 @@ def create_actor_model(
         from keras.layers import Dense, Input
         mute_tensorflow()
 
-    observed_input = Input(shape=observation_shape, name='piano_roll')
+    observed_input = Input(shape=observation_shape, name='input')
     output = Dense(n_actions, activation='softmax')(observed_input)
     model = Model(inputs=observed_input, outputs=output)
     model.compile(optimizer='sgd', loss='mse')  # Arbitrary unused values.

### 2019-10-15 rlmusician/__main__.py
feat: Show detailed scores for test episodes

@@ -71,6 +71,8 @@ def main() -> None:
     now = datetime.datetime.now().strftime("%Y-%m-%d_%H:%M:%S,%f")
     weights_path = os.path.join(results_dir, f'agent_weights_{now}.h5f')
     agent.model.save_weights(weights_path)
+
+    env.verbose = True
     agent.test(env, n_episodes=cli_args.episodes)
 
 

### 2019-10-15 rlmusician/agent/crossentropy.py

@@ -250,6 +250,7 @@ class CrossEntropyAgent:
             None
         """
         for i_episode in range(n_episodes):
+            print(f"\nEpisode {i_episode}:")
             reward = run_episode(self.model, env)
             env.render()
-            print(f"Episode {i_episode}: reward is {reward}.")
+            print(f"Reward is {reward}.")

### 2019-10-15 rlmusician/environment/environment.py

@@ -21,7 +21,7 @@ import gym
 import numpy as np
 from sinethesizer.io.piano_roll_to_tsv import write_roll_to_tsv_file
 
-from rlmusician.environment.scoring import get_scoring_functions_registry
+from rlmusician.environment.scoring import evaluate
 from rlmusician.utils import (
     create_midi_from_piano_roll, create_wav_from_events
 )
@@ -78,6 +78,7 @@ class PianoRollEnv(gym.Env):
         self.current_episode_step = None
         self.current_roll_step = None
         self.n_draws_at_current_roll_step = None
+        self.verbose = False
 
         self.action_space = gym.spaces.Discrete(n_semitones)
         self.observation_space = gym.spaces.Box(
@@ -86,18 +87,6 @@ class PianoRollEnv(gym.Env):
             shape=(n_semitones,),
             dtype=np.float32
         )
-
-    def __evaluate(self) -> float:
-        """Evaluate current state of piano roll."""
-        score = 0
-        registry = get_scoring_functions_registry()
-        for fn_name, weight in self.scoring_coefs.items():
-            fn = registry[fn_name]
-            score += weight * fn(
-                self.piano_roll,
-                **self.scoring_fn_params.get(fn_name, {})
-            )
-        return score
 
     @property
     def __decay_coefs(self) -> np.ndarray:
@@ -137,7 +126,14 @@ class PianoRollEnv(gym.Env):
         known_piano_roll = self.piano_roll[:, :self.current_roll_step + 1]
         observation = np.sum(self.__decay_coefs * known_piano_roll, axis=1)
         done = self.current_roll_step == self.n_roll_steps - 1
-        reward = self.__evaluate() if done else 0
+        reward = 0
+        if done:
+            reward = evaluate(
+                self.piano_roll,
+                self.scoring_coefs,
+                self.scoring_fn_params,
+                self.verbose
+            )
         info = {}
         return observation, reward, done, info
 

### 2019-10-15 rlmusician/environment/scoring.py

@@ -12,7 +12,7 @@ Author: Nikolay Lysenko
 """
 
 
-from typing import Callable, Dict, Optional
+from typing import Any, Callable, Dict, Optional
 
 import numpy as np
 
@@ -318,3 +318,34 @@ def get_scoring_functions_registry() -> Dict[str, Callable]:
         'usage_of_tonic': score_usage_of_tonic
     }
     return registry
+
+
+def evaluate(
+        roll: np.ndarray,
+        scoring_coefs: Dict[str, float],
+        scoring_fn_params: Dict[str, Any],
+        verbose: bool = False
+) -> float:
+    """
+    Evaluate piano roll.
+
+    :param roll:
+        piano roll
+    :param scoring_coefs:
+        mapping from scoring function names to their weights in final score
+    :param scoring_fn_params:
+        mapping from scoring function names to their parameters
+    :param verbose:
+        if it is set to `True`, scores by all functions are printed separately
+    :return:
+        overall score of piano roll quality
+    """
+    score = 0
+    registry = get_scoring_functions_registry()
+    for fn_name, weight in scoring_coefs.items():
+        fn = registry[fn_name]
+        curr_score = weight * fn(roll, **scoring_fn_params.get(fn_name, {}))
+        if verbose:
+            print(f'{fn_name:>25}: {curr_score}')  # pragma: no cover
+        score += curr_score
+    return score

### 2019-10-16 docs/paper/paper.tex
docs: Revise draft of paper

@@
 
-\title{RL-Musician: A Tool for Music Composition with Deep Reinforcement Learning}
+\title{RL-Musician: A Tool for Music Composition with Reinforcement Learning}
 
@@
 Reinforcement learning can be applied instead if the goal is to reveal more aspects of creativity.
-On the other hand, numerous recent advances in music generation are achieved with supervised deep learning.
-In this paper, a method that uses both reinforcement learning and deep learning is developed for automated music composition.
-An agent with neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment.
+On the other hand, numerous recent advances in music generation are achieved with supervisedly trained neural networks.
+In this paper, a method that involves both reinforcement learning and neural networks is developed for automated music composition.
+An agent with shallow neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment.
 The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory.
-No dataset is involved, because cross-entropy method is used instead of gradient-based methods.
+Cross-entropy method is used for training instead of gradient-based methods and so no datasets are needed.
 \end{abstract}
@@
 There is an environment with 2D table representing piano roll which means that rows of the table correspond to notes and columns of the table correspond to time steps.
-An agent observes current time step and some previous time steps.
+A cell of the table contains 1 if the note is sounding at this step and 0 otherwise.
+An agent observes vector in which for every row there is an exponential moving average over all time steps up to the current time step (inclusively).
 An action of the agent is playing a note at the current time step.
@@
 
-In algorithmic composition domain, numerous links between reinforcement learning and deep learning are established.
+Usage of computers for automatic composition of music dates back to 1950s and, what is more, approaches that can be labeled as algorithmic are known since Guido of Arezzo (11th century).
+Therefore, it is impossible to list here all studies about the subject.
+An interested reader can find more information in specialized reviews such as \cite{fernandez2013ai}.
+What is discussed below is just some examples related to reinforcement learning.
+
+Reinforcement learning can be applied to algorithmic composition either as a standalone paradigm or in connection with deep learning.
+
+As for the first case, there are not \cite{yi2007automatic}.
+
+The second group (works where connections with deep learning are established) recently gained attention and several interesting approaches were developed there.
 
@@
 To define an environment, let its state be composed of recurrent neural networks states and previously played notes, let an action be an output for current time step and let reward depend on both evaluational rules and probability of output according to initial RNN.
-Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred over cross-entropy method as a training algorithm.
+Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred as a training algorithm.
 
 Usually, softmax activation function is used in the last layer of generative RNN.
-However, it is possible to consider a family of activation functions parametrized by one parameter $t$ such that output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$).
-Also, input vector can be extended by introducing an additional part indicating origination of initial input (so called plan).
-Then an environment can be defined so that state is a pair of $t$ and plan, actions are changes in either $t$ or plan, and reward depends on evaluation of produced with these settings piece \cite{kumar2019polyphonic}.
+However, it is possible to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$.
+Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$).
+In \cite{kumar2019polyphonic}, Boltzmann softmax is used and also input vector is extended by introducing an additional part indicating origination of initial input (so called plan).
+An environment is defined so that state is a pair of temperature and plan, actions are changes in either temperature or plan, and reward depends on evaluation of produced with these settings piece.
 
@@
 
-Nevertheless, not all methods for automated music generation use deep learning.
-Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} or Markov chains \cite{farbood2001analysis}.
-Comprehensive review that covers plenty of approaches beyond deep learning can be found in \cite{fernandez2013ai}.
+% Nevertheless, not all methods for automated music generation use deep learning.
+Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} or HMMs (Hidden Markov Models) \cite{farbood2001analysis}
 

### 2019-10-22 docs/paper/paper.tex
docs: Describe RL setup

@@
 Reinforcement learning can be applied to algorithmic composition either as a standalone paradigm or in connection with deep learning.
+As for the first case, probably, there are no influential works and only separate studies like \cite{yi2007automatic} are available.
+Conversely, the second group recently gained attention and several interesting approaches were developed there.
 
-As for the first case, there are not \cite{yi2007automatic}.
-
-The second group (works where connections with deep learning are established) recently gained attention and several interesting approaches were developed there.
-
-Reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}.
+For example, reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}.
 The goal is to make generated pieces more structured and conformed with music theory rules.
@@
 However, widespread notations for music assume sequences of discrete values but classical GAN work well only with continuous data, because gradient of discrete-valued functions is uninformative.
-Techniques originating from RL can be used to overcome this obstacle \cite{yu2016seqgan,hjelm2017boundary}.
+Techniques originating from reinforcement learning can be used to overcome this obstacle \cite{yu2016seqgan,hjelm2017boundary}.
 Namely, generator is trained with policy gradient method \cite{williams1992simple}.
 Such methodology is applied to various tasks and music composition is amongst them \cite{limaguimaraes2017objective}.
-
-% Nevertheless, not all methods for automated music generation use deep learning.
-Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} or HMMs (Hidden Markov Models) \cite{farbood2001analysis}
 
@@
 Thus, a musical piece is given only up to transposition and user-defined value for a pivot note is required to define absolute pitches.
+Similarly, duration of one time step is not defined by piano roll format and must be set externally too.
+
+Another nuance is that there are no attempts to resolve ambiguity for successive cells of the same pitch.
+If values of 1 are contained by $l$ successive cells of the same pitch, this line can be played either as one sound event that lasts $l$ time steps or as $l$ sound events that last one time step each or as $k$, $1 < k < n$, sound events of varying duration.
+Piano roll format does not specify it and evaluation of a roll is independent of ways to resolve it.
+
+Denote number of rows as $n$ and number of time steps as $m$ and let piano roll be represented as matrix $R$ of size $n \times m$.
+Suppose that current time step is $t$, $1 \le t \le m$.
+Then observation is defined as:
+$$v = \sum_{i = 1}^t \beta^{t - i} R_{\cdot i},$$
+where $\beta$ is a hyperparameter related to exponential decay, $0 < \beta < 1$.
+
+It is important to distinguish between time steps of piano roll (columns of $R$) and time steps of an episode (moments when actions are taken by an agent).
+For a particular piano roll's time step $t$ there are $a$ episode time steps where $a$ is a hyperparameter.
+At each of these episode time steps, a cell from the $t$-th column may be filled with 1 or nothing may happen (if a cell that already contains 1 is chosen for action).
+
+Agent takes action by drawing it from probability distribution returned by its actor model.
+An actor model can be every function that takes observation as input and returns probability distribution over cells from the current time step.
+Here, shallow neural network is used as actor model.
+Its weights can be viewed as something that defines transition matrix for a Markov chain from which actions are generated.
+
+An episode starts with a piano roll that contains only zeros.
+Initial observation is zero vector of size $n \times 1$.
+An episode is being run until final piano roll's time step is reached.
+Then created piano roll is evaluated and resulting score is returned as reward for the episode.
 
@@
 A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}.
-The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.1.1/}}.
+The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.1.2/}}.
 
@@
 \label{sec:improvements}
-% better evaluational rules (tonality, rhythm);
-% randomized initial observations;
+% impose known constraints natively instead of trying to learn them;
 % LSTM as actor model;
 % evolutionary strategies instead of cross-entropy;
-% start from supervisedly pre-trained weigths;
 

### 2019-10-24 docs/paper/paper.tex
docs: Start describing evaluational rules

@@
 
-\date{September 30, 2019}
+\date{Ocotber 26, 2019}
 
@@
 Here, shallow neural network is used as actor model.
-Its weights can be viewed as something that defines transition matrix for a Markov chain from which actions are generated.
+Its weights can be viewed as something that defines transition matrix for a Markov chain from which sequence of piano roll's states is generated.
 
@@
 \label{subsec:setup}
+
+Eight properties of a final piano roll are evaluated at the current revision of the study.
+
+\begin{itemize}
+	\item \textbf{Consonances.} A negative or positive score is granted for each pair of sounding cells from the same column, then all these scores are summed.
+Each pair-related score depends on interval between the two corresponding pitches.
+The more consonant the interval is, the higher the score is.
+However, there is an exception -- score is zero for intervals of integer number of octaves although they are perfectly consonant.
+This is done in order to prevent doubling the same melody in different lines.
+	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by the user before training.
+A negative penalty is assigned to each sounding cell of a pitch that does not belong to the specified scale.
+	\item \textbf{Conjunct motion.} A positive score is granted for every sounding cell such that the previous cell of the same pitch is not sounding, but there is a sounding cell of close enough pitch in the previous column.
+	\item \textbf{Contrary motion.} A positive score is granted for every column such that there are both a sounding cell of pitch lower than that of closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of closest to it sounding cell from the previous column.
+    \item \textbf{Noncyclicity.} From subject matter point of view, this score penalizes pieces where the same sequence of sound events is repeated over and over again.
+Technically, it is implemented as
+\end{itemize}
 

### 2019-10-26 docs/paper/paper.tex
docs: Finish writing about evaluational rules

--- 2019-10-22 docs/paper/paper.tex
@@
 
+Eight properties of a final piano roll are evaluated at the current revision of the study.
+
+\begin{itemize}
+	\item \textbf{Consonances.} Speaking formally, total reward for consonances is:
+	$$\sum_{t = 1}^m \sum_{i, j = 1}^n [i > j]R_{it} R_{jt} c((i - j) \bmod 12),$$
+	where square brackets are Iverson brackets and $c$ is a function that maps interval in semitones into measure of its consonance, $c: \{0, \dots, 11\} \to \mathbb{R}$.
+	The interpretation is as follows.
+Since $R_{it} \in \{0, 1\}$, only pairs of sounding cells from the same column affect the total sum.
+Value of each non-zero term depends on interval between the two corresponding pitches.
+The more consonant the interval is, the higher the value should be.
+Function $c$ must be chosen in a way providing that.
+However, there is an exception -- here, $c(0) = 0$ although intervals of whole number of octaves are perfectly consonant.
+This is done in order to prevent doubling the same melody in different octaves.
+	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by user before training.
+The same negative penalty is assigned to each sounding cell of a pitch that does not belong to the specified scale and then these penalties are summed.
+	\item \textbf{Usage of the tonic.} A positive reward is assigned to a roll as a whole if share of columns where the lowest sounding note belongs to the tonic pitch class lies within a specified interval.
+	\item \textbf{Conjunct motion.} A positive reward is granted for every sounding cell such that the previous cell of the same pitch is not sounding, but there is a sounding cell of close enough pitch in the previous column.
+	\item \textbf{Contrary motion.} A positive reward is granted for every column such that there are both a sounding cell of pitch lower than that of closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of closest to it sounding cell from the previous column.
+	\item \textbf{Lines.} Ideally, a created composition should be decomposable into melodic lines.
+To have $k$ melodic lines, it is necessary (but not sufficient) that exactly $k$ cells are sounding in every column.
+So below penalty may be used as a proxy for scoring presence of desired number of melodic lines:
+	$$-\sum_{t = 1}^m \left| k - \sum_{i=1}^n R_{it} \right|.$$  
+    \item \textbf{Noncyclicity.} Pieces where the same sequence of sound events is repeated over and over again should be penalized or, as a matter of choice, pieces should be rewarded for their noncyclicity.
+Here, the latter approach is used.
+The reward is implemented as minimum number of positive cells over differences between piano roll and its shifted version for all shift values from 1 up to a certain value (currently, it is equal to 8).
+To be more precise, this score is also clipped from above in order to prevent rewarding of placing extra sounding cells.
+In literature, scores based on autocorrelation are sometimes used as alternative way to push results away from degenerate states.
+    \item \textbf{Absence of pitches played for a long time.} Partial case of a degenerate roll is a roll with some pitches that are constantly sounding.
+A negative penalty is assigned to every sounding cell such that cells of the same pitch are sounding in all $l$ previous columns where $l$ is a hyperparameter.
+\end{itemize}
+
 \section{Experimental Results}
@@
 The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.1.2/}}.
+All important settings are placed to a separate configuration file and so it is easy to experiment with them.
 
@@
 % impose known constraints natively instead of trying to learn them;
+% structure (rhythm; start, end, motiff);
 % LSTM as actor model;

### 2019-11-02 docs/paper/paper.tex
docs: Conclude paper

@@
 Actually, results reported at this draft version are far from using above advantages at full scale.
-The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:improvements}.
+The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:conclusion}.
 
@@
 
-Since algorithmic composition can be framed as training of a generative model, it sounds natural to try one of the most salient examples of generative models -- generative adversarial networks (GAN) \cite{goodfellow2014gan}.
-However, widespread notations for music assume sequences of discrete values but classical GAN work well only with continuous data, because gradient of discrete-valued functions is uninformative.
+Since algorithmic composition can be framed as training of a generative model, it sounds natural to try one of the most salient examples of generative models -- generative adversarial networks (GANs) \cite{goodfellow2014gan}.
+However, widespread notations for music assume sequences of discrete values but classical GANs work well only with continuous data, because gradient of discrete-valued functions is uninformative.
 Techniques originating from reinforcement learning can be used to overcome this obstacle \cite{yu2016seqgan,hjelm2017boundary}.
@@
 	\textbf{Output:} $\hat{w}$ -- approximate solution to the problem $\max_w \mathbb{E}_{x \sim u(\cdot, w)} f(x)$. \\
-	\textbf{Hyperparameters:} $w_0$ -- initial value of $w$; $N$ -- number of iterations, $n$ -- number of vectors to draw at each iteration, $\sigma$ -- standard deviation for vectors generation, $m$ -- number of trials for each vector; $\rho$ -- fraction of best vectors to use for update; $\alpha$ -- smoothing coefficient of updates.
+	\textbf{Hyperparameters:} $w^{(0)}$ -- initial value of $w$; $N$ -- number of iterations, $n$ -- number of vectors to draw at each iteration, $\sigma$ -- standard deviation for vectors generation, $m$ -- number of trials for each vector; $\rho$ -- fraction of best vectors to use for update; $\alpha$ -- smoothing coefficient of updates.
 	\begin{algorithmic}[1]
@@
 		    \FORALL{$j \in \{1, \dots, n\}$}
-		        \STATE{draw $w_{ij} \sim \mathcal{N}(\cdot \vert w_{i-1}, \sigma)$}
-		        \STATE{$r_j \gets \sum_{k = 1}^{m} f(x_k)$ where $x_k \sim u(\cdot, w_{ij})$}
+		        \STATE{draw $w^{(i,j)} \sim \mathcal{N}(\cdot \vert w^{(i-1)}, \sigma)$}
+		        \STATE{$r_j \gets \sum_{k = 1}^{m} f(x_k)$ where $x_k \sim u(\cdot, w^{(i,j)})$}
 		    \ENDFOR
@@
 		    \STATE{$J \gets \{j: r_j \ge r_{\mathrm{threshold}}\}$}
-		    \STATE{$w_i \gets \alpha w_{i-1} + (1 - \alpha)(\sum_{j \in J} w_{ij}) / [\rho n]$}
+		    \STATE{$w^{(i)} \gets \alpha w^{(i-1)} + (1 - \alpha)(\sum_{j \in J} w^{(i,j)}) / [\rho n]$}
 		\ENDFOR
-	    \STATE{$\hat{w} \gets w_N$}
+	    \STATE{$\hat{w} \gets w^{(N)}$}
 	\end{algorithmic}
@@
 Another nuance is that there are no attempts to resolve ambiguity for successive cells of the same pitch.
-If values of 1 are contained by $l$ successive cells of the same pitch, this line can be played either as one sound event that lasts $l$ time steps or as $l$ sound events that last one time step each or as $k$, $1 < k < n$, sound events of varying duration.
+If values of 1 are contained by $l$ successive cells of the same pitch, this line can be played either as one sound event that lasts $l$ time steps or as $l$ sound events that last one time step each or as $k$, $1 < k < l$, sound events of varying duration.
 Piano roll format does not specify it and evaluation of a roll is independent of ways to resolve it.
@@
 
-\section{Further Improvements}
-\label{sec:improvements}
-% impose known constraints natively instead of trying to learn them;
-% structure (rhythm; start, end, motiff);
-% LSTM as actor model;
-% evolutionary strategies instead of cross-entropy;
+To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (note that the last time step is silent by design, so only 32 time steps are filled).
+Training an agent from scratch on a CPU of a regular laptop takes less than an hour with these settings.
+
+As for human evaluation, created pieces lack rhythmic and tonal structure.
+Such drawback can be expected a priori, because there are no evaluational rules for structure-related properties.
+On the other hand, pieces are quite ear-pleasant.
+Thus, current version of the tool may be used by a human composer as an auxiliary source of inspiration.
+In particular, generated short parts further can be inserted into a longer piece.
 
@@
 \label{sec:conclusion}
+
+This paper studies capabilities of reinforcement learning for music composition without usage of existing pieces.
+Here, music composition is framed as optimization problem where the goal is to find values of generative model's parameters maximizing average score of pieces generated with them.
+It is found that some progress can be made in this direction.
+
+However, it is also found that vast majority of computational resources and engineering efforts are now spent on preventing an agent from taking actions that are discouraged by music theory.
+It is much easier to impose known constraints in program code than to learn them from environment responses.
+So it seems very promising to exclude discouraged actions and then focus training on things of higher level such as rhythm dynamic, presence of motiffs, proper final, and so on.
+
+If setup become more complicated after above modifications, it is worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
 

### 2019-11-03 docs/paper/paper.tex
docs: Fix minor issues

@@
 
-\date{Ocotber 26, 2019}
+\date{November 2, 2019}
 
@@
 
-\keywords{algorithmc composition \and music generation \and reinforcement learning}
+\keywords{algorithmic composition \and music generation \and reinforcement learning}
 
@@
 Currently, there are no perfect tools for algorithmic composition.
-Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that can not be distinguished from works of a talented human composer.
+Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer.
 Thus, there is an open research problem.
@@
 Agent takes action by drawing it from probability distribution returned by its actor model.
-An actor model can be every function that takes observation as input and returns probability distribution over cells from the current time step.
+An actor model can be any function that takes observation as input and returns probability distribution over cells from the current time step.
 Here, shallow neural network is used as actor model.
@@
 This is done in order to prevent doubling the same melody in different octaves.
-	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by user before training.
+	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by a user before training.
 The same negative penalty is assigned to each sounding cell of a pitch that does not belong to the specified scale and then these penalties are summed.
@@
 	\item \textbf{Conjunct motion.} A positive reward is granted for every sounding cell such that the previous cell of the same pitch is not sounding, but there is a sounding cell of close enough pitch in the previous column.
-	\item \textbf{Contrary motion.} A positive reward is granted for every column such that there are both a sounding cell of pitch lower than that of closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of closest to it sounding cell from the previous column.
+	\item \textbf{Contrary motion.} A positive reward is granted for every column such that there are both a sounding cell of pitch lower than that of the closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of the closest to it sounding cell from the previous column.
 	\item \textbf{Lines.} Ideally, a created composition should be decomposable into melodic lines.
@@
 
-To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (note that the last time step is silent by design, so only 32 time steps are filled).
+To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (recall that the last time step is silent by design, so only 32 time steps are filled).
 Training an agent from scratch on a CPU of a regular laptop takes less than an hour with these settings.
@@
 
-If setup become more complicated after above modifications, it is worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
+If setup becomes more complicated after above modifications, it might be worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
 

### 2020-01-11 docs/paper/paper.tex
docs: Start updating of paper draft

@@
 
-\date{November 2, 2019}
+\date{January 11, 2020}
 
@@
 \begin{abstract}
-The notion of creativity is wider than generating something inspired by reference pieces of art.
-It also includes generation of something that meets criteria of being an art, but is not backed by existing pieces.
-On the one hand, supervised machine learning is limited by the former scope, because it requires a dataset.
-Reinforcement learning can be applied instead if the goal is to reveal more aspects of creativity.
-On the other hand, numerous recent advances in music generation are achieved with supervisedly trained neural networks.
-In this paper, a method that involves both reinforcement learning and neural networks is developed for automated music composition.
-An agent with shallow neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment.
-The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory.
-Cross-entropy method is used for training instead of gradient-based methods and so no datasets are needed.
+In this paper, automatic composition of first species counterpoint is framed as an optimization problem close to reinforcement learning.
+An agent with shallow neural network as actor model composes polyphonic pieces sequentially one measure (bar) per time step.
+Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory.
+Cross-entropy method is used for searching weights of actor model that result in maximum expected score of a piece.
+
+Although current setup of the problem may seem simplistic, it is a step in a promising direction.
+Numerous recent advances in music generation are achieved with supervisedly trained neural networks and reinforcement learning is not used there as a standalone paradigm.
+Since collections of existing pieces are involved as datasets, results are "inspired" by these reference pieces.
+However, the notion of creativity is wider -- it also includes generation of something that meets criteria of being an art, but is not backed by existing pieces.
+Thus, new aspects of creativity can be revealed by further research of pure reinforcement learning approaches.
 \end{abstract}
 
-\keywords{algorithmic composition \and music generation \and reinforcement learning}
+\keywords{algorithmic composition \and music generation \and reinforcement learning \and species counterpoint}
 
@@
 
-Brief outline of the approach is as follows.
-There is an environment with 2D table representing piano roll which means that rows of the table correspond to notes and columns of the table correspond to time steps.
-A cell of the table contains 1 if the note is sounding at this step and 0 otherwise.
-An agent observes vector in which for every row there is an exponential moving average over all time steps up to the current time step (inclusively).
-An action of the agent is playing a note at the current time step.
-After certain number of actions are done, a movement to the next time step happens.
-When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition.
-Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions.
+The approach relates to a well-known framework for simplified music composition called species counterpoint.
+It is often used by beginning composers in order to practice in voice leading and polyphony.
+From automatic composition point of view, species counterpoint is interesting due to two its properties:
+\begin{itemize}
+	\item There are rules prescribing what to do and what not to do. So there are less degrees of freedom and the problem is computationally more feasible.
+	\item There are clear hints of what is better to avoid and what should be added. So formal rules for automatic evaluation of resulting piece can be introduced.
+\end{itemize}
+
+The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern.
+Here, first species counterpoint is covered.
+This type has trivial rhythmic structure such that only whole (semibreve) notes can be used and every note must be placed in a beginning of a measure.
+Thus, interaction between rhythm and tonality is eliminated and the problem becomes even more simpler.
+
+Brief outline of the setup is as follows.
+There is an environment that keeps a piece in progress and the piece consists of a fixed number of measures.
+Each measure must be either empty (if it is not filled yet) or containing exactly one note per a melodic line.
+Initially, only the first measure and the last measure are filled with user-defined values.
+At a particular time step of an episode, an agent observes vector in which for every pitch there is an exponential moving average over all measures up to the current one (exclusively) of binary indicators whether this pitch is played at corresponding measure.
+An action of the agent is filling of the current measure, i.e., adding a pitch to each of the melodic lines.
+When the last measure is reached, episode is ended and the agent receives reward based on evaluation of created composition.
+Trainable parameters of the agent are weights of a neural network used as a so called actor model.
 These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).
@@
 Obviously, the approach belongs to reinforcement learning, because there is an environment, but a dataset is absent.
-Nevertheless, there are strong connections with supervised deep learning.
-From generation of new pieces point of view, actor model returns probabilities of next notes given current state.
+Nevertheless, there are some connections with supervised deep learning.
+From generation of new pieces point of view, actor model returns probabilities of next sonorities (chords) given current state.
 This is similar to sampling new sequences from a next-step prediction model trained with maximum likelihood method on a dataset of existing sequences.
@@
 
-Actually, results reported at this draft version are far from using above advantages at full scale.
+Actually, results reported at this draft version are far from using above advantages at full scale (in particular, first species counterpoint is too limiting).
 The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:conclusion}.
@@
 Usually, softmax activation function is used in the last layer of generative RNN.
-However, it is possible to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$.
+However, it is a common practice in reinforcement learning domain to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$.
 Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$).
@@
 
-For this study, piano roll is chosen as music representation format, but there are some subtleties.
+%pieces are phrases.
+%define rules of motion.
 
-Here, each row of a table does not correspond to a particular note or pitch.
+There are $l$ melodic lines (usually, $l$ is 2 or 3).
+Each line is a list of pitches from a diatonic scale (i.e., only 7 notes per octave can be used).
+Line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds.
+Suppose that maximum allowed interval is $s$ scale degrees.
+If so, there are no more than $(2s + 1)^l$ options to fill the next measure given current measure.
+Actually, the number is smaller, because some options are filtered out by counterpoint rules.
+
+%For this study, piano roll is chosen as music representation format, but there are some subtleties.
+
+%Here, each row of a table does not correspond to a particular note or pitch.
 Relative axis is used instead of absolute axis.
@@
 
-Another nuance is that there are no attempts to resolve ambiguity for successive cells of the same pitch.
+%Another nuance is that there are no attempts to resolve ambiguity for successive cells of the same pitch.
 If values of 1 are contained by $l$ successive cells of the same pitch, this line can be played either as one sound event that lasts $l$ time steps or as $l$ sound events that last one time step each or as $k$, $1 < k < l$, sound events of varying duration.
@@
 
-It is important to distinguish between time steps of piano roll (columns of $R$) and time steps of an episode (moments when actions are taken by an agent).
+%It is important to distinguish between time steps of piano roll (columns of $R$) and time steps of an episode (moments when actions are taken by an agent).
 For a particular piano roll's time step $t$ there are $a$ episode time steps where $a$ is a hyperparameter.
@@
 
-Agent takes action by drawing it from probability distribution returned by its actor model.
+%Agent takes action by drawing it from probability distribution returned by its actor model.
 An actor model can be any function that takes observation as input and returns probability distribution over cells from the current time step.
@@
 
-An episode starts with a piano roll that contains only zeros.
+%An episode starts with a piano roll that contains only zeros.
 Initial observation is zero vector of size $n \times 1$.
@@
 
-Eight properties of a final piano roll are evaluated at the current revision of the study.
+Seven properties of a final piano roll are evaluated at the current revision of the study.
 
 \begin{itemize}
-	\item \textbf{Consonances.} Speaking formally, total reward for consonances is:
-	$$\sum_{t = 1}^m \sum_{i, j = 1}^n [i > j]R_{it} R_{jt} c((i - j) \bmod 12),$$
-	where square brackets are Iverson brackets and $c$ is a function that maps interval in semitones into measure of its consonance, $c: \{0, \dots, 11\} \to \mathbb{R}$.
-	The interpretation is as follows.
-Since $R_{it} \in \{0, 1\}$, only pairs of sounding cells from the same column affect the total sum.
-Value of each non-zero term depends on interval between the two corresponding pitches.
-The more consonant the interval is, the higher the value should be.
-Function $c$ must be chosen in a way providing that.
-However, there is an exception -- here, $c(0) = 0$ although intervals of whole number of octaves are perfectly consonant.
-This is done in order to prevent doubling the same melody in different octaves.
-	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by a user before training.
-The same negative penalty is assigned to each sounding cell of a pitch that does not belong to the specified scale and then these penalties are summed.
-	\item \textbf{Usage of the tonic.} A positive reward is assigned to a roll as a whole if share of columns where the lowest sounding note belongs to the tonic pitch class lies within a specified interval.
-	\item \textbf{Conjunct motion.} A positive reward is granted for every sounding cell such that the previous cell of the same pitch is not sounding, but there is a sounding cell of close enough pitch in the previous column.
-	\item \textbf{Contrary motion.} A positive reward is granted for every column such that there are both a sounding cell of pitch lower than that of the closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of the closest to it sounding cell from the previous column.
-	\item \textbf{Lines.} Ideally, a created composition should be decomposable into melodic lines.
-To have $k$ melodic lines, it is necessary (but not sufficient) that exactly $k$ cells are sounding in every column.
-So below penalty may be used as a proxy for scoring presence of desired number of melodic lines:
-	$$-\sum_{t = 1}^m \left| k - \sum_{i=1}^n R_{it} \right|.$$  
-    \item \textbf{Noncyclicity.} Pieces where the same sequence of sound events is repeated over and over again should be penalized or, as a matter of choice, pieces should be rewarded for their noncyclicity.
-Here, the latter approach is used.
-The reward is implemented as minimum number of positive cells over differences between piano roll and its shifted version for all shift values from 1 up to a certain value (currently, it is equal to 8).
-To be more precise, this score is also clipped from above in order to prevent rewarding of placing extra sounding cells.
-In literature, scores based on autocorrelation are sometimes used as alternative way to push results away from degenerate states.
-    \item \textbf{Absence of pitches played for a long time.} Partial case of a degenerate roll is a roll with some pitches that are constantly sounding.
-A negative penalty is assigned to every sounding cell such that cells of the same pitch are sounding in all $l$ previous columns where $l$ is a hyperparameter.
+	\item \textbf{Autocorrelation.} Pieces where the same sequence of sound events is repeated over and over again are worse than pieces with no repetitions.
+	\item \textbf{Entropy of pitch distribution.}
+	\item \textbf{Absence of looped pitches.} A pitch may be rearticulated, but not more than once in a row. 
+	\item \textbf{Absence of clashes.} 
+	\item \textbf{Motion.} 
+	\item \textbf{Correlation between lines.}
+    \item \textbf{Explicity of climax.}
 \end{itemize}
+
+It can be seen that the first three properties deal with non-triviality of results, next three properties deal with independence (distinguishability) of lines and the last property deals with melodic quality of each separate line.
+
+Reward is a weighted sum of above seven scores.
 
@@
 A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}.
-The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.1.2/}}.
+The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.2.0/}}.
 All important settings are placed to a separate configuration file and so it is easy to experiment with them.
 
-The implementation relies on some open-source tools \cite{brockman2016openai,chollet2015keras,oliphant2006guide,raffel2014intuitive,dong2018pypianoroll}.
+The implementation relies on some open-source tools \cite{brockman2016openai,chollet2015keras,oliphant2006guide,raffel2014intuitive}.
 
-To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (recall that the last time step is silent by design, so only 32 time steps are filled).
+%different runs are variations of the same piece.
+
+%To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (recall that the last time step is silent by design, so only 32 time steps are filled).
 Training an agent from scratch on a CPU of a regular laptop takes less than an hour with these settings.
 
-As for human evaluation, created pieces lack rhythmic and tonal structure.
+% As for human evaluation, created pieces lack rhythmic and tonal structure.
 Such drawback can be expected a priori, because there are no evaluational rules for structure-related properties.
@@
 This paper studies capabilities of reinforcement learning for music composition without usage of existing pieces.
-Here, music composition is framed as optimization problem where the goal is to find values of generative model's parameters maximizing average score of pieces generated with them.
+Here, music composition is framed as optimization problem where the goal is to find values of generative model's parameters maximizing expected score of pieces generated with them.
 It is found that some progress can be made in this direction.
 
-However, it is also found that vast majority of computational resources and engineering efforts are now spent on preventing an agent from taking actions that are discouraged by music theory.
+%However, it is also found that vast majority of computational resources and engineering efforts are now spent on preventing an agent from taking actions that are discouraged by music theory.
 It is much easier to impose known constraints in program code than to learn them from environment responses.
@@
 
-If setup becomes more complicated after above modifications, it might be worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
+%If setup becomes more complicated after above modifications, it might be worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
 

### 2020-01-12 docs/paper/paper.tex
docs: Update Setup section of a paper

@@
 
-%pieces are phrases.
-%define rules of motion.
+To start with, define representation of a musical piece.
+Suppose that counterpoint is written in $l$ melodic lines (parts, voices), where $l$ is usually 2 or 3.
+Suppose also that a diatonic scale (like C-major or A-minor) is chosen.
+Let each line be a list of pitches from the scale (i.e., only 7 notes per octave can be used and chromaticism is prohibited).
+Set length of all lines to the same number $m$, where $m$ is usually between 8 and 24.
+From music point of view, this means that a piece to be created is rather a phrase that can be inserted into a longer composition.
+Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.
+This way or that, define a piece as a matrix $P$ of size $l \times m$ where element $P_{ij}$ is a pitch played in $i$-th line in $j$-th measure.
 
-There are $l$ melodic lines (usually, $l$ is 2 or 3).
-Each line is a list of pitches from a diatonic scale (i.e., only 7 notes per octave can be used).
-Line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds.
-Suppose that maximum allowed interval is $s$ scale degrees.
+%Duration of a measure is not defined by this format and it must be set externally.
+
+Every line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds.
+Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 2$).
 If so, there are no more than $(2s + 1)^l$ options to fill the next measure given current measure.
 Actually, the number is smaller, because some options are filtered out by counterpoint rules.
+The rules enforced in this study are as follows:
+\begin{itemize}
+	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated);
+	\item If a line has submediant followed by leading tone, tonic must be used after leading tone; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
+	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad;
+	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority);
+	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant.
+\end{itemize}
 
-%For this study, piano roll is chosen as music representation format, but there are some subtleties.
+Above rules are absolute and an agent can not break them, because it selects pitches for the next measure only from options that are filtered according to them.
+If a set of such options is empty, episode terminates and an agent receives negative reward for dead end.
 
-%Here, each row of a table does not correspond to a particular note or pitch.
-Relative axis is used instead of absolute axis.
-All that is important is that each row is one semitone higher than the row right below it.
-Thus, a musical piece is given only up to transposition and user-defined value for a pivot note is required to define absolute pitches.
-Similarly, duration of one time step is not defined by piano roll format and must be set externally too.
-
-%Another nuance is that there are no attempts to resolve ambiguity for successive cells of the same pitch.
-If values of 1 are contained by $l$ successive cells of the same pitch, this line can be played either as one sound event that lasts $l$ time steps or as $l$ sound events that last one time step each or as $k$, $1 < k < l$, sound events of varying duration.
-Piano roll format does not specify it and evaluation of a roll is independent of ways to resolve it.
-
-Denote number of rows as $n$ and number of time steps as $m$ and let piano roll be represented as matrix $R$ of size $n \times m$.
-Suppose that current time step is $t$, $1 \le t \le m$.
-Then observation is defined as:
+If the set of options is not empty, agent needs an observation.
+Observation is provided in terms of so called piano roll format.
+Denote size of pitch range available to the agent as $n$ and let the piece be represented also as a matrix $R$ of size $n \times m$ where $R_{ij}$ is 1 if $i$-th pitch is played in any line in $j$-th measure and 0 otherwise.
+Denote current measure as $t$, $1 \le t \le m$.
+Then observation is defined as vector $v$ of length $n$:
 $$v = \sum_{i = 1}^t \beta^{t - i} R_{\cdot i},$$
@@
 
-%It is important to distinguish between time steps of piano roll (columns of $R$) and time steps of an episode (moments when actions are taken by an agent).
-For a particular piano roll's time step $t$ there are $a$ episode time steps where $a$ is a hyperparameter.
-At each of these episode time steps, a cell from the $t$-th column may be filled with 1 or nothing may happen (if a cell that already contains 1 is chosen for action).
+For each option, agent concatenates observation $v$ with vector $c$ of length $(2s + 1)l$ encoding this option.
+Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $c$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them.
+Concatenation of $v$ and $c$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score.
+After these scores are collected for all options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
+An action is drawn from this distribution and the agents takes it.
 
-%Agent takes action by drawing it from probability distribution returned by its actor model.
-An actor model can be any function that takes observation as input and returns probability distribution over cells from the current time step.
-Here, shallow neural network is used as actor model.
-Its weights can be viewed as something that defines transition matrix for a Markov chain from which sequence of piano roll's states is generated.
-
-%An episode starts with a piano roll that contains only zeros.
-Initial observation is zero vector of size $n \times 1$.
-An episode is being run until final piano roll's time step is reached.
-Then created piano roll is evaluated and resulting score is returned as reward for the episode.
+Remember that episode starts with the first measure and the last measure already filled.
+The agent fills empty measures one-by-one according to described above procedure.
+When the final measure is reached, episode ends, created piece is evaluated and the agent receives reward.
 
@@
 
-Seven properties of a final piano roll are evaluated at the current revision of the study.
+At the current revision of the study, seven properties of a final piece are evaluated separately and total reward is a weighted sum of these seven scores.
 
@@
 	\item \textbf{Autocorrelation.} Pieces where the same sequence of sound events is repeated over and over again are worse than pieces with no repetitions.
-	\item \textbf{Entropy of pitch distribution.}
-	\item \textbf{Absence of looped pitches.} A pitch may be rearticulated, but not more than once in a row. 
-	\item \textbf{Absence of clashes.} 
-	\item \textbf{Motion.} 
-	\item \textbf{Correlation between lines.}
-    \item \textbf{Explicity of climax.}
+To reflect this in reward, this score is used:
+	$$1 - \max_k \{|\mathrm{Autocorr}_k(R)| : 2 \le k \le 8\},$$
+	where $R$ is piano roll representation of the piece and $\mathrm{Autocorr}_k$ is autocorrelation function for lag $k$ (for matrix $R$ its autocorrelation is average autocorrelation of its rows).
+	\item \textbf{Entropy of pitch distribution.} To encourage usage of all pitches from a range corresponding to a line, there is a score equal to average over all lines actual entropy of pitch distribution within the line normalized by entropy of uniform distribution on the same set of pitches.
+	\item \textbf{Absence of looped pitches.} A pitch may be rearticulated, but not more than once in a row. Every excessive occurrence of the same pitch within a line adds constant negative reward.
+	\item \textbf{Absence of pitch or pitch class clashes.} According to species counterpoint methodology, lines must be as distinguishable from each other as possible.
+In particular, this implies that the same pitches or the same pitch classes should not sound simultaneously in different lines.
+What is more, unison or octave harmonic intervals are too stable and so they create illusion of false finish if they occur in intermediate measures.
+To take into account these considerations, there is a negative reward proportional to share of intervals with both pitches belonging to the same pitch class amongst all intervals between two simultaneously sounding pitches.
+Also there is extra negative reward proportional to share of unison intervals.
+	\item \textbf{Motion.} For every pair of lines every motion between two successive measures adds a reward that depends on a type of motion.
+Since contrary motion makes lines more distinguishable, reward for it is positive.
+Reward for oblique motion is zero, reward for similar motion is slightly negative and reward for parallel motion is negative, because parallel motion makes lines sounding like copies of each other.
+	\item \textbf{Correlation between lines.} Consider each line as a sequence of real numbers equal to pitch positions within a range of available pitches.
+Then half of difference of 1 and average over all pairs of lines correlation between two lines is returned as a reward for independence of lines.
+    \item \textbf{Explicity of climax.} Music theory suggests that good melodies must have goal-oriented motion and that such goal-orientedness may be achieved by having exactly one climax point.
+For each line, negative rewards are granted for each duplication of climax point and for not so high climax point.
 \end{itemize}
@@
 It can be seen that the first three properties deal with non-triviality of results, next three properties deal with independence (distinguishability) of lines and the last property deals with melodic quality of each separate line.
-
-Reward is a weighted sum of above seven scores.
 
@@
 %different runs are variations of the same piece.
+%pitch class clashes occur.
 

### 2020-01-17 docs/paper/paper.tex
docs: Make some small improvements

@@
 The approach relates to a well-known framework for simplified music composition called species counterpoint.
-It is often used by beginning composers in order to practice in voice leading and polyphony.
+It is often used by beginning composers in order to practice in voice leading (also known as part writing) and polyphony.
 From automatic composition point of view, species counterpoint is interesting due to two its properties:
@@
 Actually, the number is smaller, because some options are filtered out by counterpoint rules.
-The rules enforced in this study are as follows:
+The rules enforced in this study can be either rules of voice leading (part writing) or rules of harmony.
+
+Here, five rules of voice leading are imposed:
 \begin{itemize}
-	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated);
-	\item If a line has submediant followed by leading tone, tonic must be used after leading tone; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
-	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad;
-	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority);
+	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
+	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
+	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
+	\item Movement in the opposite direction must happen immediately after a skip in order to resolve tension associated with the skip;
+	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
+\end{itemize}
+In addition, there is one rule of harmony:
+\begin{itemize}
 	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant.
@@
 Concatenation of $v$ and $c$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score.
-After these scores are collected for all options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
+After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
 An action is drawn from this distribution and the agents takes it.
@@
 	$$1 - \max_k \{|\mathrm{Autocorr}_k(R)| : 2 \le k \le 8\},$$
-	where $R$ is piano roll representation of the piece and $\mathrm{Autocorr}_k$ is autocorrelation function for lag $k$ (for matrix $R$ its autocorrelation is average autocorrelation of its rows).
+	where $R$ is piano roll representation of the piece, $\mathrm{Autocorr}_k$ is autocorrelation function for lag $k$ (for matrix $R$ its autocorrelation is average autocorrelation of its rows) and maximum lag to consider is set to 8 based on common sense.
 	\item \textbf{Entropy of pitch distribution.} To encourage usage of all pitches from a range corresponding to a line, there is a score equal to average over all lines actual entropy of pitch distribution within the line normalized by entropy of uniform distribution on the same set of pitches.
-	\item \textbf{Absence of looped pitches.} A pitch may be rearticulated, but not more than once in a row. Every excessive occurrence of the same pitch within a line adds constant negative reward.
+	\item \textbf{Absence of looped pitches.} A stable pitch may be rearticulated, but not more than once in a row.
+Every excessive occurrence of the same pitch within a line adds constant negative reward.
 	\item \textbf{Absence of pitch or pitch class clashes.} According to species counterpoint methodology, lines must be as distinguishable from each other as possible.
@@
 
+%feature representation.
+
 This paper studies capabilities of reinforcement learning for music composition without usage of existing pieces.

### 2020-01-18 rlmusician/environment/__init__.py
refac: Separate counterpoint rules from line filling

@@ -5,9 +5,16 @@ Author: Nikolay Lysenko
 """
 
 
-from . import environment, piece, evaluation
+from . import environment, evaluation, piece, rules
 from .environment import CounterpointEnv
 from .piece import Piece
 
 
-__all__ = ['CounterpointEnv', 'Piece', 'environment', 'piece', 'evaluation']
+__all__ = [
+    'CounterpointEnv',
+    'Piece',
+    'environment',
+    'evaluation',
+    'piece',
+    'rules'
+]

### 2020-01-18 rlmusician/environment/environment.py

@@ -125,20 +125,20 @@ class CounterpointEnv(gym.Env):
         info = {'next_actions': self.valid_actions}
 
         # Compute `done`.
-        finish = self.piece.last_finished_measure == self.piece.n_measures - 2
+        finish = self.piece.last_finished_measure == self.piece.n_measures - 1
         no_more_actions = len(info['next_actions']) == 0
         done = finish or no_more_actions
 
         # Compute `reward`.
-        if no_more_actions:
-            reward = self.reward_for_dead_end
-        elif done:
+        if finish:
             reward = evaluate(
                 self.piece,
                 self.scoring_coefs,
                 self.scoring_fn_params,
                 self.verbose
             )
+        elif no_more_actions:
+            reward = self.reward_for_dead_end
         else:
             reward = 0
 

### 2020-01-18 rlmusician/environment/evaluation.py

@@ -171,23 +171,18 @@ def evaluate_motion_by_types(
         normalized weighted sum of scores granted for each movement
     """
     score = 0
-    for first_line, second_line in itertools.combinations(piece.lines, 2):
-        prev_first = first_line[0]
-        prev_second = second_line[0]
-        paired = zip(first_line[1:], second_line[1:])
-        for first, second in paired:
-            first_diff = first.relative_position - prev_first.relative_position
-            second_diff = second.relative_position - prev_second.relative_position
-            if first_diff == second_diff:
+    pairs = itertools.combinations(piece.passed_movements, 2)
+    for first_movements, second_movements in pairs:
+        paired = zip(first_movements, second_movements)
+        for first_movement, second_movement in paired:
+            if first_movement == second_movement:
                 score += parallel_coef
-            elif first_diff * second_diff > 0:
+            elif first_movement * second_movement > 0:
                 score += similar_coef
-            elif first_diff * second_diff == 0:
+            elif first_movement * second_movement == 0:
                 score += oblique_coef
             else:
                 score += contrary_coef
-            prev_first = first
-            prev_second = second
     n_lines = len(piece.lines)
     n_intervals = n_lines * (n_lines - 1) / 2 * (piece.n_measures - 1)
     score /= n_intervals

### 2020-01-18 rlmusician/environment/piece.py

@@ -14,11 +14,15 @@ Author: Nikolay Lysenko
 import datetime
 import itertools
 import os
-from typing import Any, Dict, List, NamedTuple, Optional
+from typing import Any, Dict, List, NamedTuple
 
 import numpy as np
 from sinethesizer.io.utils import get_note_to_position_mapping
 
+from rlmusician.environment.rules import (
+    get_voice_leading_rules_registry,
+    get_harmony_rules_registry
+)
 from rlmusician.utils import (
     create_events_from_piece,
     create_midi_from_piece,
@@ -38,11 +42,22 @@ class LineElement(NamedTuple):
     absolute_position: int
     relative_position: int
     is_from_tonic_triad: bool
-    allowed_movements: List[int]
+    permitted_movements: List[int]
 
 
 class Piece:
     """Musical piece compliant with some rules of counterpoint writing."""
+
+    names_of_voice_leading_rules = [
+        'rearticulation',
+        'skip_goal',
+        'turn_after_skip',
+        'two_unstable',
+        'step_motion_to_end'
+    ]
+    names_of_harmony_rules = [
+        'consonance'
+    ]
 
     def __init__(
             self,
@@ -90,26 +105,25 @@ class Piece:
         self.lines = []
         self.line_elements = []
         self.line_mappings = []
+        self.passed_movements = []
         for specs in line_specifications:
             self.lines.append([None for _ in range(self.n_measures)])
+            self.passed_movements.append([])
             self.__define_elements(specs)
             self.__update_range_to_show(specs)
             self.__add_end_note(specs['start_note'], 'start')
             self.__add_end_note(specs['end_note'], 'end')
         self.last_finished_measure = 0
 
-    def __get_allowed_movements(
+    def __get_permitted_movements(
             self, current_position: int, end_position: int,
-            is_from_tonic_triad: bool
     ) -> List[int]:
-        """Get all possible shifts in scale degrees from current position."""
-        allowed_movements = [
+        """Get intervals in scale degrees that do not go beyond the range."""
+        permitted_movements = [
             movement for movement in self.all_movements
             if 0 <= current_position + movement < end_position
-            # Only pitches from tonic triad may be rearticulated.
-            and (movement != 0 or is_from_tonic_triad)
         ]
-        return allowed_movements
+        return permitted_movements
 
     def __define_elements(self, specs: Dict[str, Any]) -> None:
         """Define list of pitches that can be used within a line."""
[…]

### 2020-01-18 rlmusician/environment/rules.py

@@ -1,0 +1,190 @@
+"""
+Check compliance with some rules of voice leading and harmony.
+
+Author: Nikolay Lysenko
+"""
+
+
+import itertools
+from typing import Callable, Dict, List, Optional
+
+
+def check_stability_of_rearticulated_pitch(
+        line: List[Optional['LineElement']], measure: int, movement: int,
+        **kwargs
+) -> bool:
+    """
+    Check that a pitch to be rearticulated (repeated) is stable.
+
+    :param line:
+        melodic line as list of pitches
+    :param measure:
+        last finished measure in a line
+    :param movement:
+        melodic interval in scale degrees for line continuation
+    :return:
+        indicator whether a movement is in accordance with the rule
+    """
+    if movement != 0:
+        return True
+    return line[measure].is_from_tonic_triad
+
+
+def check_that_skip_leads_to_stable_pitch(
+        line: List[Optional['LineElement']],
+        line_elements: List['LineElement'],
+        measure: int, movement: int, **kwargs
+) -> bool:
+    """
+    Check that a skip (leap) leads to a stable pitch.
+
+    :param line:
+        melodic line as list of pitches
+    :param line_elements:
+        list of pitches available for the line
+    :param measure:
+        last finished measure in a line
+    :param movement:
+        melodic interval in scale degrees for line continuation
+    :return:
+        indicator whether a movement is in accordance with the rule
+    """
+    if abs(movement) <= 1:
+        return True
+    next_position = line[measure].relative_position + movement
+    next_element = line_elements[next_position]
+    return next_element.is_from_tonic_triad
+
+
+def check_that_skip_is_followed_by_opposite_step_motion(
+        movement: int, previous_movements: List[int], **kwargs
+) -> bool:
+    """
+    Check that after a large skip there is a step motion in opposite direction.
+
+    :param movement:
+        melodic interval in scale degrees for line continuation
+    :param previous_movements:
+        list of previous movements
+    :return:
+        indicator whether a movement is in accordance with the rule
+    """
+    if len(previous_movements) == 0:
+        return True
+    previous_movement = previous_movements[-1]
+    if abs(previous_movement) <= 2:  # Skip of a second is not large enough.
+        return True
+    return movement == -previous_movement / abs(previous_movement)
+
+
+def check_resolution_of_submediant_and_leading_tone(
+        line: List[Optional['LineElement']], measure: int,
+        movement: int, previous_movements: List[int], **kwargs
+) -> bool:
+    """
[…]

### 2020-01-18 tests/environment/test_evaluation.py

@@ -335,11 +335,11 @@ def test_evaluate_absence_of_pitch_class_clashes(
             # `similar_coef`
             -0.5,
             # `oblique_coef`
-            0,
+            -0.1,
             # `contrary_coef`
             1,
             # `expected`
-            -0.125
+            -0.15
         ),
     ]
 )

### 2020-01-18 tests/environment/test_piece.py

@@ -13,7 +13,7 @@ import pytest
 from rlmusician.environment import Piece
 
 
-class TestPianoRollEnv:
+class TestPiece:
     """Tests for `Piece` class."""
 
     @pytest.mark.parametrize(
@@ -246,6 +246,42 @@ class TestPianoRollEnv:
                 # `expected`
                 [True, False, False, False]
             ),
+            (
+                # `tonic`
+                'C',
+                # `scale`
+                'major',
+                # `n_measures`
+                10,
+                # `max_skip`
+                5,
+                # `line_specifications`
+                [
+                    {
+                        'lowest_note': 'G3',
+                        'highest_note': 'G4',
+                        'start_note': 'C4',
+                        'end_note': 'C4'
+                    },
+                    {
+                        'lowest_note': 'C4',
+                        'highest_note': 'C6',
+                        'start_note': 'C5',
+                        'end_note': 'G4'
+                    },
+                ],
+                # `previous_movements`
+                [[1, 4]],
+                # `candidate_movements`
+                [
+                    [-1, -1],
+                    [1, -2],
+                    [1, 1],
+                    [1, -1],
+                ],
+                # `expected`
+                [True, False, False, False]
+            ),
         ]
     )
     def test_check_movements(
@@ -305,7 +341,6 @@ class TestPianoRollEnv:
                     [0, -1],
                     [-2, 1],
                     [1, 1],
-                    [-1, 1]
                 ],
                 # `expected_positions`
                 [
@@ -353,8 +388,8 @@ class TestPianoRollEnv:
         piece = Piece(
             tonic, scale, n_measures, max_skip, line_specifications, {}
         )
-        for mov in movements:
-            piece.add_measure(mov)
+        for movement in movements:
+            piece.add_measure(movement)
         relative_positions = [
             [element.relative_position for element in line]
             for line in piece.lines

### 2020-01-18 rlmusician/environment/evaluation.py
feat: Evaluate interestingness of line based on number of skips

@@ -249,6 +249,32 @@ def evaluate_climax_explicity(
     return score
 
 
+def evaluate_number_of_skips(
+        piece: Piece, min_n_skips: int = 1, max_n_skips: int = 3
+) -> float:
+    """
+    Evaluate interestingness of lines based on number of skips in them.
+
+    :param piece:
+        Piece` instance
+    :param min_n_skips:
+        minimum number of skips for a line to be interesting
+    :param max_n_skips:
+        maximum number of skips for a line to be still coherent
+    :return:
+        share of lines where number of skips lies within specified range
+    """
+    scores = []
+    for movements in piece.passed_movements:
+        n_skips = 0
+        for movement in movements:
+            if abs(movement) > 1:
+                n_skips += 1
+        scores.append(1 if min_n_skips <= n_skips <= max_n_skips else 0)
+    score = sum(scores) / len(scores)
+    return score
+
+
 def get_scoring_functions_registry() -> Dict[str, Callable]:
     """
     Get mapping from names of scoring functions to scoring functions.
@@ -264,6 +290,7 @@ def get_scoring_functions_registry() -> Dict[str, Callable]:
         'types_of_motion': evaluate_motion_by_types,
         'lines_correlation': evaluate_lines_correlation,
         'climax_explicity': evaluate_climax_explicity,
+        'number_of_skips': evaluate_number_of_skips,
     }
     return registry
 

### 2020-01-18 tests/environment/test_evaluation.py

@@ -16,7 +16,8 @@ from rlmusician.environment.evaluation import (
     evaluate_absence_of_pitch_class_clashes,
     evaluate_motion_by_types,
     evaluate_lines_correlation,
-    evaluate_climax_explicity
+    evaluate_climax_explicity,
+    evaluate_number_of_skips
 )
 from rlmusician.environment.piece import Piece
 
@@ -494,3 +495,55 @@ def test_evaluate_climax_explicity(
         piece, shortage_penalty, duplication_penalty
     )
     assert round(result, 4) == expected
+
+
+@pytest.mark.parametrize(
+    "piece, all_movements, min_n_skips, max_n_skips, expected",
+    [
+        (
+            # `piece`
+            Piece(
+                tonic='C',
+                scale='major',
+                n_measures=5,
+                max_skip=2,
+                line_specifications=[
+                    {
+                        'lowest_note': 'G3',
+                        'highest_note': 'G4',
+                        'start_note': 'C4',
+                        'end_note': 'G4'
+                    },
+                    {
+                        'lowest_note': 'G4',
+                        'highest_note': 'G5',
+                        'start_note': 'C5',
+                        'end_note': 'C5'
+                    },
+                ],
+                rendering_params={}
+            ),
+            # `all_movements`,
+            [
+                [2, 0],
+                [0, 0],
+                [2, 0],
+            ],
+            # `min_n_skips`
+            1,
+            # `max_n_skips`
+            2,
+            # `expected`
+            0.5
+        ),
+    ]
+)
+def test_evaluate_number_of_skips(
+        piece: Piece, all_movements: List[List[int]],
+        min_n_skips: int, max_n_skips: int, expected: float
+) -> None:
+    """Test `evaluate_number_of_skips` function."""
+    for movements in all_movements:
+        piece.add_measure(movements)
+    result = evaluate_number_of_skips(piece, min_n_skips, max_n_skips)
+    assert result == expected

### 2020-01-19 rlmusician/environment/piece.py
feat: Prohibit harmonic intervals larger than tenth

@@ -56,7 +56,8 @@ class Piece:
         'step_motion_to_end'
     ]
     names_of_harmony_rules = [
-        'consonance'
+        'consonance',
+        'absence_of_large_intervals'
     ]
 
     def __init__(

### 2020-01-19 rlmusician/environment/rules.py

@@ -177,6 +177,27 @@ def check_consonance_of_sonority(sonority: List['LineElement']) -> bool:
     return True
 
 
+def check_absence_of_large_intervals(
+        sonority: List['LineElement'], max_interval: int = 9
+) -> bool:
+    """
+    Check that there are no large intervals between adjacent pitches.
+
+    :param sonority:
+        list of simultaneously sounding pitches
+    :param max_interval:
+        maximum allowed interval in scale degrees between two adjacent pitches
+    :return:
+        indicator whether a sonority has no excessive intervals
+    """
+    if len(sonority) == 1:
+        return True
+    positions = sorted(x.relative_position for x in sonority)
+    adjacent_pairs = zip(positions, positions[1:])
+    intervals = [y - x for x, y in adjacent_pairs]
+    return max(intervals) <= max_interval
+
+
 def get_harmony_rules_registry() -> Dict[str, Callable]:
     """
     Get mapping from names to functions checking harmony rules.
@@ -185,6 +206,7 @@ def get_harmony_rules_registry() -> Dict[str, Callable]:
         registry of functions checking harmony rules
     """
     registry = {
-        'consonance': check_consonance_of_sonority
+        'consonance': check_consonance_of_sonority,
+        'absence_of_large_intervals': check_absence_of_large_intervals
     }
     return registry

### 2020-01-19 rlmusician/__main__.py
refac: Update default settings

@@ -62,7 +62,7 @@ def parse_cli_args() -> argparse.Namespace:
         help='path to configuration file'
     )
     parser.add_argument(
-        '-p', '--populations', type=int, default=10,
+        '-p', '--populations', type=int, default=15,
         help='number of populations for agent training'
     )
     parser.add_argument(

### 2020-01-21 docs/paper/paper.tex
docs: Update paper draft

@@
 Trainable parameters of the agent are weights of a neural network used as a so called actor model.
-These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).
+These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms or evolutionary strategies \cite{salimans2017evolution}).
+From optimization perspective, average over several episodes reward is estimated value of fitness function at the weights used.
 
-Obviously, the approach belongs to reinforcement learning, because there is an environment, but a dataset is absent.
+Obviously, the approach belongs to reinforcement learning, because there is an environment and an agent, but a dataset is absent.
 Nevertheless, there are some connections with supervised deep learning.
@@
 
-More details on the methodology are provided in Section \ref{sec:methodology}, but here it is appropriate to discuss its advantages.
-The reasons for not involving supervised training at all are as follows:
+More details on the methodology are provided in Section \ref{sec:methodology} but here it is appropriate to discuss its advantages.
+Of course, automatic composition of species counterpoint is not novel -- some researchers have developed optimization-based approaches not only for first species, but also for fifth species \cite{herremans2012composing}.
+Nevertheless, as far as the author knows, such approaches yield a single piece as a result of optimization, whereas in the current approach training produces an agent that creates a new piece after each run.
+
+Next-step prediction models create a new piece after each run too, but they are trained supervisedly. The reasons for not involving supervised training at all are as follows:
 \begin{itemize}
@@
 The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:conclusion}.
+% Chances are that a way of reinforcement learning combines strengths of fitness function optimization with strengths of supervised learning.
 
@@
 In context of reinforcement learning, cross-entropy method is used for searching parameters of agent that result in maximum expected reward.
-Such approach is best suitable for environments where reward becomes known only after episode is finished.
+So $X$ is a set of all possible finished episodes, $f$ is reward, and $u(\cdot, w)$ describes how an episode depends on parameters.
+Cross-entropy method is best suitable for environments where reward becomes known only after episode is finished.
 If there are rewards granted after intermediate steps, methods based on Q-values usually outperform it.
@@
 
-%Duration of a measure is not defined by this format and it must be set externally.
-
 Every line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds.
@@
 \end{itemize}
-In addition, there is one rule of harmony:
+In addition, there are two rules of harmony:
 \begin{itemize}
-	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant.
+	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
+	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
 \end{itemize}
@@
 
-For each option, agent concatenates observation $v$ with vector $c$ of length $(2s + 1)l$ encoding this option.
-Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $c$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them.
-Concatenation of $v$ and $c$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score.
+For each option, agent concatenates observation $v$ with vector $u$ of length $(2s + 1)l$ encoding this option.
+Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them.
+Concatenation of $v$ and $u$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score.
 After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
@@
 
-At the current revision of the study, seven properties of a final piece are evaluated separately and total reward is a weighted sum of these seven scores.
+At the current revision of the study, eight properties of a final piece are evaluated separately and total reward is a weighted sum of these eight scores.
 
 \begin{itemize}
-	\item \textbf{Autocorrelation.} Pieces where the same sequence of sound events is repeated over and over again are worse than pieces with no repetitions.
-To reflect this in reward, this score is used:
+	\item \textbf{Autocorrelation.} Pieces where the same sequence of sounds is repeated over and over again are worse than pieces with no repetitions. To reflect this in reward, this score is used:
 	$$1 - \max_k \{|\mathrm{Autocorr}_k(R)| : 2 \le k \le 8\},$$
@@
 For each line, negative rewards are granted for each duplication of climax point and for not so high climax point.
+    \item \textbf{Number of skips.} Positive reward is associated with every line such that number of skips in it lies within a specified range.
+The motivation behind this is that lines with almost no skips are not interesting, but lines with too many skips are not coherent.
 \end{itemize}
 
-It can be seen that the first three properties deal with non-triviality of results, next three properties deal with independence (distinguishability) of lines and the last property deals with melodic quality of each separate line.
+It can be seen that the first three properties deal with non-triviality of results, the next three properties deal with interaction of lines and the last two properties deal with melodic quality of each separate line.
 
@@
 %feature representation.
+%music theory: penalize downward skips and overlapping motion.
 

### 2020-01-23 docs/paper/paper.tex
docs: Start writing about experimental results

@@
 
+In this section, results are reported for experiments with 2 melodic lines and 16 measures starting from (G4, C5) and ending with (C4, C5).
+It is found that scores for autocorrelation and entropy of pitch distribution are redundant.
+Non-degenerate pieces are produced even without them.
+So only six properties are evaluated.
+There are weighted so that explicity of climax and absence of pitch class clashes have 3 times more impact than any other property.
+
+In Table \ref{table:scores}, scores of a typical piece created by an agent trained for 20 populations, are compared with two benchmarks.
+The first one is simple baseline -- it is score of a piece created by an agent with weights that are the best amongst 100 random weights.
+The second one are scores for a piece manually created by the author.
+
+\begin{table}[h!]
+	\caption{Scores of sample pieces}
+	\label{table:scores}
+	\begin{center}
+		\begin{tabular}{|l|c|c|c|c|}
+			\hline
+			& Baseline & Agent & Human & Maximum\\
+			\hline
+			Absence of looped pitches & & & 0 & 0\\
+			\hline
+			Absence of pitch or pitch class clashes & & & -0.643 & 0\\
+			\hline
+			Motions & & & 0.1 & 1\\
+			\hline
+			Correlation between lines & & & 0.278 & 1\\
+			\hline
+			Explicity of climax & & & 3 & 3\\
+			\hline
+			Number of skips & & & 1 & 1\\
+			\hline
+			Total reward & & & 3.735 & 6\\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{table}
+
+It can be seen that .
 %different runs are variations of the same piece.
@@
 
-%To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (recall that the last time step is silent by design, so only 32 time steps are filled).
-Training an agent from scratch on a CPU of a regular laptop takes less than an hour with these settings.
+As for subjective evaluation of music quality, it is not correct to compare the selected piece created by an agent with the piece created by the author.
+The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically.
+As a result, subjective perception of a resulting piece is not aligned with objective function used during optimization.
 
-% As for human evaluation, created pieces lack rhythmic and tonal structure.
-Such drawback can be expected a priori, because there are no evaluational rules for structure-related properties.
-On the other hand, pieces are quite ear-pleasant.
-Thus, current version of the tool may be used by a human composer as an auxiliary source of inspiration.
-In particular, generated short parts further can be inserted into a longer piece.
+Table \ref{table:piece} contains the created by an agent piece for which above scores are granted.
+
+\begin{table}[h!]
+	\caption{A piece composed by an agent}
+	\label{table:piece}
+	\begin{center}
+		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{table}
 
@@
 
-%However, it is also found that vast majority of computational resources and engineering efforts are now spent on preventing an agent from taking actions that are discouraged by music theory.
-It is much easier to impose known constraints in program code than to learn them from environment responses.
-So it seems very promising to exclude discouraged actions and then focus training on things of higher level such as rhythm dynamic, presence of motiffs, proper final, and so on.
-
-%If setup becomes more complicated after above modifications, it might be worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
-
 \bibliographystyle{unsrt}  

### 2020-01-26 docs/paper/paper.tex
docs: Write conclusion

@@
 
-\date{January 11, 2020}
+\date{January 26, 2020}
 
@@
 			\hline
-			Absence of looped pitches & & & 0 & 0\\
+			Absence of looped pitches & -1 & 0 & 0 & 0\\
 			\hline
-			Absence of pitch or pitch class clashes & & & -0.643 & 0\\
+			Absence of pitch or pitch class clashes & -0.429 & -0.429 & -0.643 & 0\\
 			\hline
-			Motions & & & 0.1 & 1\\
+			Motion & -0.1 & 0.367 & 0.1 & 1\\
 			\hline
-			Correlation between lines & & & 0.278 & 1\\
+			Correlation between lines & 0.552 & 0.599 & 0.278 & 1\\
 			\hline
-			Explicity of climax & & & 3 & 3\\
+			Explicity of climax & 0.9 & 1.65 & 3 & 3\\
 			\hline
-			Number of skips & & & 1 & 1\\
+			Number of skips & 0 & 1 & 1 & 1\\
 			\hline
-			Total reward & & & 3.735 & 6\\
+			Total reward & -0.076 & 3.187 & 3.735 & 6\\
 			\hline
@@
 
-It can be seen that .
-%different runs are variations of the same piece.
-%pitch class clashes occur.
+It can be seen that the agent's piece is better than the manually created piece by all metrics except explicity of climax.
+Since climax is higher than start note, both lines are assumed to have upward motion but the agent wrongly trades off climax score for contrary motion and lower lines correlation.
 
-As for subjective evaluation of music quality, it is not correct to compare the selected piece created by an agent with the piece created by the author.
-The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically.
-As a result, subjective perception of a resulting piece is not aligned with objective function used during optimization.
-
-Table \ref{table:piece} contains the created by an agent piece for which above scores are granted.
+Table \ref{table:piece} contains the created by the agent piece for which above scores are granted.
 
@@
 		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
+			\hline
+			& & & & & & & C6 & & & & & & & & \\
+			\hline
+			& & & & & & B5 & & & & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & & A5 & & & & A5 & & & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & & G5 & & & & G5 & & G5 & G5 & & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& & & F5 & & & & & & & & & & & & \\
+			\hline
+			& & E5 & & & & & & & & & & E5 & & & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			& D5 & & & & & & & & & & & & & D5 & \\
+			\hline
+			& & & & & & & & & & & & & & & \\
+			\hline
+			C5 & & & & & & & & & & & & & C5 & & C5 \\
 			\hline
@@
 			\hline
-			& & & & & & & & & & & & & & & \\
+			G4 & & & & & & & & & & & & & & & \\
 			\hline
@@
 			\hline
-			& & & & & & & & & & & & & & & \\
+			& F4 & & & & & & & & & & & & & & \\
+			\hline
+			& & E4 & & & E4 & & E4 & & & E4 & & & E4 & & \\
 			\hline
@@
 			\hline
-			& & & & & & & & & & & & & & & \\
+			& & & D4 & & & D4 & & D4 & & & D4 & & & D4 & \\
 			\hline
@@
 			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
+			& & & & C4 & & & & & C4 & & & C4 & & & C4 \\
 			\hline
@@
 \end{table}
+
+As for subjective evaluation of music quality, it is not correct to compare the selected piece created by the agent with the piece created by the author.
+The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically.
+That being said, subjective perception of a resulting piece is not aligned with fitness function used during optimization.
+However, it must be noted that the piece shown above looks like overfitted to the given fitness function.
+It is not a problem associated with training, it is a flaw in problem setup.
+
+Numerous experiments revealed another important issue.
+Actually, different runs of a trained agent produce variations of the same piece.
+Although there might be significant differences in scores (in particular, because some scoring functions are discrete), there are no significant differences from music point of view.
 
@@
 
+The list of suggested further improvements is as follows:
+\begin{itemize}
+	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged.
+Also downward skips can be penalized in order to have upward skips before climax and downward stepwise motion after it.
+	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes.
+However, a new term can be added to this average.
+This term must reflect how different created pieces are from each other.
+	\item \textbf{Revise feature representation for actor model.} Probably, procedure of next measure generation should be changed.
+Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation.
+To make it less raw, some features inspired by music theory may be engineered.
+	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches.
+Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates.
+Moments of time when level of tension changes may be rhythmically accented or unaccented.
+Close but not exact alignment between rhythmical accent and resolution results in aesthetic and interesting music.
+Taking into account voice leading constraints, optimization problem can be set up based on above considerations.
+\end{itemize}
+
 \bibliographystyle{unsrt}  

### 2020-01-26 rlmusician/environment/rules.py
fix: Prohibit excessively large intervals correctly

@@ -178,7 +178,7 @@ def check_consonance_of_sonority(sonority: List['LineElement']) -> bool:
 
 
 def check_absence_of_large_intervals(
-        sonority: List['LineElement'], max_interval: int = 9
+        sonority: List['LineElement'], max_interval: int = 16
 ) -> bool:
     """
     Check that there are no large intervals between adjacent pitches.
@@ -186,13 +186,13 @@ def check_absence_of_large_intervals(
     :param sonority:
         list of simultaneously sounding pitches
     :param max_interval:
-        maximum allowed interval in scale degrees between two adjacent pitches
+        maximum allowed interval in semitones between two adjacent pitches
     :return:
         indicator whether a sonority has no excessive intervals
     """
     if len(sonority) == 1:
         return True
-    positions = sorted(x.relative_position for x in sonority)
+    positions = sorted(x.absolute_position for x in sonority)
     adjacent_pairs = zip(positions, positions[1:])
     intervals = [y - x for x, y in adjacent_pairs]
     return max(intervals) <= max_interval

### 2020-01-26 docs/paper/paper.tex
docs: Finalize draft of a paper

@@
 In this paper, automatic composition of first species counterpoint is framed as an optimization problem close to reinforcement learning.
-An agent with shallow neural network as actor model composes polyphonic pieces sequentially one measure (bar) per time step.
+An agent with shallow neural network as an actor model composes polyphonic pieces sequentially one measure (bar) per time step.
 Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory.
@@
 Algorithmic music composition is automatic generation of outputs representing musical pieces and written in some formal notation.
-To name a few of common notations, there are sheet music, tablature, and MIDI standard.
+To name a few of common notations, there is sheet music, tablature, and MIDI standard.
 It is not required from output representation to unambiguously define sound waveform.
@@
 This type has trivial rhythmic structure such that only whole (semibreve) notes can be used and every note must be placed in a beginning of a measure.
-Thus, interaction between rhythm and tonality is eliminated and the problem becomes even more simpler.
+Thus, interaction between rhythm and tonality is eliminated and the problem becomes even simpler.
 
@@
 Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$).
-In \cite{kumar2019polyphonic}, Boltzmann softmax is used and also input vector is extended by introducing an additional part indicating origination of initial input (so called plan).
+In \cite{kumar2019polyphonic}, Boltzmann softmax is used and also input vector is extended by introducing an additional part (so called plan) indicating origination of initial input.
 An environment is defined so that state is a pair of temperature and plan, actions are changes in either temperature or plan, and reward depends on evaluation of produced with these settings piece.
@@
 So $X$ is a set of all possible finished episodes, $f$ is reward, and $u(\cdot, w)$ describes how an episode depends on parameters.
-Cross-entropy method is best suitable for environments where reward becomes known only after episode is finished.
+Cross-entropy method is best suitable for environments where reward becomes known only after an episode is finished.
 If there are rewards granted after intermediate steps, methods based on Q-values usually outperform it.
@@
 Every line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds.
-Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 2$).
+Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 3$).
 If so, there are no more than $(2s + 1)^l$ options to fill the next measure given current measure.
@@
 	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
-	\item Movement in the opposite direction must happen immediately after a skip in order to resolve tension associated with the skip;
+	\item Movement in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
 	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
@@
 
-Above rules are absolute and an agent can not break them, because it selects pitches for the next measure only from options that are filtered according to them.
+Above rules are absolute and an agent cannot break them, because it selects pitches for the next measure only from options that are filtered according to them.
 If a set of such options is empty, episode terminates and an agent receives negative reward for dead end.
 
-If the set of options is not empty, agent needs an observation.
+If the set of options is not empty, an agent needs an observation.
 Observation is provided in terms of so called piano roll format.
@@
 
-For each option, agent concatenates observation $v$ with vector $u$ of length $(2s + 1)l$ encoding this option.
+For each option, agent concatenates observation $v$ with encoding this option vector $u$ of length $(2s + 1)l$.
 Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them.
@@
 After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
-An action is drawn from this distribution and the agents takes it.
+An action is drawn from this distribution and the agent takes it.
 
-Remember that episode starts with the first measure and the last measure already filled.
-The agent fills empty measures one-by-one according to described above procedure.
-When the final measure is reached, episode ends, created piece is evaluated and the agent receives reward.
+Remember that an episode starts with the first and last measures already filled.
+The agent fills empty measures one-by-one according to the described above procedure.
+When the final measure is reached, the episode ends, created piece is evaluated and the agent receives reward.
 
@@
 So only six properties are evaluated.
-There are weighted so that explicity of climax and absence of pitch class clashes have 3 times more impact than any other property.
+They are weighted so that explicity of climax and absence of pitch class clashes have 3 times more impact than any other property.
 
@@
 The first one is simple baseline -- it is score of a piece created by an agent with weights that are the best amongst 100 random weights.
-The second one are scores for a piece manually created by the author.
+The second one are scores for a piece manually created by the author in about 15 minutes.
 
@@
 			\hline
-			Absence of pitch or pitch class clashes & -0.429 & -0.429 & -0.643 & 0\\
+			Absence of pitch or pitch class clashes & -0.429 & -1.286 & -0.643 & 0\\
 			\hline
-			Motion & -0.1 & 0.367 & 0.1 & 1\\
+			Motion & -0.1 & -0.033 & 0.1 & 1\\
 			\hline
-			Correlation between lines & 0.552 & 0.599 & 0.278 & 1\\
+			Correlation between lines & 0.552 & 0.168 & 0.278 & 1\\
 			\hline
-			Explicity of climax & 0.9 & 1.65 & 3 & 3\\
+			Explicity of climax & 0.9 & 3 & 3 & 3\\
 			\hline
@@
 			\hline
-			Total reward & -0.076 & 3.187 & 3.735 & 6\\
+			Total reward & -0.076 & 2.849 & 3.735 & 6\\
 			\hline
@@
 
-It can be seen that the agent's piece is better than the manually created piece by all metrics except explicity of climax.
-Since climax is higher than start note, both lines are assumed to have upward motion but the agent wrongly trades off climax score for contrary motion and lower lines correlation.
-
-Table \ref{table:piece} contains the created by the agent piece for which above scores are granted.
+It can be seen that the agent's piece consists of two lines with formally good melodies but these lines are not independent enough.
+Table \ref{table:piece} contains the created by the agent piece for which above scores are reported.
+Visual analysis of the shown piano roll confirms observation about lines dependence -- their climax points coincide and there are plenty of intermediate octave intervals.
 
@@
 			\hline
-			& & & & G5 & & & & G5 & & G5 & G5 & & & & \\
+			& & & & G5 & & & & G5 & & & & & & & \\
 			\hline
@@
 			\hline
-			& & & F5 & & & & & & & & & & & & \\
+			& & & F5 & & & & & & & & F5 & & & & \\
 			\hline
-			& & E5 & & & & & & & & & & E5 & & & \\
+			& & E5 & & & & & & & & E5 & & & & & \\
 			\hline
@@
 			\hline
-			& D5 & & & & & & & & & & & & & D5 & \\
+			& D5 & & & & & & & & & & & & D5 & & \\
 			\hline
@@
 			\hline
-			C5 & & & & & & & & & & & & & C5 & & C5 \\
+			C5 & & & & & & & C5 & & & & & C5 & & C5 & C5 \\
+			\hline
+			& & & & & & & & B4 & & & & & & & \\
 			\hline
@@
 			\hline
-			& & & & & & & & & & & & & & & \\
+			& & & & & & & & & A4 & & & & & & \\
 			\hline
@@
 			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			G4 & & & & & & & & & & & & & & & \\
+			G4 & & & & G4 & & G4 & & & & G4 & & & & & \\
 			\hline
@@
 			\hline
-			& F4 & & & & & & & & & & & & & & \\
+			& F4 & & & & F4 & & & & & & F4 & & & & \\
 			\hline
-			& & E4 & & & E4 & & E4 & & & E4 & & & E4 & & \\
+			& & & & & & & & & & & & E4 & & & \\
 			\hline
@@
 			\hline
-			& & & D4 & & & D4 & & D4 & & & D4 & & & D4 & \\
+			& & & D4 & & & & & & & & & & D4 & & \\
 			\hline
@@
 			\hline
-			& & & & C4 & & & & & C4 & & & C4 & & & C4 \\
+			& & C4 & & & & & & & & & & & & C4 & C4 \\
 			\hline
@@
 That being said, subjective perception of a resulting piece is not aligned with fitness function used during optimization.
-However, it must be noted that the piece shown above looks like overfitted to the given fitness function.
-It is not a problem associated with training, it is a flaw in problem setup.
+However, it must be noted that the piece shown above looks like excessively oriented on maximization of fitness function regardless more subtle properties important for human listeners.
+It is not a problem associated with training, it is a flaw in problem setup (more criteria for evaluation are needed).
 
@@
 
-%feature representation.
-%music theory: penalize downward skips and overlapping motion.
-
 This paper studies capabilities of reinforcement learning for music composition without usage of existing pieces.
@@
 	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged.
-Also downward skips can be penalized in order to have upward skips before climax and downward stepwise motion after it.
+Also downward skips can be penalized in order to have upward skips before climax and downward conjunct motion after it.
 	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes.
@@
 To make it less raw, some features inspired by music theory may be engineered.
+	\item \textbf{Tune hyperparameters.} In the current software implementation, cross-entropy method has five hyperparameters, an agent has two hyperparameters, and evaluation functions have more than ten hyperparameters.
+As of now, not so many computational experiments are held in order to find their optimum values.
 	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches.
@@
 Moments of time when level of tension changes may be rhythmically accented or unaccented.
-Close but not exact alignment between rhythmical accent and resolution results in aesthetic and interesting music.
-Taking into account voice leading constraints, optimization problem can be set up based on above considerations.
+Close but not exact alignment between rhythmical accent and resolution often results in aesthetic and interesting music.
+Taking into account constraints of smooth voice leading, optimization problem can be set up based on above considerations.
 \end{itemize}

### 2020-01-26 tests/environment/test_environment.py
ci: Fix tests broken by rule of max harmonic interval

@@ -97,7 +97,7 @@ class TestCounterpointEnv:
                 # `actions`
                 [16, 6],
                 # `expected`
-                [6, 7, 9, 10, 12, 13, 14, 16, 18, 19, 20, 22, 24]
+                [6, 10, 12, 16, 18, 20, 22, 24]
             ),
         ]
     )

### 2020-01-26 tests/environment/test_piece.py

@@ -259,8 +259,8 @@ class TestPiece:
                 [
                     {
                         'lowest_note': 'G3',
-                        'highest_note': 'G4',
-                        'start_note': 'C4',
+                        'highest_note': 'G5',
+                        'start_note': 'G4',
                         'end_note': 'C4'
                     },
                     {
@@ -271,13 +271,13 @@ class TestPiece:
                     },
                 ],
                 # `previous_movements`
-                [[1, 4]],
+                [[0, 4]],
                 # `candidate_movements`
                 [
                     [-1, -1],
                     [1, -2],
                     [1, 1],
-                    [1, -1],
+                    [0, -1],
                 ],
                 # `expected`
                 [True, False, False, False]

### 2020-01-26 rlmusician/environment/rules.py
chore: Highlight a subtle nuance

@@ -178,7 +178,8 @@ def check_consonance_of_sonority(sonority: List['LineElement']) -> bool:
 
 
 def check_absence_of_large_intervals(
-        sonority: List['LineElement'], max_interval: int = 16
+        sonority: List['LineElement'],
+        max_interval: int = 16  # NB: Changing this value may break tests.
 ) -> bool:
     """
     Check that there are no large intervals between adjacent pitches.

### 2020-03-09 docs/paper/paper.tex
refac: Name policy properly, do not call it actor model

@@
 \begin{abstract}
-In this paper, automatic composition of first species counterpoint is framed as an optimization problem close to reinforcement learning.
-An agent with shallow neural network as an actor model composes polyphonic pieces sequentially one measure (bar) per time step.
+In this paper, automatic composition of first species counterpoint is framed as a black-box optimization problem close to reinforcement learning.
+An agent with shallow neural network as a policy composes polyphonic pieces sequentially one measure (bar) per time step.
 Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory.
-Cross-entropy method is used for searching weights of actor model that result in maximum expected score of a piece.
+Cross-entropy method is used for searching policy weights that result in maximum expected score of a piece.
 
@@
 When the last measure is reached, episode is ended and the agent receives reward based on evaluation of created composition.
-Trainable parameters of the agent are weights of a neural network used as a so called actor model.
-These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms or evolutionary strategies \cite{salimans2017evolution}).
+Trainable parameters of the agent are weights of a neural network used as a policy.
+These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with evolutionary strategies \cite{salimans2017evolution}, genetic algorithms, simulated annealing, or other black-box optimization methods).
 From optimization perspective, average over several episodes reward is estimated value of fitness function at the weights used.
@@
 Nevertheless, there are some connections with supervised deep learning.
-From generation of new pieces point of view, actor model returns probabilities of next sonorities (chords) given current state.
+From generation of new pieces point of view, a policy returns probabilities of next sonorities (chords) given current state.
 This is similar to sampling new sequences from a next-step prediction model trained with maximum likelihood method on a dataset of existing sequences.
@@
 Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them.
-Concatenation of $v$ and $u$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score.
+Concatenation of $v$ and $u$ is passed as input to a policy which is a shallow neural network returning a single real-valued score.
 After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
@@
 However, a new term can be added to this average.
-This term must reflect how different created pieces are from each other.
-	\item \textbf{Revise feature representation for actor model.} Probably, procedure of next measure generation should be changed.
+This term must quantify how different created pieces are from each other.
+	\item \textbf{Revise feature representation for policy.} Probably, procedure of next measure generation should be changed.
 Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation.

### 2020-03-09 rlmusician/__main__.py

@@ -16,7 +16,7 @@ import yaml
 
 from rlmusician.agent import (
     CounterpointEnvAgent,
-    create_actor_model, extract_initial_weights, optimize_with_cem
+    create_policy, extract_initial_weights, optimize_with_cem
 )
 from rlmusician.environment import CounterpointEnv, Piece
 
@@ -28,7 +28,7 @@ def evaluate_agent_weights(
         agent_params: Dict[str, Any]
 ) -> float:
     """
-    Evaluate weights of actor model for an agent.
+    Evaluate weights of a policy for an agent.
 
     :param flat_weights:
         1D array of weights to be evaluated
@@ -94,7 +94,7 @@ def main() -> None:
     piece = Piece(**settings['piece'])
     env = CounterpointEnv(piece, **settings['environment'])
     agent_params = {
-        'model_fn': create_actor_model,
+        'policy_fn': create_policy,
         'observation_len': env.observation_space.shape[0],
         'n_lines': len(piece.lines),
         'n_movements_per_line': len(piece.all_movements),
@@ -118,7 +118,7 @@ def main() -> None:
     agent.set_weights(best_weights)
     now = datetime.datetime.now().strftime("%Y-%m-%d_%H:%M:%S,%f")
     weights_path = os.path.join(results_dir, f'agent_weights_{now}.h5f')
-    agent.model.save_weights(weights_path)
+    agent.policy.save_weights(weights_path)
 
     env.verbose = True
     for i_episode in range(cli_args.episodes):

### 2020-03-09 rlmusician/agent/__init__.py

@@ -5,18 +5,18 @@ Author: Nikolay Lysenko
 """
 
 
-from . import actor_model, agent, crossentropy
-from .actor_model import create_actor_model
+from . import agent, crossentropy, policy
 from .agent import CounterpointEnvAgent, extract_initial_weights
 from .crossentropy import optimize_with_cem
+from .policy import create_policy
 
 
 __all__ = [
     'CounterpointEnvAgent',
-    'actor_model',
     'agent',
-    'create_actor_model',
+    'create_policy',
     'crossentropy',
     'extract_initial_weights',
-    'optimize_with_cem'
+    'optimize_with_cem',
+    'policy'
 ]

### 2020-03-09 rlmusician/agent/agent.py

@@ -19,7 +19,7 @@ class CounterpointEnvAgent:
 
     def __init__(
             self,
-            model_fn: Callable[..., 'keras.models.Model'],
+            policy_fn: Callable[..., 'keras.models.Model'],
             observation_len: int,
             n_lines: int,
             n_movements_per_line: int,
@@ -29,8 +29,8 @@ class CounterpointEnvAgent:
         """
         Initialize instance.
 
-        :param model_fn:
-            function that creates actor model
+        :param policy_fn:
+            function that creates shallow neural network policy
         :param observation_len:
             length of observation returned by environment
         :param n_lines:
@@ -38,12 +38,12 @@ class CounterpointEnvAgent:
         :param n_movements_per_line:
             maximum number of line movements supported by environment
         :param hidden_layer_size:
-            size of actor model's hidden layer
+            size of policy's hidden layer
         :param softmax_temperature:
             temperature parameter for Boltzmann softmax which is used for
             mapping scores of candidate actions to their probabilities
         """
-        self.model_fn = model_fn
+        self.policy_fn = policy_fn
         self.observation_len = observation_len
         self.n_lines = n_lines
         self.n_movements_per_line = n_movements_per_line
@@ -51,16 +51,16 @@ class CounterpointEnvAgent:
         self.softmax_temperature = softmax_temperature
 
         self.input_size = observation_len + n_lines * n_movements_per_line
-        self.model = model_fn((self.input_size,), hidden_layer_size)
-        self.shapes = [w.shape for w in self.model.get_weights()]
-        self.sizes = [w.size for w in self.model.get_weights()]
+        self.policy = policy_fn((self.input_size,), hidden_layer_size)
+        self.shapes = [w.shape for w in self.policy.get_weights()]
+        self.sizes = [w.size for w in self.policy.get_weights()]
         self.n_weights = sum(self.sizes)
 
     def represent_actions(
             self, observation: np.ndarray, actions: List[int]
     ) -> np.ndarray:
         """
-        Create batch of potential actions' representation for actor model.
+        Create batch of potential actions' representation for policy.
 
         :param observation:
             observation returned by environment
@@ -87,7 +87,7 @@ class CounterpointEnvAgent:
 
     def set_weights(self, flat_weights: np.ndarray) -> None:
         """
-        Set weights of model.
+        Set weights of policy.
 
         :param flat_weights:
[…]

### 2020-03-09 rlmusician/agent/policy.py

@@ -1,7 +1,7 @@
 """
-Define actor model.
+Define a policy.
 
-Here, actor model maps a pair of an observation and an encoded action to
+Here, a policy maps a pair of an observation and an encoded action to
 score of this action.
 
 Author: Nikolay Lysenko
@@ -39,18 +39,18 @@ def import_keras_silently() -> None:
     sys.stderr = stderr
 
 
-def create_actor_model(
+def create_policy(
         input_shape: Tuple[int], hidden_layer_size: int
 ) -> 'keras.models.Model':
     """
-    Create simple actor model for `CounterpointEnv`.
+    Create simple policy for `CounterpointEnv`.
 
     :param input_shape:
         shape of input without the first dimension (reserved for batch size)
     :param hidden_layer_size:
         number of hidden units
     :return:
-        actor model as a neural network with one hidden layer
+        policy as a neural network with one hidden layer
     """
     # Avoid mandatory importing `keras`-related stuff in the main process.
     import warnings

### 2020-03-09 setup.py

@@ -17,7 +17,7 @@ with open(os.path.join(current_dir, 'README.md'), encoding='utf-8') as f:
 
 setup(
     name='rl-musician',
-    version='0.2.1',
+    version='0.2.2',
     description=description,
     long_description=long_description,
     long_description_content_type='text/markdown',

### 2020-03-09 tests/agent/test_agent.py

@@ -11,7 +11,7 @@ import numpy as np
 import pytest
 
 from rlmusician.agent.agent import CounterpointEnvAgent
-from rlmusician.agent.actor_model import create_actor_model
+from rlmusician.agent.policy import create_policy
 from rlmusician.environment import CounterpointEnv, Piece
 
 
@@ -60,7 +60,7 @@ class TestCounterpointEnvAgent:
     ) -> None:
         """Test `represent_actions` method."""
         agent = CounterpointEnvAgent(
-            create_actor_model,
+            create_policy,
             len(observation),
             n_lines,
             n_movements_per_line,
@@ -95,7 +95,7 @@ class TestCounterpointEnvAgent:
     ) -> None:
         """Test `set_weights` method."""
         agent = CounterpointEnvAgent(
-            create_actor_model,
+            create_policy,
             observation_len,
             n_lines,
             n_movements_per_line,
@@ -103,9 +103,9 @@ class TestCounterpointEnvAgent:
             softmax_temperature=1
         )
         agent.set_weights(flat_weights)
-        model_weights = agent.model.get_weights()
-        flat_model_weights = np.hstack((x.flatten() for x in model_weights))
-        np.testing.assert_equal(flat_model_weights, flat_weights)
+        policy_weights = agent.policy.get_weights()
+        flat_policy_weights = np.hstack((x.flatten() for x in policy_weights))
+        np.testing.assert_equal(flat_policy_weights, flat_weights)
 
     @pytest.mark.parametrize(
         "observation_len, n_lines, n_movements_per_line, hidden_layer_size, "
@@ -186,7 +186,7 @@ class TestCounterpointEnvAgent:
     ) -> None:
         """Test that `run_episode` method does not fail."""
         agent = CounterpointEnvAgent(
-            create_actor_model,
+            create_policy,
             observation_len,
             n_lines,
             n_movements_per_line,

### 2020-03-09 rlmusician/environment/rules.py
chore: Extend a comment with dependencies of a rule

@@ -109,9 +109,9 @@ def check_resolution_of_submediant_and_leading_tone(
         return True
     if line[measure - 1].is_from_tonic_triad:
         return True
-    # NB: This function relies on assumption that every skip must lead to a
-    # pitch from the tonic triad.
-    # Under this assumption, if there are two unstable pitches in a row,
+    # NB: This implementation relies on assumptions that every skip must lead
+    # to a pitch from the tonic triad and only triad pitches may be repeated.
+    # Under these assumptions, if there are two unstable pitches in a row,
     # one of them is submediant and the other is leading tone.
     return movement == previous_movements[-1]
 

### 2020-04-09 rlmusician/__main__.py
refac: Speed up Monte Carlo Beam Search

@@ -11,7 +11,7 @@ from pkg_resources import resource_filename
 
 import yaml
 
-from rlmusician.agent import optimize_with_mcbs
+from rlmusician.agent import optimize_with_monte_carlo_beam_search
 from rlmusician.environment import CounterpointEnv, Piece
 
 
@@ -47,7 +47,9 @@ def main() -> None:
 
     piece = Piece(**settings['piece'])
     env = CounterpointEnv(piece, **settings['environment'])
-    best_action_sequences = optimize_with_mcbs(env, **settings['agent'])
+    best_action_sequences = optimize_with_monte_carlo_beam_search(
+        env, **settings['agent']
+    )
 
     env.verbose = True
     for i_episode, action_sequence in enumerate(best_action_sequences):

### 2020-04-09 rlmusician/agent/__init__.py

@@ -6,7 +6,10 @@ Author: Nikolay Lysenko
 
 
 from . import monte_carlo_beam_search
-from .monte_carlo_beam_search import optimize_with_mcbs
+from .monte_carlo_beam_search import optimize_with_monte_carlo_beam_search
 
 
-__all__ = ['monte_carlo_beam_search.py', 'optimize_with_mcbs']
+__all__ = [
+    'monte_carlo_beam_search',
+    'optimize_with_monte_carlo_beam_search'
+]

### 2020-04-09 rlmusician/agent/monte_carlo_beam_search.py

@@ -5,6 +5,7 @@ Author: Nikolay Lysenko
 """
 
 
+from copy import deepcopy
 from random import choice
 from typing import Any, Dict, List, Optional, Tuple
 
@@ -12,28 +13,38 @@ from rlmusician.environment import CounterpointEnv
 from rlmusician.utils import map_in_parallel
 
 
-def continue_stub(
-        env: CounterpointEnv,
-        stub: List[int]
-) -> Tuple[List[int], float]:
+def roll_in(env: CounterpointEnv, actions: List[int]) -> CounterpointEnv:
     """
-    Continue stub with random actions until episode is finished.
+    Do roll-in actions.
 
     :param env:
         environment
-    :param stub:
-        sequence of actions
+    :param actions:
+        sequence of roll-in actions
+    :return:
+    """
+    env.reset()
+    for action in actions:
+        env.step(action)
+    return env
+
+
+def roll_out_randomly(
+        env: CounterpointEnv,
+        past_actions: List[int]
+) -> Tuple[List[int], float]:
+    """
+    Continue an episode in progress with random actions until it is finished.
+
+    :param env:
+        environment
+    :param past_actions:
+        sequence of actions that have been taken before
     :return:
         finalized sequence of actions and reward for the episode
     """
-    env.reset()
-    done = False
-    past_actions = []
     valid_actions = env.valid_actions
-    for action in stub:
-        observation, reward, done, info = env.step(action)
-        past_actions.append(action)
-        valid_actions = info['next_actions']
+    done = False
     while not done:
         action = choice(valid_actions)
         observation, reward, done, info = env.step(action)
@@ -43,13 +54,13 @@ def continue_stub(
     return record
 
 
-def update_stubs(
+def create_stubs(
         records: List[Tuple[List[int], float]],
         n_stubs: int,
         stub_length: int
 ) -> List[List[int]]:
     """
-    Create new stubs based on collected statistics.
+    Create roll-in sequences (stubs) based on collected statistics.
 
     :param records:
         sorted statistics of played episodes as sequences of actions
@@ -72,7 +83,7 @@ def update_stubs(
     return stubs
 
 
-def optimize_with_mcbs(
+def optimize_with_monte_carlo_beam_search(
         env: CounterpointEnv,
         beam_width: int,
         n_records_to_keep: int,
@@ -80,7 +91,7 @@ def optimize_with_mcbs(
[…]

### 2020-04-09 tests/agent/test_monte_carlo_beam_search.py

@@ -10,10 +10,42 @@ from typing import Any, Dict, List, Tuple
 import pytest
 
 from rlmusician.agent.monte_carlo_beam_search import (
-    optimize_with_mcbs,
-    update_stubs
+    create_stubs,
+    optimize_with_monte_carlo_beam_search,
 )
 from rlmusician.environment import CounterpointEnv, Piece
+
+
+@pytest.mark.parametrize(
+    "records, n_stubs, stub_length, expected",
+    [
+        (
+            # `records`
+            [
+                ([1, 2, 3], 5),
+                ([1, 2, 2], 4),
+                ([3, 2, 1], 3),
+                ([2, 3, 2], 2),
+            ],
+            # `n_stubs`
+            2,
+            # `stub_length`
+            2,
+            # `expected`
+            [
+                [1, 2],
+                [3, 2],
+            ]
+        ),
+    ]
+)
+def test_create_stubs(
+        records: List[Tuple[List[int], float]], n_stubs: int, stub_length: int,
+        expected: List[List[int]]
+) -> None:
+    """Test `create_stubs` function."""
+    result = create_stubs(records, n_stubs, stub_length)
+    assert result == expected
 
 
 @pytest.mark.parametrize(
@@ -75,48 +107,16 @@ from rlmusician.environment import CounterpointEnv, Piece
         ),
     ]
 )
-def test_optimize_with_mcbs(
+def test_optimize_with_monte_carlo_beam_search(
         env: CounterpointEnv,
         beam_width: int,
         n_records_to_keep: int,
         n_trials_schedule: List[int],
         paralleling_params: Dict[str, Any]
 ) -> None:
-    """Test that `optimize_with_mcbs` runs without failures."""
-    results = optimize_with_mcbs(
+    """Test that `optimize_with_monte_carlo_beam_search` has no failures."""
+    results = optimize_with_monte_carlo_beam_search(
         env, beam_width, n_records_to_keep, n_trials_schedule,
         paralleling_params
     )
     assert len(results[0]) == env.piece.n_measures - 2
-
-
-@pytest.mark.parametrize(
-    "records, n_stubs, stub_length, expected",
-    [
-        (
-            # `records`
-            [
-                ([1, 2, 3], 5),
-                ([1, 2, 2], 4),
-                ([3, 2, 1], 3),
-                ([2, 3, 2], 2),
-            ],
-            # `n_stubs`
-            2,
-            # `stub_length`
-            2,
-            # `expected`
[…]

### 2020-04-11 docs/paper/paper.tex
docs: Update beginning of the paper

@@
 
-\title{RL-Musician: A Tool for Music Composition with Reinforcement Learning}
+\title{RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition}
 
-\date{January 26, 2020}
+\date{April 4, 2020}
 
@@
 \begin{abstract}
-In this paper, automatic composition of first species counterpoint is framed as a black-box optimization problem close to reinforcement learning.
-An agent with shallow neural network as a policy composes polyphonic pieces sequentially one measure (bar) per time step.
-Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory.
-Cross-entropy method is used for searching policy weights that result in maximum expected score of a piece.
+In this paper, automatic composition of first species counterpoint is framed as a tree search problem close to reinforcement learning.
+An environment representing a polyphonic piece in progress is introduced.
+An action is filling of the current measure (bar) for all melodic lines simultaneously.
+Once a specified number of measures is added, an episode terminates and the resulting piece is evaluated based on some hand-written rules derived from music theory.
+Best sequences of actions are found with a variation of Monte Carlo Beam Search.
 
@@
 Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}.
+Meanwhile, fitness function optimization also yields prominent results\cite{herremans2012first, herremans2013fifth}.
 
@@
 The approach relates to a well-known framework for simplified music composition called species counterpoint.
-It is often used by beginning composers in order to practice in voice leading (also known as part writing) and polyphony.
+It is usually used by beginning composers in order to practice in voice leading (also known as part writing) and polyphony.
 From automatic composition point of view, species counterpoint is interesting due to two its properties:
@@
 Each measure must be either empty (if it is not filled yet) or containing exactly one note per a melodic line.
-Initially, only the first measure and the last measure are filled with user-defined values.
-At a particular time step of an episode, an agent observes vector in which for every pitch there is an exponential moving average over all measures up to the current one (exclusively) of binary indicators whether this pitch is played at corresponding measure.
-An action of the agent is filling of the current measure, i.e., adding a pitch to each of the melodic lines.
-When the last measure is reached, episode is ended and the agent receives reward based on evaluation of created composition.
-Trainable parameters of the agent are weights of a neural network used as a policy.
-These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with evolutionary strategies \cite{salimans2017evolution}, genetic algorithms, simulated annealing, or other black-box optimization methods).
-From optimization perspective, average over several episodes reward is estimated value of fitness function at the weights used.
+Initially, only the first measure and the last measure are filled with user-defined values and the current measure is the second one.
+An action is filling of the current measure, i.e., adding a pitch to each of the melodic lines.
+After an action, the next measure becomes the current one.
+When the last measure is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.
 
-Obviously, the approach belongs to reinforcement learning, because there is an environment and an agent, but a dataset is absent.
-Nevertheless, there are some connections with supervised deep learning.
-From generation of new pieces point of view, a policy returns probabilities of next sonorities (chords) given current state.
-This is similar to sampling new sequences from a next-step prediction model trained with maximum likelihood method on a dataset of existing sequences.
-
-More details on the methodology are provided in Section \ref{sec:methodology} but here it is appropriate to discuss its advantages.
-Of course, automatic composition of species counterpoint is not novel -- some researchers have developed optimization-based approaches not only for first species, but also for fifth species \cite{herremans2012composing}.
-Nevertheless, as far as the author knows, such approaches yield a single piece as a result of optimization, whereas in the current approach training produces an agent that creates a new piece after each run.
-
-Next-step prediction models create a new piece after each run too, but they are trained supervisedly. The reasons for not involving supervised training at all are as follows:
-\begin{itemize}
-	\item Finding new ways of music creation is a more challenging task than imitation of famous pieces.
-If no known pieces are used, chances are that the harder problem is considered and it is not replaced with the simpler problem of imitation.
-	\item There are tuning systems other than equal temperament (for instance, in microtonal music).
-For some of them it may be impossible to collect dataset large enough to allow training models in a supervised fashion.
-However, developers of a tuning system should know some underlying principles and so (at least, in theory) it is possible to create evaluational rules and train an agent based on them.
-\end{itemize}
-
-Actually, results reported at this draft version are far from using above advantages at full scale (in particular, first species counterpoint is too limiting).
-The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:conclusion}.
-% Chances are that a way of reinforcement learning combines strengths of fitness function optimization with strengths of supervised learning.
+The remainder of the paper is organized as follows.
+Section \ref{sec:literature} describes the most relevant works and the place of the current paper amongst them.
+Namely, Section \ref{subsec:composition} discusses algorithmic composition, whereas Section \ref{subsec:montecarlo} contains preliminaries on Monte Carlo Beam Search and lists modifications made in this study.
+Details of the setup (in particular, rules used for evaluation of pieces) are defined in Section \ref{sec:methodology}.
+Section \ref{sec:results} reports experimental results and the way to reproduce them.
+Finally, Section \ref{sec:conclusion} concludes.
 
@@
 An interested reader can find more information in specialized reviews such as \cite{fernandez2013ai}.
-What is discussed below is just some examples related to reinforcement learning.
+What is discussed below is just some examples related to the current study.
 
-Reinforcement learning can be applied to algorithmic composition either as a standalone paradigm or in connection with deep learning.
-As for the first case, probably, there are no influential works and only separate studies like \cite{yi2007automatic} are available.
+These examples can be divided into two groups.
+The first one is formed by approaches where given fitness function is optimized.
+The second group is formed by composition of music with reinforcement learning in its narrow sense (i.e., methods based on value function estimation).
+This group is something in between the first group and supervised learning approaches.
+On the one hand, reward function is a domain-specific name for fitness function.
+On the other hand, data generated during past episodes are used for training supervised models.
+
+The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group.
+A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function.
+What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint.
+However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes.
+Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations.
+As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces can not be created.
+This way of that, the core drawback of the discussed papers is that download links to software implementation named Optimuse are broken and so the results can not be reproduced.
+
+There are other optimization-based approaches as well.
+Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method) or construction metaheuristics (ant colony algorithms) can be used.
+However, such papers are less relevant to the current study and so let us switch our attention to reinforcement learning.
+It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning.
+As for the first case, probably, there are no influential works and only separate studies like \cite{smith2012reinforcement, yi2007automatic} are available.
 Conversely, the second group recently gained attention and several interesting approaches were developed there.
 
-For example, reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}.
+For example, reinforcement learning can be used for altering weights of recurrent neural networks (RNNs) trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}.
 The goal is to make generated pieces more structured and conformed with music theory rules.
 To define an environment, let its state be composed of recurrent neural networks states and previously played notes, let an action be an output for current time step and let reward depend on both evaluational rules and probability of output according to initial RNN.
-Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred as a training algorithm.
+Rewards are granted immediately after a step and DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred as a training algorithm.
 
@@
 
-\subsection{Cross-Entropy Method}
-\label{subsec:crossentropy}
+Let us highlight again that above cases of reinforcement learning usage still require supervised learning and datasets. The arguments for not involving them at all are as follows:
+\begin{itemize}
+	\item Finding new ways of music creation is a more challenging task than imitation of famous pieces.
+If no known pieces are used, chances are that the harder problem is considered and it is not replaced with the simpler problem of imitation.
+	\item There are tuning systems other than equal temperament (for instance, in microtonal music).
+For some of them it may be impossible to collect a dataset large enough to allow training models in a supervised fashion.
+However, developers of a tuning system should know some underlying principles and so (at least, in theory) it is possible to define a fitness function and find pieces that optimize it or to define a reward and train an agent based on it.
+\end{itemize}
+This implies that optimization-based approaches and pure reinforcement learning approaches must not be overlooked due to advances in supervised learning application.
 
-Initially, cross-entropy method was developed for estimation of rare events probability \cite{rubinstein1997optimization}.
-However, it was found that it is also appropriate for solving optimization problems.
-More detailed discussion of cross-entropy method can be found in \cite{boer2005tutorial}.
+\subsection{Monte Carlo Beam Search}
+\label{subsec:montecarlo}
 
-\begin{algorithm}
-	\caption{Cross-entropy method for optimization} \label{alg:crossentropy}
-	\textbf{Input:} $X$ -- set of elements, $f: X \to \mathbb{R}$ -- target function, $u(\cdot, w)$ -- probabilistic distribution over $X$ parametrized by vector $w$. \\
-	\textbf{Output:} $\hat{w}$ -- approximate solution to the problem $\max_w \mathbb{E}_{x \sim u(\cdot, w)} f(x)$. \\
-	\textbf{Hyperparameters:} $w^{(0)}$ -- initial value of $w$; $N$ -- number of iterations, $n$ -- number of vectors to draw at each iteration, $\sigma$ -- standard deviation for vectors generation, $m$ -- number of trials for each vector; $\rho$ -- fraction of best vectors to use for update; $\alpha$ -- smoothing coefficient of updates.
-	\begin{algorithmic}[1]
-		\FORALL{$i \in \{1, \dots, N\}$}
-		    \FORALL{$j \in \{1, \dots, n\}$}
-		        \STATE{draw $w^{(i,j)} \sim \mathcal{N}(\cdot \vert w^{(i-1)}, \sigma)$}
-		        \STATE{$r_j \gets \sum_{k = 1}^{m} f(x_k)$ where $x_k \sim u(\cdot, w^{(i,j)})$}
-		    \ENDFOR
-		    \STATE{$r_{\mathrm{threshold}} \gets$ $[\rho n]$-th highest value of $\{r_j: j \in \{1, \dots, n\}\}$}
-		    \STATE{$J \gets \{j: r_j \ge r_{\mathrm{threshold}}\}$}
-		    \STATE{$w^{(i)} \gets \alpha w^{(i-1)} + (1 - \alpha)(\sum_{j \in J} w^{(i,j)}) / [\rho n]$}
-		\ENDFOR
-	    \STATE{$\hat{w} \gets w^{(N)}$}
-	\end{algorithmic}
-\end{algorithm}
-
-Algorithm \ref{alg:crossentropy} defines a variant of cross-entropy method for optimization.
-Sometimes, hyperparameter $m$ is omitted and intermediate results are not aggregated over multiple trials.
-In case of $u(\cdot, w)$ that acts like a deterministic function of $w$, $m$ is redundant, but, in general case, terminal result can be improved by setting $m > 1$.
-
-In context of reinforcement learning, cross-entropy method is used for searching parameters of agent that result in maximum expected reward.
-So $X$ is a set of all possible finished episodes, $f$ is reward, and $u(\cdot, w)$ describes how an episode depends on parameters.
-Cross-entropy method is best suitable for environments where reward becomes known only after an episode is finished.
-If there are rewards granted after intermediate steps, methods based on Q-values usually outperform it.
+Consider a tree search problem.
 
@@
 Suppose also that a diatonic scale (like C-major or A-minor) is chosen.
-Let each line be a list of pitches from the scale (i.e., only 7 notes per octave can be used and chromaticism is prohibited).
+Let each line be a list of pitches from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited).
 Set length of all lines to the same number $m$, where $m$ is usually between 8 and 24.
@@
 
-The implementation relies on some open-source tools \cite{brockman2016openai,chollet2015keras,oliphant2006guide,raffel2014intuitive}.
+The implementation relies on some open-source tools \cite{brockman2016openai,oliphant2006guide,raffel2014intuitive}.
 

### 2020-04-11 docs/paper/paper.tex
docs: Describe Monte Carlo Beam Search in the draft

@@
 Thus, there is an open research problem.
-Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{smith2012reinforcement} are applicable.
+Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{yi2007automatic, smith2012reinforcement} are applicable.
 Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}.
@@
 As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces can not be created.
-This way of that, the core drawback of the discussed papers is that download links to software implementation named Optimuse are broken and so the results can not be reproduced.
+This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results can not be reproduced.
 
@@
 It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning.
-As for the first case, probably, there are no influential works and only separate studies like \cite{smith2012reinforcement, yi2007automatic} are available.
+As for the first case, probably, there are no influential works and only separate studies like \cite{yi2007automatic, smith2012reinforcement} are available.
 Conversely, the second group recently gained attention and several interesting approaches were developed there.
@@
 Consider a tree search problem.
+To be more detailed, suppose that there is a tree such that every its leaf is mapped to a reward and the goal is to find the leaf with the maximum reward.
+However, leaves and corresponding to them rewards are not known initially.
+Information about a leaf becomes available only after a path from the root of the tree to this leaf is constructed.
+%In other words, the goal is to find optimum sequence of nodes starting at the root and ending at a leaf.
+
+The described in Section \ref{sec:introduction} procedure of first species counterpoint writing can be considered a tree search problem.
+The root is the initial state of an environment.
+Each valid option to fill the current measure is an edge from the current node of the tree to a new node.
+Leaves are pieces where all measures are filled.
+
+If size of a tree is small enough, exhaustive search is possible.
+There are well-known classical algorithms such as depth-first search (DFS) that allow iterating over all paths from the root to leaves.
+Unfortunately, trees are large in numerous real problems.
+For instance, a rough estimate of total number of leaves for first species counterpoint with 2 voices and 16 measures is about $7^{14}$ which is equal to several hundreds of billions.
+
+An obvious idea for a large tree is to use random search.
+To create a path, every next node is chosen at random until a leaf is reached.
+After specified number of paths is created, the best one amongst them is chosen as an approximate solution.
+
+This idea can be further improved in the following way.
+Let an optimum path be constructed sequentially with adding one new node per a series of random trials.
+If so, there is a stub initially containing only the root.
+At every iteration, there are several trials to continue stub at random until a leaf is reached.
+The best one amongst them (and amongst trials from previous iterations if applicable) is chosen and the first non-stub node from there is added to the stub.
+This method can be called Monte Carlo Search.
+The gain in comparison with plain random search is due to allocating more exploration to subtrees that contain the current best leaf.
+
+Above version of Monte Carlo Search is greedy, because it ignores that some subtrees might be underexplored.
+A well-known method that explores more than greedy search, but uses less computational resources than exhaustive search, is called beam search.
+Instead of just one stub, $w$ distinct stubs are constructed and used, where $w$ is a hyperparameter named beam width.
+It is natural to call this method Monte Carlo Beam Search.
+
+Probably, the term Monte Carlo Beam Search was coined in \cite{cazenave2012beam}.
+However, it has different meaning there.
+In \cite{cazenave2009nested}, the same author defined Nested Monte Carlo Search as a combination of Monte Carlo Search with DFS.
+Given a current stub, all its extensions by $n$ new nodes (where $n$ is a hyperparameter named level of nestedness) are found with DFS and then continued at random.
+So pre-defined number of trials per iteration is absent, because number of trials is equal to number of extensions by $n$ nodes ahead.
+Comparing with Monte Carlo Search, Nested Monte Carlo Search distributes exploration more evenly between short extensions of a stub.
+This way or that, in \cite{cazenave2012beam}, Monte Carlo Beam Search is defined as beam search based on Nested Monte Carlo Search.
+
+Nevertheless, here, beam search is based on Monte Carlo Search.
+The reason for it is quite mundane -- parallel implementation of DFS looks like overcomplication that may yield too few.
+The second difference with \cite{cazenave2012beam} is that dependence of trials number on stub length is introduced.
+The longer the stub is, the smaller a subtree to be explored is.
+Therefore, it is better to reallocate computational resources in order to explore more when a stub is short and to explore less when a stub is long.
 
@@
 A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}.
-The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.2.0/}}.
+The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.3.0/}}.
 All important settings are placed to a separate configuration file and so it is easy to experiment with them.

### 2020-04-12 docs/paper/paper.tex
docs: Update experimental results

@@
 There are other optimization-based approaches as well.
-Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method) or construction metaheuristics (ant colony algorithms) can be used.
+Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method), or construction metaheuristics (ant colony algorithms) can be used.
 However, such papers are less relevant to the current study and so let us switch our attention to reinforcement learning.
@@
 To create a path, every next node is chosen at random until a leaf is reached.
-After specified number of paths is created, the best one amongst them is chosen as an approximate solution.
+After specified number of paths are created, the best one amongst them is chosen as an approximate solution.
 
@@
 \end{itemize}
-In addition, there are two rules of harmony:
+In addition, there are three rules of harmony:
 \begin{itemize}
 	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
+	\item Within a measure, all pitch classes must be different unless the measure is the first one or the last one;
 	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
@@
 
-Above rules are absolute and an agent cannot break them, because it selects pitches for the next measure only from options that are filtered according to them.
-If a set of such options is empty, episode terminates and an agent receives negative reward for dead end.
-
-If the set of options is not empty, an agent needs an observation.
-Observation is provided in terms of so called piano roll format.
-Denote size of pitch range available to the agent as $n$ and let the piece be represented also as a matrix $R$ of size $n \times m$ where $R_{ij}$ is 1 if $i$-th pitch is played in any line in $j$-th measure and 0 otherwise.
-Denote current measure as $t$, $1 \le t \le m$.
-Then observation is defined as vector $v$ of length $n$:
-$$v = \sum_{i = 1}^t \beta^{t - i} R_{\cdot i},$$
-where $\beta$ is a hyperparameter related to exponential decay, $0 < \beta < 1$.
-
-For each option, agent concatenates observation $v$ with encoding this option vector $u$ of length $(2s + 1)l$.
-Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them.
-Concatenation of $v$ and $u$ is passed as input to a policy which is a shallow neural network returning a single real-valued score.
-After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions.
-An action is drawn from this distribution and the agent takes it.
-
-Remember that an episode starts with the first and last measures already filled.
-The agent fills empty measures one-by-one according to the described above procedure.
-When the final measure is reached, the episode ends, created piece is evaluated and the agent receives reward.
+Above rules are absolute and it is not possible to break any of them, because pitches for the next measure can be selected only from options that are filtered according to these rules.
+If a set of such options is empty, episode terminates and an unfinished sequence of actions receives negative reward for dead end.
 
@@
 
-At the current revision of the study, eight properties of a final piece are evaluated separately and total reward is a weighted sum of these eight scores.
+At the current revision of the study, seven properties of a final piece are evaluated separately and total reward is a weighted sum of these seven scores.
 
 \begin{itemize}
-	\item \textbf{Autocorrelation.} Pieces where the same sequence of sounds is repeated over and over again are worse than pieces with no repetitions. To reflect this in reward, this score is used:
-	$$1 - \max_k \{|\mathrm{Autocorr}_k(R)| : 2 \le k \le 8\},$$
-	where $R$ is piano roll representation of the piece, $\mathrm{Autocorr}_k$ is autocorrelation function for lag $k$ (for matrix $R$ its autocorrelation is average autocorrelation of its rows) and maximum lag to consider is set to 8 based on common sense.
-	\item \textbf{Entropy of pitch distribution.} To encourage usage of all pitches from a range corresponding to a line, there is a score equal to average over all lines actual entropy of pitch distribution within the line normalized by entropy of uniform distribution on the same set of pitches.
 	\item \textbf{Absence of looped pitches.} A stable pitch may be rearticulated, but not more than once in a row.
 Every excessive occurrence of the same pitch within a line adds constant negative reward.
-	\item \textbf{Absence of pitch or pitch class clashes.} According to species counterpoint methodology, lines must be as distinguishable from each other as possible.
-In particular, this implies that the same pitches or the same pitch classes should not sound simultaneously in different lines.
-What is more, unison or octave harmonic intervals are too stable and so they create illusion of false finish if they occur in intermediate measures.
-To take into account these considerations, there is a negative reward proportional to share of intervals with both pitches belonging to the same pitch class amongst all intervals between two simultaneously sounding pitches.
-Also there is extra negative reward proportional to share of unison intervals.
+    \item \textbf{Absence of looped fragments.} Define a fragment as one or more consecutive measures (at a piece level, not at a melodic line level).
+For every fragment followed by exactly the same fragment, a constant negative reward is added.
 	\item \textbf{Motion.} For every pair of lines every motion between two successive measures adds a reward that depends on a type of motion.
@@
 	\item \textbf{Correlation between lines.} Consider each line as a sequence of real numbers equal to pitch positions within a range of available pitches.
-Then half of difference of 1 and average over all pairs of lines correlation between two lines is returned as a reward for independence of lines.
+Then average over all pairs of lines correlation between two lines can quantify independence of lines.
+Actually, reward is rescaled to be from 0 to 1 -- it is half of difference of 1 and the above average correlation.
     \item \textbf{Explicity of climax.} Music theory suggests that good melodies must have goal-oriented motion and that such goal-orientedness may be achieved by having exactly one climax point.
-For each line, negative rewards are granted for each duplication of climax point and for not so high climax point.
+For each line, constant reward is decreased for each duplication of climax point and for not so high climax point.
     \item \textbf{Number of skips.} Positive reward is associated with every line such that number of skips in it lies within a specified range.
 The motivation behind this is that lines with almost no skips are not interesting, but lines with too many skips are not coherent.
+    \item \textbf{Absence of downward skips}.
+Counterpoint theory prescribes to have downward step motion.
+To encourage its usage, there are negative rewards for every downward skip and the large a skip is, the higher a penalty its.
 \end{itemize}
 
-It can be seen that the first three properties deal with non-triviality of results, the next three properties deal with interaction of lines and the last two properties deal with melodic quality of each separate line.
+It can be seen that the first two properties deal with non-triviality of results, the next two properties deal with interaction of lines and the last three properties deal with melodic quality of each separate line.
 
@@
 In this section, results are reported for experiments with 2 melodic lines and 16 measures starting from (G4, C5) and ending with (C4, C5).
-It is found that scores for autocorrelation and entropy of pitch distribution are redundant.
-Non-degenerate pieces are produced even without them.
-So only six properties are evaluated.
-They are weighted so that explicity of climax and absence of pitch class clashes have 3 times more impact than any other property.
-
-In Table \ref{table:scores}, scores of a typical piece created by an agent trained for 20 populations, are compared with two benchmarks.
-The first one is simple baseline -- it is score of a piece created by an agent with weights that are the best amongst 100 random weights.
-The second one are scores for a piece manually created by the author in about 15 minutes.
+Properties to be evaluated are weighted so that explicity of climax has 3 times more impact than any other property.
 
@@
 			\hline
-			& Baseline & Agent & Human & Maximum\\
+			& Human & MCS & MCBS & Maximum\\
 			\hline
-			Absence of looped pitches & -1 & 0 & 0 & 0\\
+			Absence of looped pitches & 0 & 0 & 0 & 0\\
 			\hline
-			Absence of pitch or pitch class clashes & -0.429 & -1.286 & -0.643 & 0\\
+			Absence of looped fragments & -1 & 0 & 0 & 0\\
 			\hline
-			Motion & -0.1 & -0.033 & 0.1 & 1\\
+			Motion & -0.267 & -0.1 & -0.033 & 1\\
 			\hline
-			Correlation between lines & 0.552 & 0.168 & 0.278 & 1\\
+			Correlation between lines & 0.269 & 0.321 & 0.321 & 1\\
 			\hline
-			Explicity of climax & 0.9 & 3 & 3 & 3\\
+			Explicity of climax & 3 & 3 & 3 & 3\\
 			\hline
-			Number of skips & 0 & 1 & 1 & 1\\
+			Number of skips & 1 & 1 & 1 & 1\\
 			\hline
-			Total reward & -0.076 & 2.849 & 3.735 & 6\\
+			Absence of downward skips & 0 & -0.6 & 0 & 0\\
+			\hline
+			Total reward & 3.003 & 3.621 & 4.288 & 6\\
 			\hline
@@
 
-It can be seen that the agent's piece consists of two lines with formally good melodies but these lines are not independent enough.
-Table \ref{table:piece} contains the created by the agent piece for which above scores are reported.
-Visual analysis of the shown piano roll confirms observation about lines dependence -- their climax points coincide and there are plenty of intermediate octave intervals.
+In Table \ref{table:scores}, scores of a piece found with Monte Carlo Beam Search are compared with benchmarks.
+The piece was found with beam width set to 5 and number of trials schedule set to ($10000 \times 3$, $5000 \times 3$, $1000 \times 8$) -- this means that the second measure of the piece was chosen after 10000 random continuations and, for example, the eighth measure of the piece was chosen after 1000 random continuations.
+One of the benchmarks is achieved by Monte Carlo Search with 43000 random trials (i.e., total number of random trials is the same).
+The other benchmark relates to a piece manually created by the author in about 10 minutes.
+Ideally, scores should be collected during multiple independent experiments and then significance of difference should be tested, but it is not done yet.
+
+The results look like scores for climax explicity and absence of looped pitches, absence of looped fragments, and absence of downward skips impose hard constraints.
+If so, the goal is to maximize sum of scores for lines independence subject to counterpoint rules and above new constraints.
+
+Table \ref{table:piece} contains found by Monte Carlo Beam Search piece for which the scores are reported.
 
 \begin{table}[h!]
-	\caption{A piece composed by an agent}
+	\caption{The piece found with Monte Carlo Beam Search}
 	\label{table:piece}
@@
 			\hline
-			& & & & & & & C6 & & & & & & & & \\
+			& & & C6 & & & & & & & & & & & & \\
 			\hline
-			& & & & & & B5 & & & & & & & & & \\
+			& & & & B5 & & & & & & & & & & & \\
 			\hline
@@
 			\hline
-			& & & & & A5 & & & & A5 & & & & & & \\
+			& & & & & A5 & & & &  & & & & & & \\
 			\hline
@@
 			\hline
-			& & & & G5 & & & & G5 & & & & & & & \\
+			& & G5 & & & & G5 & & & & & & & & & \\
 			\hline
@@
 			\hline
-			& & & F5 & & & & & & & & F5 & & & & \\
+			& & & & & & & F5 & & & & & & & & \\
 			\hline
-			& & E5 & & & & & & & & E5 & & & & & \\
+			& E5 & & & & & & & E5 & & & & & & & \\
 			\hline
@@
 			\hline
-			& D5 & & & & & & & & & & & & D5 & & \\
+			& & & & & & & & & D5 & & & & & & \\
 			\hline
@@
 			\hline
-			C5 & & & & & & & C5 & & & & & C5 & & C5 & C5 \\
+			C5 & C5 & & & & & & & & & C5 & & C5 & C5 & & C5 \\
 			\hline
-			& & & & & & & & B4 & & & & & & & \\
+			& & B4 & & & & & & & & & B4 & & & B4 & \\
 			\hline
@@
 			\hline
-			& & & & & & & & & A4 & & & & & & \\
+			& & & A4 & & & & & & & & & & & & \\
 			\hline
@@
 			\hline
-			G4 & & & & G4 & & G4 & & & & G4 & & & & & \\
+			G4 & & & & G4 & & & & G4 & & G4 & G4 & & & & \\
 			\hline
@@
 			\hline
-			& F4 & & & & F4 & & & & & & F4 & & & & \\
+			& & & & & F4 & & & & F4 & & & F4 & & & \\
 			\hline
-			& & & & & & & & & & & & E4 & & & \\
+			& & & & & & E4 & & & & & & & E4 & & \\
 			\hline
@@
 			\hline
-			& & & D4 & & & & & & & & & & D4 & & \\
+			& & & & & & & D4 & & & & & & & D4 & \\
 			\hline
@@
 			\hline
-			& & C4 & & & & & & & & & & & & C4 & C4 \\
+			& & & & & & & & & & & & & & & C4 \\
 			\hline
@@
 
-As for subjective evaluation of music quality, it is not correct to compare the selected piece created by the agent with the piece created by the author.
-The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically.
-That being said, subjective perception of a resulting piece is not aligned with fitness function used during optimization.
-However, it must be noted that the piece shown above looks like excessively oriented on maximization of fitness function regardless more subtle properties important for human listeners.
-It is not a problem associated with training, it is a flaw in problem setup (more criteria for evaluation are needed).
-
-Numerous experiments revealed another important issue.
-Actually, different runs of a trained agent produce variations of the same piece.
-Although there might be significant differences in scores (in particular, because some scoring functions are discrete), there are no significant differences from music point of view.
+Subjective evaluation of music quality can be used for revising setup of the problem. Probably, 16 measures is too much, because long step motions are resembling descending scales.
 
@@
 
-This paper studies capabilities of reinforcement learning for music composition without usage of existing pieces.
-Here, music composition is framed as optimization problem where the goal is to find values of generative model's parameters maximizing expected score of pieces generated with them.
-It is found that some progress can be made in this direction.
+This paper introduces a new approach for music composition without usage of existing pieces.
 
@@
 \begin{itemize}
-	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged.
-Also downward skips can be penalized in order to have upward skips before climax and downward conjunct motion after it.
-	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes.
-However, a new term can be added to this average.
-This term must quantify how different created pieces are from each other.
-	\item \textbf{Revise feature representation for policy.} Probably, procedure of next measure generation should be changed.
-Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation.
-To make it less raw, some features inspired by music theory may be engineered.
-	\item \textbf{Tune hyperparameters.} In the current software implementation, cross-entropy method has five hyperparameters, an agent has two hyperparameters, and evaluation functions have more than ten hyperparameters.
-As of now, not so many computational experiments are held in order to find their optimum values.
-	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches.
-Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates.
-Moments of time when level of tension changes may be rhythmically accented or unaccented.
-Close but not exact alignment between rhythmical accent and resolution often results in aesthetic and interesting music.
-Taking into account constraints of smooth voice leading, optimization problem can be set up based on above considerations.
+	\item Use 
 \end{itemize}

### 2020-04-14 docs/paper/paper.tex
docs: Update conclusion of a paper draft

@@
 
-\date{April 4, 2020}
+\date{April 14, 2020}
 
@@
 Currently, there are no perfect tools for algorithmic composition.
-Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer.
+Plenty of various approaches to composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer.
 Thus, there is an open research problem.
@@
 These examples can be divided into two groups.
-The first one is formed by approaches where given fitness function is optimized.
+The first one is formed by approaches where an explicitly given fitness function is optimized.
 The second group is formed by composition of music with reinforcement learning in its narrow sense (i.e., methods based on value function estimation).
@@
 Information about a leaf becomes available only after a path from the root of the tree to this leaf is constructed.
-%In other words, the goal is to find optimum sequence of nodes starting at the root and ending at a leaf.
 
@@
 Counterpoint theory prescribes to have downward step motion.
-To encourage its usage, there are negative rewards for every downward skip and the large a skip is, the higher a penalty its.
+To encourage its usage, there are negative rewards for every downward skip and the larger a skip is, the higher a penalty its.
 \end{itemize}
@@
 
-The results look like scores for climax explicity and absence of looped pitches, absence of looped fragments, and absence of downward skips impose hard constraints.
+The results look like scores for climax explicity, absence of looped pitches, absence of looped fragments, and absence of downward skips impose hard constraints.
 If so, the goal is to maximize sum of scores for lines independence subject to counterpoint rules and above new constraints.
 
-Table \ref{table:piece} contains found by Monte Carlo Beam Search piece for which the scores are reported.
+Table \ref{table:piece} demonstrates found by Monte Carlo Beam Search piece for which the scores are reported.
 
@@
 
-This paper introduces a new approach for music composition without usage of existing pieces.
+This paper introduces a new approach to composition of music without usage of existing pieces.
+For a problem of first species counterpoint writing, the approach finds pieces that maximize weighted sum of associated with particular musical properties scores subject to boundary constraints.
+These boundary constraints (the first measure and the last measure) are user-defined and weights for the sum of scores are user-defined too.
 
-The list of suggested further improvements is as follows:
-\begin{itemize}
-	\item Use 
-\end{itemize}
+Results are good from formal metrics point of view -- the method outperforms random search and manual trials.
+However, created pieces are too simple to be enjoyable music.
+This is an inevitable consequence of too strict limitations of first species counterpoint.
+To eliminate them, higher species can be chosen or frameworks other than species counterpoint can be explored.
+
+If the problem becomes computationally complex after introducing more musical freedom, concepts from reinforcement learning such as value function can be involved.
+State-free search can be replaced with a search where decision are made based on observations and estimated values of states and actions.
 

### 2020-04-15 docs/paper/paper.tex
docs: Fix some typos

@@
 Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations.
-As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces can not be created.
-This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results can not be reproduced.
+As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces cannot be created.
+This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results cannot be reproduced.
 
@@
 Above version of Monte Carlo Search is greedy, because it ignores that some subtrees might be underexplored.
-A well-known method that explores more than greedy search, but uses less computational resources than exhaustive search, is called beam search.
+A well-known method that explores more than greedy search but uses less computational resources than exhaustive search, is called beam search.
 Instead of just one stub, $w$ distinct stubs are constructed and used, where $w$ is a hyperparameter named beam width.
@@
 	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
-	\item Movement in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
+	\item Movement by step in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
 	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
@@
 The piece was found with beam width set to 5 and number of trials schedule set to ($10000 \times 3$, $5000 \times 3$, $1000 \times 8$) -- this means that the second measure of the piece was chosen after 10000 random continuations and, for example, the eighth measure of the piece was chosen after 1000 random continuations.
-One of the benchmarks is achieved by Monte Carlo Search with 43000 random trials (i.e., total number of random trials is the same).
+One of the benchmarks is achieved by Monte Carlo Search with 53000 random trials (i.e., total number of random trials is the same).
 The other benchmark relates to a piece manually created by the author in about 10 minutes.
@@
 
-Subjective evaluation of music quality can be used for revising setup of the problem. Probably, 16 measures is too much, because long step motions are resembling descending scales.
+Subjective evaluation of music quality can be used for revising setup of the problem. Probably, 16 measures are too much, because long step motions are resembling descending scales.
 

### 2020-04-16 docs/paper/paper.tex

@@
 The other benchmark relates to a piece manually created by the author in about 10 minutes.
-Ideally, scores should be collected during multiple independent experiments and then significance of difference should be tested, but it is not done yet.
+Ideally, scores should be collected during multiple independent experiments and then statistical significance of difference should be tested, but it is not done yet.
 

### 2020-04-18 rlmusician/utils/__init__.py
feat: Revise work with scales

@@ -11,15 +11,15 @@ from .io import (
     create_wav_from_events
 )
 from .misc import convert_to_base, map_in_parallel
-from .music_theory import create_scale, slice_scale
+from .music_theory import Scale, ScaleElement
 
 
 __all__ = [
+    'Scale',
+    'ScaleElement',
     'convert_to_base',
     'create_events_from_piece',
     'create_midi_from_piece',
-    'create_scale',
     'create_wav_from_events',
     'map_in_parallel',
-    'slice_scale'
 ]

### 2020-04-18 rlmusician/utils/music_theory.py

@@ -11,61 +11,101 @@ from sinethesizer.io.utils import get_note_to_position_mapping
 
 
 NOTE_TO_POSITION = get_note_to_position_mapping()
+TONIC_TRIAD_DEGREES = (1, 3, 5)
 
 
 class ScaleElement(NamedTuple):
-    """A pitch from a scale."""
+    """A pitch from a diatonic scale."""
 
-    absolute_position: int
+    note: str
+    position_in_semitones: int
+    position_in_degrees: int
     degree: int
+    is_from_tonic_triad: bool
 
 
-def create_scale(tonic: str, scale_type: str) -> List[ScaleElement]:
-    """
-    Create scale.
+class Scale:
+    """A diatonic scale."""
 
-    :param tonic:
-        tonic pitch class represented by letter (like C or A#)
-    :param scale_type:
-        type of scale (currently, 'major', 'natural_minor', and
-        'harmonic_minor' are supported)
-    :return:
-        scale
-    """
-    patterns = {
-        'major': [1, 0, 2, 0, 3, 4, 0, 5, 0, 6, 0, 7],
-        'natural_minor': [1, 0, 2, 3, 0, 4, 0, 5, 6, 0, 7, 0],
-        'harmonic_minor': [1, 0, 2, 3, 0, 4, 0, 5, 6, 0, 0, 7],
-    }
-    pattern = patterns[scale_type]
-    tonic_position = NOTE_TO_POSITION[tonic + '1']
-    scale = []
-    for x in range(len(NOTE_TO_POSITION)):
-        degree = pattern[(x - tonic_position) % len(pattern)]
-        if degree > 0:
-            scale.append(ScaleElement(absolute_position=x, degree=degree))
-    return scale
+    def __init__(self, tonic: str, scale_type: str):
+        """
+        Initialize an instance.
 
+        :param tonic:
+            tonic pitch class represented by letter (like C or A#)
+        :param scale_type:
+            type of scale (currently, 'major', 'natural_minor', and
+            'harmonic_minor' are supported)
+        """
+        self.tonic = tonic
+        self.scale_type = scale_type
 
-def slice_scale(
-        scale: List[ScaleElement], lowest_note: str, highest_note: str
-) -> List[ScaleElement]:
-    """
-    Keep only pitches that are within a specified range.
+        self.elements = self.__create_elements()
+        self.note_to_element = {
+            element.note: element for element in self.elements
+        }
+        self.position_in_semitones_to_element = {
[…]

### 2020-04-18 rlmusician/environment/piece.py
feat: Start preparing Piece class to florid counterpoint

@@ -2,7 +2,7 @@
 Define data structure that represents musical piece compliant with some rules.
 
 This data structure contains two nested data structures:
-1) instrumental lines (lists of pitches),
+1) melodic lines (lists of pitches),
 2) piano roll (`numpy` 2D-array with rows corresponding to notes,
    columns corresponding to time steps, and cells containing zeros and ones
    and indicating whether a note is played).
@@ -12,7 +12,6 @@ Author: Nikolay Lysenko
 
 
 import datetime
-import itertools
 import os
 from typing import Any, Dict, List, NamedTuple
 
@@ -24,38 +23,36 @@ from rlmusician.environment.rules import (
     get_harmony_rules_registry
 )
 from rlmusician.utils import (
+    Scale,
+    ScaleElement,
     create_events_from_piece,
     create_midi_from_piece,
-    create_scale,
     create_wav_from_events,
-    slice_scale
 )
 
 
 NOTE_TO_POSITION = get_note_to_position_mapping()
-TONIC_TRIAD_DEGREES = (1, 3, 5)
+N_EIGHTS_PER_MEASURE = 8
 
 
 class LineElement(NamedTuple):
-    """A pitch that can be used within a line."""
+    """An element of a melodic line."""
 
-    absolute_position: int
-    relative_position: int
-    degree: int
-    is_from_tonic_triad: bool
-    feasible_movements: List[int]
+    scale_element: ScaleElement
+    start_time_in_eights: int
+    end_time_in_eights: int
 
 
 class Piece:
-    """Musical piece compliant with some rules of counterpoint writing."""
+    """Piece where florid counterpoint line is created given cantus firmus."""
 
     def __init__(
             self,
             tonic: str,
             scale_type: str,
             n_measures: int,
-            max_skip: int,
-            line_specifications: List[Dict[str, Any]],
+            cantus_firmus: List[str],
+            counterpoint_specifications: Dict[str, Any],
             voice_leading_rules: Dict[str, Any],
             harmony_rules: Dict[str, Any],
             rendering_params: Dict[str, Any]
@@ -70,126 +67,151 @@ class Piece:
             'harmonic_minor' are supported)
         :param n_measures:
             duration of piece in measures
-        :param max_skip:
-            maximum allowed skip (in scale degrees) between successive notes
-            from the same line
-        :param line_specifications:
-            parameters of lines
+        :param cantus_firmus:
+            cantus firmus as a sequence of notes
+        :param counterpoint_specifications:
[…]

### 2020-04-18 rlmusician/environment/piece.py
feat: Support rules of rhythm

@@ -19,8 +19,9 @@ import numpy as np
 from sinethesizer.io.utils import get_note_to_position_mapping
 
 from rlmusician.environment.rules import (
+    get_harmony_rules_registry,
+    get_rhythm_rules_registry,
     get_voice_leading_rules_registry,
-    get_harmony_rules_registry
 )
 from rlmusician.utils import (
     Scale,
@@ -53,6 +54,7 @@ class Piece:
             n_measures: int,
             cantus_firmus: List[str],
             counterpoint_specifications: Dict[str, Any],
+            rhythm_rules: Dict[str, Any],
             voice_leading_rules: Dict[str, Any],
             harmony_rules: Dict[str, Any],
             rendering_params: Dict[str, Any]
@@ -71,6 +73,8 @@ class Piece:
             cantus firmus as a sequence of notes
         :param counterpoint_specifications:
             parameters of a counterpoint line
+        :param rhythm_rules:
+            names of applicable rhythm rules and their parameters
         :param voice_leading_rules:
             names of applicable voice leading rules and their parameters
         :param harmony_rules:
@@ -82,6 +86,8 @@ class Piece:
         self.scale_type = scale_type
         self.n_measures = n_measures
         self.counterpoint_specifications = counterpoint_specifications
+        self.names_of_rhythm_rules = rhythm_rules['names']
+        self.rhythm_rules_params = rhythm_rules['params']
         self.names_of_voice_leading_rules = voice_leading_rules['names']
         self.voice_leading_rules_params = voice_leading_rules['params']
         self.names_of_harmony_rules = harmony_rules['names']
@@ -91,7 +97,10 @@ class Piece:
         self.scale = Scale(tonic, scale_type)
         self.max_skip = counterpoint_specifications['max_skip']
         self.all_movements = list(range(-self.max_skip, self.max_skip + 1))
-        self.current_time_in_eights = N_EIGHTS_PER_MEASURE
+        self.current_time_in_eights = None
+        self.current_measure_durations = None
+        self.past_movements = None
+        self.__set_defaults_to_runtime_variables()
 
         self.cantus_firmus = self.__create_cantus_firmus(cantus_firmus)
         self.counterpoint = self.__create_beginning_of_counterpoint()
@@ -100,11 +109,19 @@ class Piece:
         self.end_scale_element = self.scale.get_element_by_note(end_note)
         self.__validate_boundary_notes()
 
+        # TODO: Range for counterpoint line.
+
[…]

### 2020-04-18 rlmusician/environment/rules.py

@@ -1,5 +1,5 @@
 """
-Check compliance with some rules of voice leading and harmony.
+Check compliance with some rules of rhythm, voice leading, and harmony.
 
 Author: Nikolay Lysenko
 """
@@ -7,6 +7,46 @@ Author: Nikolay Lysenko
 
 import itertools
 from typing import Callable, Dict, List, Optional
+
+
+def check_validity_of_rhythmic_pattern(durations: List[int]) -> bool:
+    """
+    Check that current measure is properly divided by notes.
+
+    :param durations:
+        durations (in eights) of all notes from a current measure and
+        a new note
+    :return:
+        indicator whether a movement is in accordance with the rule
+    """
+    valid_patterns = [
+        [4, 4],
+        [4, 2, 2],
+        [4, 2, 1, 1],
+        [2, 2, 2, 2],
+        [2, 2, 2, 1, 1],
+        [2, 1, 1, 2, 2],
+        [4, 8],
+        [2, 2, 8],
+        [2, 1, 1, 8],
+    ]
+    for valid_pattern in valid_patterns:
+        if valid_pattern[:len(durations)] == durations:
+            return True
+    return False
+
+
+def get_rhythm_rules_registry() -> Dict[str, Callable]:
+    """
+    Get mapping from names to functions checking rhythm rules.
+
+    :return:
+        registry of functions checking rhythm rules
+    """
+    registry = {
+        'pattern_validity': check_validity_of_rhythmic_pattern,
+    }
+    return registry
 
 
 def check_stability_of_rearticulated_pitch(

### 2020-04-18 rlmusician/environment/piece.py
feat: Check range of countrpoint line

@@ -97,6 +97,7 @@ class Piece:
         self.scale = Scale(tonic, scale_type)
         self.max_skip = counterpoint_specifications['max_skip']
         self.all_movements = list(range(-self.max_skip, self.max_skip + 1))
+
         self.current_time_in_eights = None
         self.current_measure_durations = None
         self.past_movements = None
@@ -105,11 +106,16 @@ class Piece:
         self.cantus_firmus = self.__create_cantus_firmus(cantus_firmus)
         self.counterpoint = self.__create_beginning_of_counterpoint()
 
-        end_note = self.counterpoint_specifications['end_note']
+        end_note = counterpoint_specifications['end_note']
         self.end_scale_element = self.scale.get_element_by_note(end_note)
         self.__validate_boundary_notes()
 
-        # TODO: Range for counterpoint line.
+        self.lowest_element = self.scale.get_element_by_note(
+            counterpoint_specifications['lowest_note']
+        )
+        self.highest_element = self.scale.get_element_by_note(
+            counterpoint_specifications['highest_note']
+        )
 
         self._piano_roll = None
         self.__initialize_piano_roll()
@@ -200,14 +206,8 @@ class Piece:
         cantus_firmus_lower_bound = min(cantus_firmus_positions)
         cantus_firmus_upper_bound = max(cantus_firmus_positions)
 
-        lowest_element = self.scale.get_element_by_note(
-            self.counterpoint_specifications['lowest_note']
-        )
-        counterpoint_lower_bound = lowest_element.position_in_semitones
-        highest_element = self.scale.get_element_by_note(
-            self.counterpoint_specifications['highest_note']
-        )
-        counterpoint_upper_bound = highest_element.position_in_semitones
+        counterpoint_lower_bound = self.lowest_element.position_in_semitones
+        counterpoint_upper_bound = self.highest_element.position_in_semitones
 
         self.lowest_row_to_show = min(
             cantus_firmus_lower_bound,
@@ -218,30 +218,17 @@ class Piece:
             counterpoint_upper_bound
         )
 
-    def __finalize_if_needed(self) -> None:
-        """Add final measure of counterpoint line if the piece is finished."""
-        penultimate_measure_end = N_EIGHTS_PER_MEASURE * (self.n_measures - 1)
-        if self.current_time_in_eights == penultimate_measure_end:
-            end_line_element = LineElement(
-                self.end_scale_element,
-                penultimate_measure_end,
[…]

### 2020-04-18 rlmusician/environment/piece.py
feat: Support different timbres for different lines

@@ -373,7 +373,7 @@ class Piece:
         if self.current_time_in_eights == piece_duration:
             raise RuntimeError("Attempt to add notes to a finished piece.")
         if not self.check_rules(movement, duration):
-            raise ValueError('The suggested step breaks some rules.')
+            raise ValueError('The suggested continuation breaks some rules.')
         next_line_element = self.__find_next_element(movement, duration)
         self.counterpoint.append(next_line_element)
         self.__add_to_piano_roll(next_line_element)

### 2020-04-18 rlmusician/utils/io.py

@@ -14,10 +14,17 @@ from sinethesizer.io import (
 from sinethesizer.io.utils import get_list_of_notes
 
 
+N_EIGHTS_PER_MEASURE = 8
+
+
 def create_midi_from_piece(
         piece: 'rlmusician.environment.Piece',
-        midi_path: str, measure_in_seconds: float,
-        instrument: int, velocity: int, trailing_silence_in_measures: int = 2
+        midi_path: str,
+        measure_in_seconds: float,
+        cantus_firmus_instrument: int,
+        counterpoint_instrument: int,
+        velocity: int,
+        trailing_silence_in_measures: int = 2
 ) -> None:
     """
     Create MIDI file from a piece created by this package.
@@ -28,8 +35,12 @@ def create_midi_from_piece(
         path where resulting MIDI file is going to be saved
     :param measure_in_seconds:
         duration of one measure in seconds
-    :param instrument:
-        ID (number) of instrument according to General MIDI specification
+    :param cantus_firmus_instrument:
+        for an instrument that plays cantus firmus, its ID (number)
+        according to General MIDI specification
+    :param counterpoint_instrument:
+        for an instrument that plays counterpoint line, its ID (number)
+        according to General MIDI specification
     :param velocity:
         one common velocity for all notes
     :param trailing_silence_in_measures:
@@ -38,13 +49,26 @@ def create_midi_from_piece(
         None
     """
     numeration_shift = pretty_midi.note_name_to_number('A0')
-    pretty_midi_instrument = pretty_midi.Instrument(program=instrument)
-    for line in piece.lines:
-        for measure, element in enumerate(line):
-            if element is None:  # Dead end occurred during piece creation.
-                continue
-            start_time = measure * measure_in_seconds
-            end_time = start_time + measure_in_seconds
+    lines = [
+        piece.cantus_firmus,
+        piece.counterpoint
+    ]
+    pretty_midi_instruments = [
+        pretty_midi.Instrument(program=cantus_firmus_instrument),
+        pretty_midi.Instrument(program=counterpoint_instrument)
+    ]
+    for line, pretty_midi_instrument in zip(lines, pretty_midi_instruments):
+        for element in line:
+            start_time = (
+                element.start_time_in_eights
+                / N_EIGHTS_PER_MEASURE
+                * measure_in_seconds
+            )
+            end_time = (
+                element.end_time_in_eights
+                / N_EIGHTS_PER_MEASURE
[…]

### 2020-04-18 rlmusician/environment/piece.py
refac: Check all rules in a single function

@@ -2,10 +2,13 @@
 Define data structure that represents musical piece compliant with some rules.
 
 This data structure contains two nested data structures:
-1) melodic lines (lists of pitches),
+1) melodic lines (loosely speaking, lists of pitches),
 2) piano roll (`numpy` 2D-array with rows corresponding to notes,
    columns corresponding to time steps, and cells containing zeros and ones
    and indicating whether a note is played).
+
+The rules for a piece and its structure come from 5th species counterpoint
+(also known as florid counterpoint).
 
 Author: Nikolay Lysenko
 """
@@ -18,11 +21,7 @@ from typing import Any, Dict, List, NamedTuple
 import numpy as np
 from sinethesizer.io.utils import get_note_to_position_mapping
 
-from rlmusician.environment.rules import (
-    get_harmony_rules_registry,
-    get_rhythm_rules_registry,
-    get_voice_leading_rules_registry,
-)
+from rlmusician.environment.rules import get_rules_registry
 from rlmusician.utils import (
     Scale,
     ScaleElement,
@@ -54,9 +53,7 @@ class Piece:
             n_measures: int,
             cantus_firmus: List[str],
             counterpoint_specifications: Dict[str, Any],
-            rhythm_rules: Dict[str, Any],
-            voice_leading_rules: Dict[str, Any],
-            harmony_rules: Dict[str, Any],
+            rules: Dict[str, Any],
             rendering_params: Dict[str, Any]
     ):
         """
@@ -73,12 +70,8 @@ class Piece:
             cantus firmus as a sequence of notes
         :param counterpoint_specifications:
             parameters of a counterpoint line
-        :param rhythm_rules:
-            names of applicable rhythm rules and their parameters
-        :param voice_leading_rules:
-            names of applicable voice leading rules and their parameters
-        :param harmony_rules:
-            names of applicable harmony rules and their parameters
+        :param rules:
+            names of applicable rules and parameters of these rules
         :param rendering_params:
             settings of saving the piece to TSV, MIDI, and WAV files
         """
@@ -86,12 +79,8 @@ class Piece:
         self.scale_type = scale_type
         self.n_measures = n_measures
         self.counterpoint_specifications = counterpoint_specifications
-        self.names_of_rhythm_rules = rhythm_rules['names']
-        self.rhythm_rules_params = rhythm_rules['params']
-        self.names_of_voice_leading_rules = voice_leading_rules['names']
[…]

### 2020-04-18 rlmusician/environment/rules.py

@@ -14,8 +14,11 @@ def check_validity_of_rhythmic_pattern(durations: List[int]) -> bool:
     Check that current measure is properly divided by notes.
 
     :param durations:
-        durations (in eights) of all notes from a current measure and
-        a new note
+        durations (in eights) of all notes from a current measure
+        (including a new note); if a new note prolongs to the next measure,
+        its full duration is included; however, if the first note starts
+        in the previous measure, only its duration within the current measure
+        is included
     :return:
         indicator whether a movement is in accordance with the rule
     """
@@ -34,19 +37,6 @@ def check_validity_of_rhythmic_pattern(durations: List[int]) -> bool:
         if valid_pattern[:len(durations)] == durations:
             return True
     return False
-
-
-def get_rhythm_rules_registry() -> Dict[str, Callable]:
-    """
-    Get mapping from names to functions checking rhythm rules.
-
-    :return:
-        registry of functions checking rhythm rules
-    """
-    registry = {
-        'pattern_validity': check_validity_of_rhythmic_pattern,
-    }
-    return registry
 
 
 def check_stability_of_rearticulated_pitch(
@@ -184,23 +174,6 @@ def check_step_motion_to_final_pitch(
     return degrees_to_end_note <= measures_left
 
 
-def get_voice_leading_rules_registry() -> Dict[str, Callable]:
-    """
-    Get mapping from names to functions checking voice leading rules.
-
-    :return:
-        registry of functions checking voice leading rules
-    """
-    registry = {
-        'rearticulation': check_stability_of_rearticulated_pitch,
-        'destination_of_skip': check_that_skip_leads_to_stable_pitch,
-        'turn_after_skip': check_that_skip_is_followed_by_opposite_step_motion,
-        'VI_VII_resolution': check_resolution_of_submediant_and_leading_tone,
-        'step_motion_to_end': check_step_motion_to_final_pitch
-    }
-    return registry
-
-
 def check_consonance_of_sonority(sonority: List['LineElement']) -> bool:
     """
     Check that sonority is consonant.
@@ -261,14 +234,24 @@ def check_absence_of_pitch_class_clashes(
     return True
 
 
-def get_harmony_rules_registry() -> Dict[str, Callable]:
+def get_rules_registry() -> Dict[str, Callable]:
     """
-    Get mapping from names to functions checking harmony rules.
+    Get mapping from names to corresponding functions that check rules.
 
     :return:
[…]

### 2020-04-19 rlmusician/environment/piece.py
feat: Create states for faster rules checking

@@ -25,6 +25,7 @@ from rlmusician.environment.rules import get_rules_registry
 from rlmusician.utils import (
     Scale,
     ScaleElement,
+    check_consonance,
     create_events_from_piece,
     create_midi_from_piece,
     create_wav_from_events,
@@ -75,6 +76,7 @@ class Piece:
         :param rendering_params:
             settings of saving the piece to TSV, MIDI, and WAV files
         """
+        # Initial inputs.
         self.tonic = tonic
         self.scale_type = scale_type
         self.n_measures = n_measures
@@ -83,22 +85,21 @@ class Piece:
         self.rules_params = rules['params']
         self.rendering_params = rendering_params
 
+        # Calculated attributes.
         self.scale = Scale(tonic, scale_type)
         self.max_skip = counterpoint_specifications['max_skip']
         self.all_movements = list(range(-self.max_skip, self.max_skip + 1))
 
-        self.current_time_in_eights = None
-        self.current_measure_durations = None
-        self.past_movements = None
-        self.__set_defaults_to_runtime_variables()
-
+        # Melodic lines.
         self.cantus_firmus = self.__create_cantus_firmus(cantus_firmus)
         self.counterpoint = self.__create_beginning_of_counterpoint()
 
+        # Boundaries.
         end_note = counterpoint_specifications['end_note']
         self.end_scale_element = self.scale.get_element_by_note(end_note)
         self.__validate_boundary_notes()
 
+        # Vertical range of a counterpoint line.
         self.lowest_element = self.scale.get_element_by_note(
             counterpoint_specifications['lowest_note']
         )
@@ -106,17 +107,20 @@ class Piece:
             counterpoint_specifications['highest_note']
         )
 
+        # Piano roll.
         self._piano_roll = None
         self.__initialize_piano_roll()
         self.lowest_row_to_show = None
         self.highest_row_to_show = None
         self.__set_range_to_show()
 
-    def __set_defaults_to_runtime_variables(self) -> None:
-        """Set default values to variables that change at runtime."""
-        self.current_time_in_eights = N_EIGHTS_PER_MEASURE
-        self.current_measure_durations = []
-        self.past_movements = []
+        # Runtime variables.
+        self.current_time_in_eights = None
+        self.current_measure_durations = None
+        self.past_movements = None
+        self.current_motion_start_element = None
+        self.is_last_element_dissonant = None
[…]

### 2020-04-19 rlmusician/environment/rules.py

@@ -9,7 +9,7 @@ import itertools
 from typing import Callable, Dict, List, Optional
 
 
-def check_validity_of_rhythmic_pattern(durations: List[int]) -> bool:
+def check_validity_of_rhythmic_pattern(durations: List[int], **kwargs) -> bool:
     """
     Check that current measure is properly divided by notes.
 

### 2020-04-19 rlmusician/utils/__init__.py

@@ -11,12 +11,13 @@ from .io import (
     create_wav_from_events
 )
 from .misc import convert_to_base, map_in_parallel
-from .music_theory import Scale, ScaleElement
+from .music_theory import Scale, ScaleElement, check_consonance
 
 
 __all__ = [
     'Scale',
     'ScaleElement',
+    'check_consonance',
     'convert_to_base',
     'create_events_from_piece',
     'create_midi_from_piece',

### 2020-04-19 rlmusician/utils/music_theory.py

@@ -109,3 +109,39 @@ class Scale:
             raise ValueError(
                 f"Position {pos} is out of available positions range."
             )
+
+
+def check_consonance(
+        first: ScaleElement, second: ScaleElement,
+        is_perfect_fourth_consonant: bool = False
+) -> bool:
+    """
+    Check whether an interval between two pitches is consonant.
+
+    :param first:
+        first pitch
+    :param second:
+        second pitch
+    :param is_perfect_fourth_consonant:
+        indicator whether to consider perfect fourth a consonant interval
+        (for two voices, it usually considered dissonant)
+    :return:
+        indicator whether the interval is consonant
+    """
+    n_semitones_to_consonance = {
+        0: True,
+        1: False,
+        2: False,
+        3: True,
+        4: True,
+        5: is_perfect_fourth_consonant,
+        6: False,
+        7: True,
+        8: True,
+        9: True,
+        10: False,
+        11: False
+    }
+    interval = first.position_in_semitones - second.position_in_semitones
+    interval %= len(n_semitones_to_consonance)
+    return n_semitones_to_consonance[interval]

### 2020-05-06 rlmusician/agent/monte_carlo_beam_search.py
refac: Limit RAM consumption

@@ -6,11 +6,17 @@ Author: Nikolay Lysenko
 
 
 import random
-from copy import deepcopy
 from typing import Any, Dict, List, Optional, NamedTuple
 
 from rlmusician.environment import CounterpointEnv
-from rlmusician.utils import map_in_parallel
+from rlmusician.utils import generate_deep_copies, imap_in_parallel
+
+
+class EnvWithActions(NamedTuple):
+    """A tuple of `CounterpointEnv` and actions previously applied to it."""
+
+    env: CounterpointEnv
+    actions: List[int]
 
 
 class Record(NamedTuple):
@@ -20,7 +26,7 @@ class Record(NamedTuple):
     reward: float
 
 
-def roll_in(env: CounterpointEnv, actions: List[int]) -> CounterpointEnv:
+def roll_in(env: CounterpointEnv, actions: List[int]) -> EnvWithActions:
     """
     Do roll-in actions.
 
@@ -34,24 +40,22 @@ def roll_in(env: CounterpointEnv, actions: List[int]) -> CounterpointEnv:
     env.reset()
     for action in actions:
         env.step(action)
-    return env
+    env_with_actions = EnvWithActions(env, actions)
+    return env_with_actions
 
 
-def roll_out_randomly(
-        env: CounterpointEnv,
-        past_actions: List[int]
-) -> Record:
+def roll_out_randomly(env_with_actions: EnvWithActions) -> Record:
     """
     Continue an episode in progress with random actions until it is finished.
 
-    :param env:
-        environment
-    :param past_actions:
-        sequence of actions that have been taken before
+    :param env_with_actions:
+        environment and sequence of actions that have been taken before
     :return:
         finalized sequence of actions and reward for the episode
     """
     random.seed()  # Reseed to have independent results amongst processes.
+    env = env_with_actions.env
+    past_actions = env_with_actions.actions
     done = False
     valid_actions = env.valid_actions
     while not done:
@@ -89,11 +93,10 @@ def add_records(
         corresponding to them rewards
     """
     for stub in stubs:
-        env = roll_in(env, stub)
-        records_for_stub = map_in_parallel(
+        env_with_actions = roll_in(env, stub)
+        records_for_stub = imap_in_parallel(
             roll_out_randomly,
-            # FIXME: Excessive memory consumption happens here.
-            [(deepcopy(env), deepcopy(stub)) for _ in range(n_trials)],
+            generate_deep_copies(env_with_actions, n_trials),
             paralleling_params
         )
         records.extend(records_for_stub)

### 2020-05-06 rlmusician/utils/__init__.py

@@ -10,7 +10,7 @@ from .io import (
     create_midi_from_piece,
     create_wav_from_events
 )
-from .misc import convert_to_base, map_in_parallel
+from .misc import convert_to_base, imap_in_parallel, generate_deep_copies
 from .music_theory import Scale, ScaleElement, check_consonance
 
 
@@ -22,5 +22,6 @@ __all__ = [
     'create_events_from_piece',
     'create_midi_from_piece',
     'create_wav_from_events',
-    'map_in_parallel',
+    'imap_in_parallel',
+    'generate_deep_copies',
 ]

### 2020-05-06 rlmusician/utils/misc.py

@@ -6,47 +6,8 @@ Author: Nikolay Lysenko
 
 
 import multiprocessing as mp
-from typing import Any, Callable, Dict, List, Optional, Tuple
-
-
-def map_in_parallel(
-        fn: Callable,
-        args: List[Tuple[Any, ...]],
-        pool_kwargs: Optional[Dict[str, Any]] = None
-) -> List[Any]:
-    """
-    Apply function to each tuple of arguments from given list in parallel.
-
-    This function contains boilerplate code that is needed for correct work
-    of `pytest-cov`. Usage of `mp.Pool` as context manager is not alternative
-    to this function, because:
-    1) not all covered lines of code may be marked as covered;
-    2) some files with names like '.coverage.hostname.*' may be not deleted.
-
-    See more: https://github.com/pytest-dev/pytest-cov/issues/250.
-
-    :param fn:
-        function
-    :param args:
-        list of tuples of arguments
-    :param pool_kwargs:
-        parameters of pool such as number of processes and maximum number of
-        tasks for a worker before it is replaced with a new one
-    :return:
-        results of applying the function to the arguments
-    """
-    pool_kwargs = pool_kwargs or {}
-    pool_kwargs['processes'] = pool_kwargs.get('n_processes')
-    pool_kwargs['maxtasksperchild'] = pool_kwargs.get('max_tasks_per_child')
-    old_keys = ['n_processes', 'max_tasks_per_child']
-    pool_kwargs = {k: v for k, v in pool_kwargs.items() if k not in old_keys}
-    pool = mp.Pool(**pool_kwargs)
-    try:
-        results = pool.starmap(fn, args)
-    finally:
-        pool.close()
-        pool.join()
-    return results
+from copy import deepcopy
+from typing import Any, Callable, Dict, Iterator, List, Optional
 
 
 def convert_to_base(
@@ -76,3 +37,58 @@ def convert_to_base(
         digits.extend(padding)
     digits = digits[::-1]
     return digits
+
+
+def imap_in_parallel(
+        fn: Callable,
+        args: Iterator[Any],
+        pool_kwargs: Optional[Dict[str, Any]] = None
+) -> Iterator[Any]:
+    """
+    Apply function to each argument from given iterable in parallel.
+
+    This function contains boilerplate code that is needed for correct work
+    of `pytest-cov`. Usage of `mp.Pool` as context manager is not alternative
+    to this function, because:
+    1) not all covered lines of code may be marked as covered;
[…]

### 2020-05-06 rlmusician/environment/piece.py
feat: Prohibit overlapping motion

@@ -246,15 +246,19 @@ class Piece:
         )
         return next_line_element
 
-    def __find_cantus_firmus_elements(
-            self, duration: int
-    ) -> List[LineElement]:
+    def __find_cf_elements(self, duration: int) -> List[LineElement]:
         """Find what in cantus firmus sounds simultaneously with a new note."""
         start_index = self.current_time_in_eights // N_EIGHTS_PER_MEASURE
         end_time = self.current_time_in_eights + duration
         end_index = (end_time - 1) // N_EIGHTS_PER_MEASURE + 1
         results = self.cantus_firmus[start_index:end_index]
         return results
+
+    def __find_previous_cf_element(self) -> LineElement:
+        """Find what in cantus firmus sounds before a new note."""
+        index = (self.current_time_in_eights - 1) // N_EIGHTS_PER_MEASURE
+        result = self.cantus_firmus[index]
+        return result
 
     def __check_range(self, movement: int) -> bool:
         """Check that movement does not lead beyond a range of a line."""
@@ -275,7 +279,8 @@ class Piece:
         registry = get_rules_registry()
         continuation = self.__find_next_element(movement, duration)
         durations = [x for x in self.current_measure_durations] + [duration]
-        cantus_firmus_elements = self.__find_cantus_firmus_elements(duration)
+        cantus_firmus_elements = self.__find_cf_elements(duration)
+        previous_cantus_firmus_element = self.__find_previous_cf_element()
         state = {
             'line': self.counterpoint,
             'counterpoint_continuation': continuation,
@@ -285,6 +290,7 @@ class Piece:
             'current_measure_durations': self.current_measure_durations,
             'durations': durations,
             'cantus_firmus_elements': cantus_firmus_elements,
+            'previous_cantus_firmus_element': previous_cantus_firmus_element,
             'current_motion_start_element': self.current_motion_start_element,
             'is_last_element_consonant': self.is_last_element_consonant,
             'is_counterpoint_above': self.is_counterpoint_above,
@@ -339,7 +345,7 @@ class Piece:
 
     def __update_indicator_of_consonance(self, duration: int) -> None:
         """Update indicator of current vertical consonance between lines."""
-        cantus_firmus_elements = self.__find_cantus_firmus_elements(duration)
+        cantus_firmus_elements = self.__find_cf_elements(duration)
         cantus_firmus_element = cantus_firmus_elements[-1].scale_element
         counterpoint_element = self.counterpoint[-1].scale_element
         self.is_last_element_consonant = check_consonance(

### 2020-05-06 rlmusician/environment/rules.py

@@ -415,6 +415,31 @@ def check_absence_of_lines_crossing(
     return True
 
 
+def check_absence_of_overlapping_motion(
+        counterpoint_continuation: 'LineElement',
+        previous_cantus_firmus_element: 'LineElement',
+        is_counterpoint_above: bool,
+        **kwargs
+) -> bool:
+    """
+    Check that there is no overlapping motion.
+
+    :param counterpoint_continuation:
+        current continuation of counterpoint line
+    :param previous_cantus_firmus_element:
+        the latest element of cantus firmus that sounds simultaneously
+        with the last counterpoint element (excluding its continuation)
+    :param is_counterpoint_above:
+        indicator whether counterpoint must be above cantus firmus
+    :return:
+        indicator whether a continuation is in accordance with the rule
+    """
+    initial_sign = 1 if is_counterpoint_above else -1
+    cpt_pitch = counterpoint_continuation.scale_element.position_in_semitones
+    cf_pitch = previous_cantus_firmus_element.scale_element.position_in_semitones
+    return initial_sign * (cpt_pitch - cf_pitch) > 0
+
+
 # Registry.
 
 def get_rules_registry() -> Dict[str, Callable]:
@@ -443,5 +468,6 @@ def get_rules_registry() -> Dict[str, Callable]:
         'resolution_of_suspended_dissonance': check_resolution_of_suspended_dissonance,
         'absence_of_large_intervals': check_absence_of_large_intervals,
         'absence_of_lines_crossing': check_absence_of_lines_crossing,
+        'absence_of_overlapping_motion': check_absence_of_overlapping_motion,
     }
     return registry

### 2020-05-06 tests/environment/test_rules.py

@@ -14,6 +14,7 @@ from rlmusician.environment.rules import (
     check_absence_of_large_intervals,
     check_absence_of_lines_crossing,
     check_absence_of_monotonous_long_motion,
+    check_absence_of_overlapping_motion,
     check_absence_of_skip_series,
     check_absence_of_stalled_pitches,
     check_consonance_on_strong_beat,
@@ -129,6 +130,49 @@ def test_check_absence_of_monotonous_long_motion(
         counterpoint_continuation,
         current_motion_start_element,
         max_distance_in_semitones
+    )
+    assert result == expected
+
+
+@pytest.mark.parametrize(
+    "counterpoint_continuation, previous_cantus_firmus_element, "
+    "is_counterpoint_above, expected",
+    [
+        (
+            LineElement(ScaleElement('C1', 3, 2, 1, True), 24, 28),
+            LineElement(ScaleElement('C2', 15, 9, 1, True), 16, 24),
+            True,
+            False
+        ),
+        (
+            LineElement(ScaleElement('C1', 3, 2, 1, True), 24, 28),
+            LineElement(ScaleElement('C2', 15, 9, 1, True), 16, 24),
+            False,
+            True
+        ),
+        (
+            LineElement(ScaleElement('C1', 3, 2, 1, True), 24, 28),
+            LineElement(ScaleElement('C1', 3, 2, 1, True), 16, 24),
+            True,
+            False
+        ),
+        (
+            LineElement(ScaleElement('C1', 3, 2, 1, True), 24, 28),
+            LineElement(ScaleElement('C1', 3, 2, 1, True), 16, 24),
+            False,
+            False
+        ),
+    ]
+)
+def test_check_absence_of_overlapping_motion(
+        counterpoint_continuation: LineElement,
+        previous_cantus_firmus_element: LineElement,
+        is_counterpoint_above: bool, expected: bool
+) -> None:
+    """Test `check_absence_of_overlapping_motion` function."""
+    result = check_absence_of_overlapping_motion(
+        counterpoint_continuation, previous_cantus_firmus_element,
+        is_counterpoint_above
     )
     assert result == expected
 

### 2020-05-07 rlmusician/environment/evaluation.py
feat: Evaluate absence of narrow ranges

@@ -12,6 +12,7 @@ import numpy as np
 from scipy.stats import entropy
 
 from rlmusician.environment.piece import Piece
+from rlmusician.utils import rolling_aggregate
 
 
 def evaluate_absence_of_looped_fragments(
@@ -65,6 +66,37 @@ def evaluate_entropy(piece: Piece) -> float:
     max_entropy_distribution = [1 / len(elements) for _ in elements]
     denominator = entropy(max_entropy_distribution)
     score = raw_score / denominator
+    return score
+
+
+def evaluate_absence_of_narrow_ranges(
+        piece: Piece, min_size: int = 9,
+        penalties: Optional[Dict[int, float]] = None
+) -> float:
+    """
+    Evaluate melodic fluency based on absence of narrow ranges.
+
+    :param piece:
+        `Piece` instance
+    :param min_size:
+        minimum size of narrow range (in line elements)
+    :param penalties:
+        mapping from width of a range (in scale degrees) to penalty
+        applicable to ranges of not greater width
+    :return:
+        multiplied by -1 count of narrow ranges weighted based on their width
+    """
+    penalties = penalties or {2: 1, 3: 0.5}
+    pitches = [x.scale_element.position_in_degrees for x in piece.counterpoint]
+    rolling_mins = rolling_aggregate(pitches, min, min_size)[min_size-1:]
+    rolling_maxs = rolling_aggregate(pitches, max, min_size)[min_size-1:]
+    borders = zip(rolling_mins, rolling_maxs)
+    score = 0
+    for lower_border, upper_border in borders:
+        range_width = upper_border - lower_border
+        curr_penalties = [v for k, v in penalties.items() if k >= range_width]
+        penalty = max(curr_penalties) if curr_penalties else 0
+        score -= penalty
     return score
 
 
@@ -136,6 +168,7 @@ def get_scoring_functions_registry() -> Dict[str, Callable]:
     registry = {
         'looped_fragments': evaluate_absence_of_looped_fragments,
         'entropy': evaluate_entropy,
+        'narrow_ranges': evaluate_absence_of_narrow_ranges,
         'climax_explicity': evaluate_climax_explicity,
         'number_of_skips': evaluate_number_of_skips,
     }

### 2020-05-07 rlmusician/utils/__init__.py

@@ -8,9 +8,14 @@ Author: Nikolay Lysenko
 from .io import (
     create_events_from_piece,
     create_midi_from_piece,
-    create_wav_from_events
+    create_wav_from_events,
 )
-from .misc import convert_to_base, imap_in_parallel, generate_deep_copies
+from .misc import (
+    convert_to_base,
+    imap_in_parallel,
+    generate_deep_copies,
+    rolling_aggregate,
+)
 from .music_theory import Scale, ScaleElement, check_consonance
 
 
@@ -24,4 +29,5 @@ __all__ = [
     'create_wav_from_events',
     'imap_in_parallel',
     'generate_deep_copies',
+    'rolling_aggregate',
 ]

### 2020-05-07 rlmusician/utils/misc.py

@@ -86,9 +86,36 @@ def generate_deep_copies(something: Any, n_copies: int) -> Iterator[Any]:
     :param something:
         object to be copied
     :param n_copies:
-        requested number of copies
+        number of copies to be generated
     :return:
         deep copies
     """
     for _ in range(n_copies):
         yield deepcopy(something)
+
+
+def rolling_aggregate(
+        values: List[float],
+        aggregation_fn: Callable[[List[float]], float],
+        window_size: int
+) -> List[float]:
+    """
+    Compute rolling aggregate.
+
+    :param values:
+        list of values to be aggregated
+    :param aggregation_fn:
+        aggregation function
+    :param window_size:
+        size of rolling window
+    :return:
+        list of rolling aggregates
+    """
+    window = []
+    results = []
+    for value in values:
+        if len(window) == window_size:
+            window.pop(0)
+        window.append(value)
+        results.append(aggregation_fn(window))
+    return results

### 2020-05-07 tests/environment/test_evaluation.py

@@ -5,12 +5,13 @@ Author: Nikolay Lysenko
 """
 
 
-from typing import List, Optional, Tuple
+from typing import Dict, List, Optional, Tuple
 
 import pytest
 
 from rlmusician.environment.evaluation import (
     evaluate_absence_of_looped_fragments,
+    evaluate_absence_of_narrow_ranges,
     evaluate_climax_explicity,
     evaluate_entropy,
     evaluate_number_of_skips,
@@ -89,6 +90,109 @@ def test_evaluate_absence_of_looped_fragments(
     for movement, duration in steps:
         piece.add_line_element(movement, duration)
     result = evaluate_absence_of_looped_fragments(piece, min_size, max_size)
+    assert result == expected
+
+
+@pytest.mark.parametrize(
+    "piece, steps, min_size, penalties, expected",
+    [
+        (
+            # `piece`
+            Piece(
+                tonic='C',
+                scale_type='major',
+                cantus_firmus=['C4', 'D4', 'E4', 'D4', 'C4'],
+                counterpoint_specifications={
+                    'start_note': 'E4',
+                    'end_note': 'G3',
+                    'lowest_note': 'G3',
+                    'highest_note': 'G4',
+                    'start_pause_in_eights': 4,
+                    'max_skip_in_degrees': 3,
+                },
+                rules={
+                    'names': ['rearticulation_stability'],
+                    'params': {}
+                },
+                rendering_params={}
+            ),
+            # `steps`,
+            [(1, 4), (1, 4), (-1, 4), (-1, 4), (-3, 4), (-1, 4)],
+            # `min_size`
+            4,
+            # `penalties`
+            {1: 1, 2: 0.6, 3: 0.1},
+            # `expected`
+            -1.2
+        ),
+        (
+            # `piece`
+            Piece(
+                tonic='C',
+                scale_type='major',
+                cantus_firmus=['C4', 'D4', 'E4', 'D4', 'C4'],
+                counterpoint_specifications={
+                    'start_note': 'E4',
+                    'end_note': 'G3',
+                    'lowest_note': 'G3',
+                    'highest_note': 'G4',
+                    'start_pause_in_eights': 4,
+                    'max_skip_in_degrees': 3,
+                },
+                rules={
+                    'names': ['rearticulation_stability'],
+                    'params': {}
+                },
+                rendering_params={}
+            ),
+            # `steps`,
[…]

### 2020-05-07 tests/utils/test_misc.py

@@ -5,11 +5,11 @@ Author: Nikolay Lysenko
 """
 
 
-from typing import List, Optional
+from typing import Callable, List, Optional
 
 import pytest
 
-from rlmusician.utils.misc import convert_to_base
+from rlmusician.utils.misc import convert_to_base, rolling_aggregate
 
 
 @pytest.mark.parametrize(
@@ -26,4 +26,19 @@ def test_convert_to_base(
 ) -> None:
     """Test `convert_to_base` function."""
     result = convert_to_base(number, base, min_length)
+    assert result == expected
+
+
+@pytest.mark.parametrize(
+    "values, aggregation_fn, window_size, expected",
+    [
+        ([0, 5, 2, 1, -3, 6, 4, 7], min, 3, [0, 0, 0, 1, -3, -3, -3, 4]),
+    ]
+)
+def test_rolling_aggregate(
+        values: List[float], aggregation_fn: Callable[[List[float]], float],
+        window_size: int, expected: List[float]
+) -> None:
+    """Test `rolling_aggregate` function."""
+    result = rolling_aggregate(values, aggregation_fn, window_size)
     assert result == expected

### 2020-05-08 docs/paper/paper.tex
docs: Update first three sections of the draft

@@
 
-\date{April 14, 2020}
+\date{May 11, 2020}
 
@@
 \begin{abstract}
-In this paper, automatic composition of first species counterpoint is framed as a tree search problem close to reinforcement learning.
-An environment representing a polyphonic piece in progress is introduced.
-An action is filling of the current measure (bar) for all melodic lines simultaneously.
-Once a specified number of measures is added, an episode terminates and the resulting piece is evaluated based on some hand-written rules derived from music theory.
-Best sequences of actions are found with a variation of Monte Carlo Beam Search.
+In this paper, automatic composition of fifth species counterpoint is framed as a tree search problem close to reinforcement learning.
+The goal in this problem is to compose a melodic line given another melodic line consisting of whole notes only.
+The line in progress has fixed ends and the rest is filled sequentially from start to finish.
+Once it is done, an episode terminates and the resulting piece is evaluated based on some hand-written rules derived from music theory.
+Best sequences of notes are found with a variation of Monte Carlo Beam Search.
 
@@
 The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern.
-Here, first species counterpoint is covered.
-This type has trivial rhythmic structure such that only whole (semibreve) notes can be used and every note must be placed in a beginning of a measure.
-Thus, interaction between rhythm and tonality is eliminated and the problem becomes even simpler.
+Here, fifth species counterpoint is covered.
+This type is the highest one and the less restrictive one.
+Like in other species, there is a given melodic line consisting entirely of whole (semibreve) notes.
+This line is called cantus firmus.
+Again, like in other species, a new melodic line of the same duration is going to be created (so called counterpoint line) and these two lines should sound aesthetically pleasant when played simultaneously.
+The difference between fifth species and other species is that a counterpoint line may change its rhythmic pattern: half notes, quarter notes, eights notes, and suspended whole notes can be used.
 
 Brief outline of the setup is as follows.
-There is an environment that keeps a piece in progress and the piece consists of a fixed number of measures.
-Each measure must be either empty (if it is not filled yet) or containing exactly one note per a melodic line.
-Initially, only the first measure and the last measure are filled with user-defined values and the current measure is the second one.
-An action is filling of the current measure, i.e., adding a pitch to each of the melodic lines.
-After an action, the next measure becomes the current one.
-When the last measure is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.
+There is an environment that keeps a piece in progress.
+Initially, the piece consists of a user-defined cantus firmus, the first note of counterpoint line, and the last note of counterpoint line.
+An action is extension of the counterpoint left part by one new note.
+Not every action is valid, there are rules prohibiting some of them.
+Invalid actions can not be done by design.
+When the last note is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.
 
@@
 On the one hand, reward function is a domain-specific name for fitness function.
-On the other hand, data generated during past episodes are used for training supervised models.
+On the other hand, data generated during past episodes are used for training supervised models that approximate value function.
 
 The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group.
-A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function.
+A local search algorithm named Variable Neighborhood Search (VNS) is used there in order to create first and fifth species counterpoint pieces that maximize fitness function.
 What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint.
-However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes.
-Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations.
-As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces cannot be created.
-This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results cannot be reproduced.
+Despite that, there are important differences:
+\begin{itemize}
+	\item In \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations.
+As a result, there are only evaluation rules and there are no rules of composition.
+If a piece breaks some fundamental rules of counterpoint, it is still evaluated (but it can receive score making it unfeasible).
+Conversely, in the current study, both rules of composition and rules of evaluation are present as separate entities.
+An action that contradicts a rule of composition can not be done and so only compliant pieces are evaluated.
+	\item In \cite{herremans2012first, herremans2013fifth}, evaluation rules accurately follow counterpoint theory, but here it is not reproduced too literally.
+From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sound unpleasant, because some of these prescriptions are style-specific.
+	\item Output of VNS is a fixed piece.
+If cantus firmus is changed, VNS should be run from scratch.
+Actually, output of Monte Carlo Beam Search is a fixed piece too and search should be run from scratch for every new cantus firmus as well.
+However, the current approach can be upgraded to a true reinforcement learning approach.
+Such upgrade is possible, because the problem is formulated in RL terms: composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions updates.
+If so, there can be an agent that observes current state of a piece and writes counterpoint line based on observations.
+Cantus firmus can be changed from episode to episode during training of the agent in order to teach the agent to deal with various cantus firmi.
+After each episode played by the trained agent, a new output piece can be created even for a fixed cantus firmus.
+	\item Methodology of the two discussed papers has a software implementation named Optimuse, but download links are broken and so results can not be reproduced.
+Unlike this, links from the present paper lead to time-tested sites such as GitHub and PyPI.
+\end{itemize}
 
@@
 It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning.
-As for the first case, probably, there are no influential works and only separate studies like \cite{yi2007automatic, smith2012reinforcement} are available.
-Conversely, the second group recently gained attention and several interesting approaches were developed there.
+As for the first case, some separate studies like \cite{yi2007automatic, smith2012reinforcement} are available, but the second group recently gained attention and several interesting approaches were developed there.
 
@@
 
-The described in Section \ref{sec:introduction} procedure of first species counterpoint writing can be considered a tree search problem.
+The described in Section \ref{sec:introduction} procedure of fifth species counterpoint writing can be considered a tree search problem.
 The root is the initial state of an environment.
-Each valid option to fill the current measure is an edge from the current node of the tree to a new node.
-Leaves are pieces where all measures are filled.
+Each valid option to extend counterpoint line is an edge from the current node of the tree to a new node.
+Leaves are pieces where counterpoint line is finished.
 
@@
 Unfortunately, trees are large in numerous real problems.
-For instance, a rough estimate of total number of leaves for first species counterpoint with 2 voices and 16 measures is about $7^{14}$ which is equal to several hundreds of billions.
 

### 2020-05-12 docs/paper/paper.tex
docs: Update section about problem setup

@@
 Usually, softmax activation function is used in the last layer of generative RNN.
-However, it is a common practice in reinforcement learning domain to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$.
+However, it is a common practice in reinforcement learning domain to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted by $t$.
 Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$).
@@
 There are well-known classical algorithms such as depth-first search (DFS) that allow iterating over all paths from the root to leaves.
-Unfortunately, trees are large in numerous real problems.
+Unfortunately, trees are large in numerous real-world problems.
 
@@
 To start with, define representation of a musical piece.
-Suppose that counterpoint is written in $l$ melodic lines (parts, voices), where $l$ is usually 2 or 3.
-Suppose also that a diatonic scale (like C-major or A-minor) is chosen.
-Let each line be a list of pitches from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited).
-Set length of all lines to the same number $m$, where $m$ is usually between 8 and 24.
+Suppose that a diatonic scale (like C-major or A-minor) is chosen.
+Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eights of measure).
+In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint.
+Definition of cantus firmus implies that very element of cantus firmus has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus.
+Denote this length by $m$, where $m$ is usually between 8 and 24.
 From music point of view, this means that a piece to be created is rather a phrase that can be inserted into a longer composition.
 Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.
-This way or that, define a piece as a matrix $P$ of size $l \times m$ where element $P_{ij}$ is a pitch played in $i$-th line in $j$-th measure.
 
 Every line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds.
-Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 3$).
-If so, there are no more than $(2s + 1)^l$ options to fill the next measure given current measure.
+Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 7$).
+If so, there are no more than $(2s + 1)$ options to select the next pitch given the current pitch and there are no more than $4(2s + 1)$ options to add the next note given the current note (where $4 = \#\{1, 2, 4, 8\}$).
 Actually, the number is smaller, because some options are filtered out by counterpoint rules.
-The rules enforced in this study can be either rules of voice leading (part writing) or rules of harmony.
+The rules enforced in this study can be rules of rhythm, rules of voice leading (part writing), and rules of harmony.
 
-Here, five rules of voice leading are imposed:
+Here, only one rule of rhythm is present:
+\begin{itemize}
+	\item There must be no conflicts between strong beats and longer notes.
+In other words, every measure must be split in a manner that allows a listener to unambiguously define meter.
+For example, if a measure is split like this: (4, 2, 1, 1) or (2, 1, 1, 8) where 8 stands for a tied over bar note, this is correct, but if a measure is split like this: (2, 4, 2), this is not correct, because downbeat note is shorter than the next note and intermediate strong beat is inside a note, not at a beginning of a note.
+\end{itemize}
+Further, seven rules of voice leading are imposed:
 \begin{itemize}
 	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
+	\item No pitches can be rearticulated two times in a row, because it makes a line stalling;
+	\item A line can not move from an intermediate pitch by more than 9 semitones without any changes in direction (rearticulation is not a change of direction);
 	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
-	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
+	\item There can be no more than 2 skips (leaps) in a row;
 	\item Movement by step in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
@@
 \end{itemize}
-In addition, there are three rules of harmony:
+Finally, there are seven rules of harmony:
 \begin{itemize}
-	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
-	\item Within a measure, all pitch classes must be different unless the measure is the first one or the last one;
-	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
+	\item If a note from a counterpoint line starts on a strong beat (downbeat or measure middle), harmonic interval between the pitch and simultaneously sounding pitch from cantus firmus, must be consonant (although, if counterpoint note is suspended, it can be dissonant with the second simultaneously sounding note from cantus firmus);
+	\item If a note from a counterpoint line forms a dissonance with cantus firmus, it must be reached by step motion;
+	\item If a note from a counterpoint line forms a dissonance with cantus firmus, it must be left by step motion;
+	\item If a dissonance is created by suspended note, this dissonance must be resolved by downward step motion;
+	\item Harmonic intervals larger than tenth are prohibited;
+	\item Voice crossing is prohibited (i.e., counterpoint can be either strictly above cantus firmus or strictly below it);
+	\item Overlapping motion is prohibited (i.e., counterpoint can be either strictly above lagged cantus firmus or strictly below it).
 \end{itemize}
 
-Above rules are absolute and it is not possible to break any of them, because pitches for the next measure can be selected only from options that are filtered according to these rules.
+Above rules are absolute and it is not possible to break any of them, because line continuations can be selected only from options that are filtered according to these rules.
 If a set of such options is empty, episode terminates and an unfinished sequence of actions receives negative reward for dead end.
+
+It worths to be mentioned that the listed rules are based on counterpoint theory, but they do not reproduce it exactly.
+For example, there is a pattern named Double Neighbor and it is allowed in counterpoint music, but it is prohibited by the rule about step motion to a dissonating element.
 

### 2020-05-15 rlmusician/environment/evaluation.py
refac: Penalize measures with one-pitch only

@@ -16,7 +16,7 @@ from rlmusician.utils import rolling_aggregate
 
 
 def evaluate_absence_of_looped_fragments(
-        piece: Piece, min_size: int = 8, max_size: Optional[int] = None
+        piece: Piece, min_size: int = 4, max_size: Optional[int] = None
 ) -> float:
     """
     Evaluate non-triviality of a piece based on absence of looped fragments.
@@ -33,7 +33,10 @@ def evaluate_absence_of_looped_fragments(
     score = 0
     max_size = max_size or piece.total_duration_in_eights // 2
     for size in range(min_size, max_size + 1):
-        for position in range(0, piece.total_duration_in_eights - 2*size + 1):
+        max_position = piece.total_duration_in_eights - 2 * size
+        penultimate_measure_end = piece.total_duration_in_eights - 8 - 1
+        max_position = min(max_position, penultimate_measure_end)
+        for position in range(0, max_position + 1):
             fragment = piece.piano_roll[:, position:position+size]
             next_fragment = piece.piano_roll[:, position+size:position+2*size]
             if np.array_equal(fragment, next_fragment):

### 2020-05-16 rlmusician/environment/evaluation.py
feat: Grant continuous reward for number of skips

@@ -139,25 +139,25 @@ def evaluate_climax_explicity(
 
 
 def evaluate_number_of_skips(
-        piece: Piece, min_n_skips: int = 1, max_n_skips: int = 3
+        piece: Piece, rewards: Optional[Dict[int, float]] = None
 ) -> float:
     """
-    Evaluate interestingness of counterpoint line based on number of skips.
+    Evaluate interestingness/coherency of counterpoint based on skips number.
 
     :param piece:
         `Piece` instance
-    :param min_n_skips:
-        minimum number of skips for a line to be interesting
-    :param max_n_skips:
-        maximum number of skips for a line to be still coherent
+    :param rewards:
+        mapping from number of skips to reward
     :return:
-        indicator whether number of skips lies within a specified range
+        reward assigned to balancing between interestingess and coherency
+        of counterpoint line
     """
+    rewards = rewards or {1: 0.8, 2: 0.9, 3: 1, 4: 0.9, 5: 0.5, 6: 0.25}
     n_skips = 0
     for movement in piece.past_movements:
         if abs(movement) > 1:
             n_skips += 1
-    score = 1 if min_n_skips <= n_skips <= max_n_skips else 0
+    score = rewards.get(n_skips, 0)
     return score
 
 

### 2020-05-16 tests/environment/test_evaluation.py

@@ -338,7 +338,7 @@ def test_evaluate_entropy(
 
 
 @pytest.mark.parametrize(
-    "piece, steps, min_n_skips, max_n_skips, expected",
+    "piece, steps, rewards, expected",
     [
         (
             # `piece`
@@ -362,21 +362,19 @@ def test_evaluate_entropy(
             ),
             # `steps`,
             [(1, 4), (1, 4), (-3, 4), (-1, 4), (-1, 4), (-1, 4)],
-            # `min_n_skips`
-            1,
-            # `max_n_skips`
-            2,
+            # `rewards`
+            {1: 0.5, 2: 1, 3: 0.5},
             # `expected`
             1
         ),
     ]
 )
 def test_evaluate_number_of_skips(
-        piece: Piece, steps: List[Tuple[int, int]],
-        min_n_skips: int, max_n_skips: int, expected: float
+        piece: Piece, steps: List[Tuple[int, int]], rewards: Dict[int, float],
+        expected: float
 ) -> None:
     """Test `evaluate_number_of_skips` function."""
     for movement, duration in steps:
         piece.add_line_element(movement, duration)
-    result = evaluate_number_of_skips(piece, min_n_skips, max_n_skips)
+    result = evaluate_number_of_skips(piece, rewards)
     assert result == expected

### 2020-05-16 docs/paper/paper.tex
docs: Update the rest of the draft

@@
 \usepackage{algorithmic}    % pseudocode
+\usepackage{graphicx}       % images
 
@@
 
-\date{May 11, 2020}
+\date{May 16, 2020}
 
@@
 The goal in this problem is to compose a melodic line given another melodic line consisting of whole notes only.
-The line in progress has fixed ends and the rest is filled sequentially from start to finish.
+The line in progress has fixed ends and the gap between these ends is filled sequentially from start to finish.
 Once it is done, an episode terminates and the resulting piece is evaluated based on some hand-written rules derived from music theory.
@@
 	\item In \cite{herremans2012first, herremans2013fifth}, evaluation rules accurately follow counterpoint theory, but here it is not reproduced too literally.
-From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sound unpleasant, because some of these prescriptions are style-specific.
+From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sounds unpleasant, because some of these prescriptions are style-specific.
 	\item Output of VNS is a fixed piece.
@@
 In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint.
-Definition of cantus firmus implies that very element of cantus firmus has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus.
+Definition of cantus firmus implies that every its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus.
 Denote this length by $m$, where $m$ is usually between 8 and 24.
@@
 Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.
+
+Alternatively, a piece can be represented by a piano roll which is a matrix $P$ of size $n \times 8m$ where $n$ is size of available pitches range (hence, $n \le 88$, since there are 88 piano keys) and $8m$ is number of eights of measure in a piece lasting $m$ measures.
+Matrix element $P_{ij}$ is 1 if $i$-th pitch is sounding at $j$-th eight and 0 otherwise.
 
@@
 Actually, the number is smaller, because some options are filtered out by counterpoint rules.
-The rules enforced in this study can be rules of rhythm, rules of voice leading (part writing), and rules of harmony.
+The rules enforced in this study can be rules of rhythm, rules of voice leading, and rules of harmony.
 
@@
 In other words, every measure must be split in a manner that allows a listener to unambiguously define meter.
-For example, if a measure is split like this: (4, 2, 1, 1) or (2, 1, 1, 8) where 8 stands for a tied over bar note, this is correct, but if a measure is split like this: (2, 4, 2), this is not correct, because downbeat note is shorter than the next note and intermediate strong beat is inside a note, not at a beginning of a note.
+For example, if a measure is split like this: (4, 2, 1, 1) or (2, 1, 1, 8) where 8 stands for a tied over bar note, this is correct, but if a measure is split like this: (2, 4, 2), this is not correct, because downbeat note is shorter than the next note and intermediate strong beat is inside a span of a note, not at a beginning of a note.
 \end{itemize}
@@
 	\item Movement by step in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
-	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
+	\item For every note from counterpoint line, there must be a way to reach the final note of this line with step motion (remember that the final note is set initially), because resolution must be present during last measures.
 \end{itemize}
@@
 
+Initially, cantus firmus, start note of counterpoint line (occupying the whole first measure except opening pause), end note of counterpoint line (occupying the whole last measure), and range of pitches available for usage in counterpoint line, are set externally.
+These values are the same for all episodes of Monte Carlo Beam Search.
+During each episode, a gap between the first note and the last note is reduced by one new note added to the left.
+This is done until the counterpoint line is finished.
+
 \subsection{Evaluational Rules}
@@
 
-At the current revision of the study, seven properties of a final piece are evaluated separately and total reward is a weighted sum of these seven scores.
+At the current revision of the study, five properties of a final piece are evaluated separately and total reward is a weighted sum of these five scores.
 
 \begin{itemize}
-	\item \textbf{Absence of looped pitches.} A stable pitch may be rearticulated, but not more than once in a row.
-Every excessive occurrence of the same pitch within a line adds constant negative reward.
-    \item \textbf{Absence of looped fragments.} Define a fragment as one or more consecutive measures (at a piece level, not at a melodic line level).
+    \item \textbf{Absence of looped fragments.} Define a fragment as content of four or more consecutive columns of a piano roll $P$.
 For every fragment followed by exactly the same fragment, a constant negative reward is added.
-	\item \textbf{Motion.} For every pair of lines every motion between two successive measures adds a reward that depends on a type of motion.
-Since contrary motion makes lines more distinguishable, reward for it is positive.
-Reward for oblique motion is zero, reward for similar motion is slightly negative and reward for parallel motion is negative, because parallel motion makes lines sounding like copies of each other.
-	\item \textbf{Correlation between lines.} Consider each line as a sequence of real numbers equal to pitch positions within a range of available pitches.
-Then average over all pairs of lines correlation between two lines can quantify independence of lines.
-Actually, reward is rescaled to be from 0 to 1 -- it is half of difference of 1 and the above average correlation.
+    \item \textbf{Entropy of pitch distribution.} Let $C$ be a set of all pitches available for usage in a counterpoint line and let $\mu$ be the probability distribution on this set $C$ derived from frequencies of pitches in a counterpoint line.
+Divide entropy of $\mu$ by entropy of the uniform distribution on $C$.
+The result is a real number from 0 to 1.
+This number is returned as a reward for non-degeneracy of pitch usage.
     \item \textbf{Explicity of climax.} Music theory suggests that good melodies must have goal-oriented motion and that such goal-orientedness may be achieved by having exactly one climax point.
-For each line, constant reward is decreased for each duplication of climax point and for not so high climax point.
-    \item \textbf{Number of skips.} Positive reward is associated with every line such that number of skips in it lies within a specified range.
+For a counterpoint line, a constant reward is decreased for each duplication of climax point and for not so high climax point.
+    \item \textbf{Absence of aimless fluctuations.} If melody fluctuates within a narrow range of pitches for a long time, it sounds boring.
+So for every 9 or more successive notes from a counterpoint line such that the lowest of them and the highest of them are no more than third apart, constant negative reward is assigned.
+Also, if an interval between the lowest of 9 or more successive notes and the highest of them is a fourth, half of this negative reward is assigned.
+    \item \textbf{Number of skips.} Depending on number of skips in counterpoint line, non-negative reward is granted.
 The motivation behind this is that lines with almost no skips are not interesting, but lines with too many skips are not coherent.
-    \item \textbf{Absence of downward skips}.
-Counterpoint theory prescribes to have downward step motion.
-To encourage its usage, there are negative rewards for every downward skip and the larger a skip is, the higher a penalty its.
+So the reward is 0 when number of skips is 0, then it increases, reaches maximum at 3 skips and decreases after that becoming 0 again for 7 or more skips.
 \end{itemize}
 
-It can be seen that the first two properties deal with non-triviality of results, the next two properties deal with interaction of lines and the last three properties deal with melodic quality of each separate line.
+It can be seen that the first two properties deal with non-triviality of results, while the other three properties deal with melodic quality of a counterpoint line.
 
@@
 A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}.
-The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.3.0/}}.
+The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.4.1/}}.
 All important settings are placed to a separate configuration file and so it is easy to experiment with them.
@@
 
-In this section, results are reported for experiments with 2 melodic lines and 16 measures starting from (G4, C5) and ending with (C4, C5).
-Properties to be evaluated are weighted so that explicity of climax has 3 times more impact than any other property.
+In this section, results are reported for experiments with a cantus firmus attributed to Johann Joseph Fux, a theorist of species counterpoint.
+This cantus firmus lasts 12 measures.
+Counterpoint line starts with G4, ends with C5, and is allowed to have pitches from C4 to B5.
+Properties to be evaluated have equal weights.
+
+Below figure shows a piece produced given this setup. As Table \ref{table:scores} shows, the piece is close to theoretical optimum.
+
+\begin{figure}[h!]
+	\label{image:sheet}
+	\caption{An example of a generated piece.}
+	\centering
+	\includegraphics[width=0.9\textwidth]{sheet_music.png}
+\end{figure}
 
 \begin{table}[h!]
-	\caption{Scores of sample pieces}
+	\caption{Scores of the above piece.}
 	\label{table:scores}
 	\begin{center}
-		\begin{tabular}{|l|c|c|c|c|}
+		\begin{tabular}{|l|c|c|}
 			\hline
-			& Human & MCS & MCBS & Maximum\\
+			& Score & Maximum\\
 			\hline
-			Absence of looped pitches & 0 & 0 & 0 & 0\\
+			Absence of looped fragments & 0 & 0\\
 			\hline
-			Absence of looped fragments & -1 & 0 & 0 & 0\\
+			Entropy of pitch distribution & 0.868 & 1\\
 			\hline
-			Motion & -0.267 & -0.1 & -0.033 & 1\\
+			Explicity of climax & 1 & 1\\
 			\hline
-			Correlation between lines & 0.269 & 0.321 & 0.321 & 1\\
+			Absence of aimless fluctuations & 0 & 0\\
 			\hline
-			Explicity of climax & 3 & 3 & 3 & 3\\
+			Number of skips & 1 & 1\\
 			\hline
-			Number of skips & 1 & 1 & 1 & 1\\
-			\hline
-			Absence of downward skips & 0 & -0.6 & 0 & 0\\
-			\hline
-			Total reward & 3.003 & 3.621 & 4.288 & 6\\
+			\textbf{Total reward} & 2.868 & 3\\
 			\hline
@@
 
-In Table \ref{table:scores}, scores of a piece found with Monte Carlo Beam Search are compared with benchmarks.
-The piece was found with beam width set to 5 and number of trials schedule set to ($10000 \times 3$, $5000 \times 3$, $1000 \times 8$) -- this means that the second measure of the piece was chosen after 10000 random continuations and, for example, the eighth measure of the piece was chosen after 1000 random continuations.
-One of the benchmarks is achieved by Monte Carlo Search with 53000 random trials (i.e., total number of random trials is the same).
-The other benchmark relates to a piece manually created by the author in about 10 minutes.
-Ideally, scores should be collected during multiple independent experiments and then statistical significance of difference should be tested, but it is not done yet.
+The piece was found with beam width set to 5 and number of trials schedule set to ($30000$, $10000 \times 5$, $5000 \times 3$, $1000 \times 5$, $100 \times \dots$) -- this means that the second note of the counterpoint line was chosen after 30000 random continuations and, for example, the eighth note of the counterpoint line was chosen after 5000 random continuations.
 
-The results look like scores for climax explicity, absence of looped pitches, absence of looped fragments, and absence of downward skips impose hard constraints.
-If so, the goal is to maximize sum of scores for lines independence subject to counterpoint rules and above new constraints.
-
-Table \ref{table:piece} demonstrates found by Monte Carlo Beam Search piece for which the scores are reported.
-
-\begin{table}[h!]
-	\caption{The piece found with Monte Carlo Beam Search}
-	\label{table:piece}
-	\begin{center}
-		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
-			\hline
-			& & & C6 & & & & & & & & & & & & \\
-			\hline
-			& & & & B5 & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & A5 & & & &  & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & G5 & & & & G5 & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & F5 & & & & & & & & \\
-			\hline
-			& E5 & & & & & & & E5 & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & D5 & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			C5 & C5 & & & & & & & & & C5 & & C5 & C5 & & C5 \\
-			\hline
-			& & B4 & & & & & & & & & B4 & & & B4 & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & A4 & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			G4 & & & & G4 & & & & G4 & & G4 & G4 & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & F4 & & & & F4 & & & F4 & & & \\
-			\hline
-			& & & & & & E4 & & & & & & & E4 & & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & D4 & & & & & & & D4 & \\
-			\hline
-			& & & & & & & & & & & & & & & \\
-			\hline
-			& & & & & & & & & & & & & & & C4 \\
-			\hline
-		\end{tabular}
-	\end{center}
-\end{table}
-
-Subjective evaluation of music quality can be used for revising setup of the problem. Probably, 16 measures are too much, because long step motions are resembling descending scales.
+Multiple runs of Monte Carlo Beam Search were held and almost all of them led to scores similar to the reported above scores.
+Pieces found by different runs are not variations of each other, but they may share something in common.
+Taking into account that cantus firmus is the same for all of them, this is expected.
+Moreover, for the chosen cantus firmus and the chosen settings, climax pitch of a counterpoint line can be B5 (the highest allowed pitch) only in the eighth measure and this limitation obviously reduces variability.
 
@@
 This paper introduces a new approach to composition of music without usage of existing pieces.
-For a problem of first species counterpoint writing, the approach finds pieces that maximize weighted sum of associated with particular musical properties scores subject to boundary constraints.
-These boundary constraints (the first measure and the last measure) are user-defined and weights for the sum of scores are user-defined too.
+For a problem of fifth species counterpoint writing, the approach finds pieces that maximize weighted sum of associated with particular musical properties scores subject to boundary constraints.
+These boundary constraints (the first note and the last note) are user-defined and weights for the sum of scores are user-defined too.
 
-Results are good from formal metrics point of view -- the method outperforms random search and manual trials.
-However, created pieces are too simple to be enjoyable music.
-This is an inevitable consequence of too strict limitations of first species counterpoint.
-To eliminate them, higher species can be chosen or frameworks other than species counterpoint can be explored.
+Because the problem is a reduced version of general music composition problem, results can be virtually perfect from formal metrics point of view.
+However, several aspects of real music are ignored here:
+\begin{itemize}
+	\item \textbf{Long-term structure.} The developed method produces short phrases (up to a pair of dozens of measures). When creating a longer piece, new concepts arise: say, motifs.
+	\item \textbf{Higher number of voices.} In most Western pieces, four voices are used. Usually, one of them forms main melody and the others form accompaniment.
+	\item \textbf{Freer textures and rhythmical patterns.} Presence of a melody entirely consisting of whole notes is not a mandatory requirement.
+Further, pauses can be used within lines.
+Finally, the higher number of voices is, the more rhythmical patterns become available.
+\end{itemize}
 
-If the problem becomes computationally complex after introducing more musical freedom, concepts from reinforcement learning such as value function can be involved.
-State-free search can be replaced with a search where decision are made based on observations and estimated values of states and actions.
+It is impossible to take these aspects into account while staying within species counterpoint. A framework going beyond it should be involved.
 

### 2020-05-16 tests/environment/test_environment.py
ci: Update some tests

@@ -42,7 +42,7 @@ class TestCounterpointEnv:
                     ),
                     scoring_coefs={'number_of_skips': 1},
                     scoring_fn_params={
-                        'number_of_skips': {'min_n_skips': 1, 'max_n_skips': 1}
+                        'number_of_skips': {'rewards': {1: 1}}
                     },
                     reward_for_dead_end=-100,
                 ),
@@ -108,7 +108,7 @@ class TestCounterpointEnv:
                     ),
                     scoring_coefs={'number_of_skips': 1},
                     scoring_fn_params={
-                        'number_of_skips': {'min_n_skips': 1, 'max_n_skips': 1}
+                        'number_of_skips': {'rewards': {1: 1}}
                     },
                     reward_for_dead_end=-100,
                 ),
@@ -155,7 +155,7 @@ class TestCounterpointEnv:
                     ),
                     scoring_coefs={'number_of_skips': 1},
                     scoring_fn_params={
-                        'number_of_skips': {'min_n_skips': 1, 'max_n_skips': 1}
+                        'number_of_skips': {'rewards': {1: 1}}
                     },
                     reward_for_dead_end=-100,
                 ),
@@ -187,7 +187,7 @@ class TestCounterpointEnv:
                     ),
                     scoring_coefs={'number_of_skips': 1},
                     scoring_fn_params={
-                        'number_of_skips': {'min_n_skips': 4, 'max_n_skips': 4}
+                        'number_of_skips': {'rewards': {4: 1}}
                     },
                     reward_for_dead_end=-100,
                 ),
@@ -223,7 +223,7 @@ class TestCounterpointEnv:
                     ),
                     scoring_coefs={'number_of_skips': 1},
                     scoring_fn_params={
-                        'number_of_skips': {'min_n_skips': 1, 'max_n_skips': 1}
+                        'number_of_skips': {'rewards': {1: 1}}
                     },
                     reward_for_dead_end=-100,
                 ),
@@ -269,7 +269,7 @@ class TestCounterpointEnv:
                     ),
                     scoring_coefs={'number_of_skips': 1},
                     scoring_fn_params={
-                        'number_of_skips': {'min_n_skips': 1, 'max_n_skips': 1}
+                        'number_of_skips': {'rewards': {1: 1}}
                     },
                     reward_for_dead_end=-100,
                 ),

### 2020-05-17 rlmusician/environment/environment.py
chore: Correct typo in a word

@@ -109,8 +109,8 @@ class CounterpointEnv(gym.Env):
         observation = self.piece.piano_roll
         info = {'next_actions': self.valid_actions}
 
-        past_duration = self.piece.current_time_in_eights
-        piece_duration = self.piece.total_duration_in_eights
+        past_duration = self.piece.current_time_in_eighths
+        piece_duration = self.piece.total_duration_in_eighths
         finished = past_duration == piece_duration
         no_more_actions = len(info['next_actions']) == 0
         done = finished or no_more_actions

### 2020-05-17 rlmusician/environment/evaluation.py

@@ -24,18 +24,18 @@ def evaluate_absence_of_looped_fragments(
     :param piece:
         `Piece` instance
     :param min_size:
-        minimum duration of a fragment (in eights)
+        minimum duration of a fragment (in eighths)
     :param max_size:
-        maximum duration of a fragment (in eights)
+        maximum duration of a fragment (in eighths)
     :return:
         multiplied by -1 number of looped fragments
     """
     score = 0
-    max_size = max_size or piece.total_duration_in_eights // 2
+    max_size = max_size or piece.total_duration_in_eighths // 2
     for size in range(min_size, max_size + 1):
-        max_position = piece.total_duration_in_eights - 2 * size
-        penultimate_measure_end = piece.total_duration_in_eights - 8 - 1
-        max_position = min(max_position, penultimate_measure_end)
+        max_position = piece.total_duration_in_eighths - 2 * size
+        penultimate_measure_end = piece.total_duration_in_eighths - 8
+        max_position = min(max_position, penultimate_measure_end - 1)
         for position in range(0, max_position + 1):
             fragment = piece.piano_roll[:, position:position+size]
             next_fragment = piece.piano_roll[:, position+size:position+2*size]

### 2020-05-17 rlmusician/environment/piece.py

@@ -33,15 +33,15 @@ from rlmusician.utils import (
 
 
 NOTE_TO_POSITION = get_note_to_position_mapping()
-N_EIGHTS_PER_MEASURE = 8
+N_EIGHTHS_PER_MEASURE = 8
 
 
 class LineElement(NamedTuple):
     """An element of a melodic line."""
 
     scale_element: ScaleElement
-    start_time_in_eights: int
-    end_time_in_eights: int
+    start_time_in_eighths: int
+    end_time_in_eighths: int
 
 
 class Piece:
@@ -86,7 +86,9 @@ class Piece:
         self.max_skip = counterpoint_specifications['max_skip_in_degrees']
         self.all_movements = list(range(-self.max_skip, self.max_skip + 1))
         self.n_measures = len(cantus_firmus)
-        self.total_duration_in_eights = N_EIGHTS_PER_MEASURE * self.n_measures
+        self.total_duration_in_eighths = (
+            N_EIGHTHS_PER_MEASURE * self.n_measures
+        )
 
         # Melodic lines.
         self.cantus_firmus = self.__create_cantus_firmus(cantus_firmus)
@@ -115,7 +117,7 @@ class Piece:
         self.__set_range_to_show()
 
         # Runtime variables.
-        self.current_time_in_eights = None
+        self.current_time_in_eighths = None
         self.current_measure_durations = None
         self.past_movements = None
         self.current_motion_start_element = None
@@ -129,8 +131,8 @@ class Piece:
         cantus_firmus = [
             LineElement(
                 scale_element=self.scale.get_element_by_note(note),
-                start_time_in_eights=N_EIGHTS_PER_MEASURE * i,
-                end_time_in_eights=N_EIGHTS_PER_MEASURE * (i+1)
+                start_time_in_eighths=N_EIGHTHS_PER_MEASURE * i,
+                end_time_in_eighths=N_EIGHTHS_PER_MEASURE * (i+1)
             )
             for i, note in enumerate(cantus_firmus_as_notes)
         ]
@@ -141,8 +143,8 @@ class Piece:
         start_note = self.counterpoint_specifications['start_note']
         start_element = LineElement(
             self.scale.get_element_by_note(start_note),
-            self.counterpoint_specifications['start_pause_in_eights'],
-            N_EIGHTS_PER_MEASURE
+            self.counterpoint_specifications['start_pause_in_eighths'],
+            N_EIGHTHS_PER_MEASURE
         )
[…]

### 2020-05-17 rlmusician/environment/rules.py

@@ -11,7 +11,7 @@ from typing import Callable, Dict, List
 from rlmusician.utils.music_theory import ScaleElement, check_consonance
 
 
-N_EIGHTS_PER_MEASURE = 8
+N_EIGHTHS_PER_MEASURE = 8
 
 
 # Rhythm rules.
@@ -21,7 +21,7 @@ def check_validity_of_rhythmic_pattern(durations: List[int], **kwargs) -> bool:
     Check that current measure is properly divided by notes.
 
     :param durations:
-        durations (in eights) of all notes from a current measure
+        durations (in eighths) of all notes from a current measure
         (including a new note); if a new note prolongs to the next measure,
         its full duration is included; however, if the first note starts
         in the previous measure, only its duration within the current measure
@@ -216,7 +216,7 @@ def check_step_motion_to_final_pitch(
     :param counterpoint_end:
         element that ends counterpoint line
     :param piece_duration:
-        total duration of piece (in eights)
+        total duration of piece (in eighths)
     :param prohibit_rearticulation:
         if it is set to `True`, the last but one pitch can not be the same as
         the final pitch
@@ -227,11 +227,11 @@ def check_step_motion_to_final_pitch(
         counterpoint_continuation.scale_element.position_in_degrees
         - counterpoint_end.position_in_degrees
     )
-    eights_left = (
-        (piece_duration - N_EIGHTS_PER_MEASURE)
-        - counterpoint_continuation.end_time_in_eights
+    eighths_left = (
+        (piece_duration - N_EIGHTHS_PER_MEASURE)
+        - counterpoint_continuation.end_time_in_eighths
     )
-    quarters_left = ceil(eights_left / 2)
+    quarters_left = ceil(eighths_left / 2)
     if quarters_left == 0 and degrees_to_end_note == 0:
         return not prohibit_rearticulation
     return degrees_to_end_note <= quarters_left + 1
@@ -255,7 +255,7 @@ def check_consonance_on_strong_beat(
     :return:
         indicator whether a continuation is in accordance with the rule
     """
-    if counterpoint_continuation.start_time_in_eights % 4 != 0:
+    if counterpoint_continuation.start_time_in_eighths % 4 != 0:
         return True
     return check_consonance(
         counterpoint_continuation.scale_element,
@@ -340,8 +340,10 @@ def check_resolution_of_suspended_dissonance(
     :return:
         indicator whether a continuation is in accordance with the rule
     """
[…]

### 2020-05-17 rlmusician/utils/io.py

@@ -14,7 +14,7 @@ from sinethesizer.io import (
 from sinethesizer.io.utils import get_list_of_notes
 
 
-N_EIGHTS_PER_MEASURE = 8
+N_EIGHTHS_PER_MEASURE = 8
 
 
 def create_midi_from_piece(
@@ -64,13 +64,13 @@ def create_midi_from_piece(
                 + numeration_shift
             )
             start_time = (
-                element.start_time_in_eights
-                / N_EIGHTS_PER_MEASURE
+                element.start_time_in_eighths
+                / N_EIGHTHS_PER_MEASURE
                 * measure_in_seconds
             )
             end_time = (
-                element.end_time_in_eights
-                / N_EIGHTS_PER_MEASURE
+                element.end_time_in_eighths
+                / N_EIGHTHS_PER_MEASURE
                 * measure_in_seconds
             )
             note = pretty_midi.Note(
@@ -137,13 +137,13 @@ def create_events_from_piece(
     for line, timbre in zip(lines, timbres):
         for element in line:
             start_time = (
-                element.start_time_in_eights
-                / N_EIGHTS_PER_MEASURE
+                element.start_time_in_eighths
+                / N_EIGHTHS_PER_MEASURE
                 * measure_in_seconds
             )
             duration = (
-                (element.end_time_in_eights - element.start_time_in_eights)
-                / N_EIGHTS_PER_MEASURE
+                (element.end_time_in_eighths - element.start_time_in_eighths)
+                / N_EIGHTHS_PER_MEASURE
                 * measure_in_seconds
             )
             pitch_id = element.scale_element.position_in_semitones

### 2020-05-17 tests/agent/test_monte_carlo_beam_search.py

@@ -103,7 +103,7 @@ def test_create_stubs(
                         'end_note': 'E4',
                         'lowest_note': 'G3',
                         'highest_note': 'G4',
-                        'start_pause_in_eights': 4,
+                        'start_pause_in_eighths': 4,
                         'max_skip_in_degrees': 2,
                     },
                     rules={

### 2020-05-17 tests/environment/test_environment.py

@@ -31,7 +31,7 @@ class TestCounterpointEnv:
                             'end_note': 'E4',
                             'lowest_note': 'G3',
                             'highest_note': 'G4',
-                            'start_pause_in_eights': 4,
+                            'start_pause_in_eighths': 4,
                             'max_skip_in_degrees': 2,
                         },
                         rules={
@@ -92,7 +92,7 @@ class TestCounterpointEnv:
                             'end_note': 'E4',
                             'lowest_note': 'G3',
                             'highest_note': 'G4',
-                            'start_pause_in_eights': 4,
+                            'start_pause_in_eighths': 4,
                             'max_skip_in_degrees': 2,
                         },
                         rules={
@@ -144,7 +144,7 @@ class TestCounterpointEnv:
                             'end_note': 'E4',
                             'lowest_note': 'G3',
                             'highest_note': 'G4',
-                            'start_pause_in_eights': 4,
+                            'start_pause_in_eighths': 4,
                             'max_skip_in_degrees': 2,
                         },
                         rules={
@@ -176,7 +176,7 @@ class TestCounterpointEnv:
                             'end_note': 'E4',
                             'lowest_note': 'G3',
                             'highest_note': 'G4',
-                            'start_pause_in_eights': 4,
+                            'start_pause_in_eighths': 4,
                             'max_skip_in_degrees': 2,
                         },
                         rules={
@@ -208,7 +208,7 @@ class TestCounterpointEnv:
                             'end_note': 'E4',
                             'lowest_note': 'G3',
                             'highest_note': 'G4',
-                            'start_pause_in_eights': 4,
+                            'start_pause_in_eighths': 4,
                             'max_skip_in_degrees': 2,
                         },
                         rules={
@@ -258,7 +258,7 @@ class TestCounterpointEnv:
                             'end_note': 'E4',
                             'lowest_note': 'G3',
                             'highest_note': 'G4',
-                            'start_pause_in_eights': 4,
+                            'start_pause_in_eighths': 4,
                             'max_skip_in_degrees': 2,
                         },
                         rules={
@@ -303,4 +303,4 @@ class TestCounterpointEnv:
             env.step(action)
         observation = env.reset()
         np.testing.assert_equal(observation, expected)
-        assert env.piece.current_time_in_eights == 8
+        assert env.piece.current_time_in_eighths == 8

### 2020-05-17 tests/environment/test_evaluation.py

@@ -33,7 +33,7 @@ from rlmusician.environment.piece import Piece
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 rules={
@@ -62,7 +62,7 @@ from rlmusician.environment.piece import Piece
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 rules={
@@ -107,7 +107,7 @@ def test_evaluate_absence_of_looped_fragments(
                     'end_note': 'G3',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 3,
                 },
                 rules={
@@ -136,7 +136,7 @@ def test_evaluate_absence_of_looped_fragments(
                     'end_note': 'G3',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 3,
                 },
                 rules={
@@ -165,7 +165,7 @@ def test_evaluate_absence_of_looped_fragments(
                     'end_note': 'G3',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 3,
                 },
                 rules={
@@ -210,7 +210,7 @@ def test_evaluate_absence_of_narrow_ranges(
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'E4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 rules={
@@ -239,7 +239,7 @@ def test_evaluate_absence_of_narrow_ranges(
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
[…]

### 2020-05-17 tests/environment/test_piece.py

@@ -35,7 +35,7 @@ class TestPiece:
                     'end_note': 'E4',
                     'lowest_note': 'G4',
                     'highest_note': 'G3',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -59,7 +59,7 @@ class TestPiece:
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -83,7 +83,7 @@ class TestPiece:
                     'end_note': 'D4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -107,7 +107,7 @@ class TestPiece:
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -131,7 +131,7 @@ class TestPiece:
                     'end_note': 'E#4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -173,7 +173,7 @@ class TestPiece:
                     'highest_note': 'G4',
                     'start_note': 'F#4',
                     'end_note': 'D4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -234,7 +234,7 @@ class TestPiece:
                     'end_note': 'E4',
                     'lowest_note': 'G3',
                     'highest_note': 'G4',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2,
                 },
                 # `rules`
@@ -262,7 +262,7 @@ class TestPiece:
                     'end_note': 'E4',
[…]

### 2020-05-17 tests/utils/test_io.py

@@ -30,7 +30,7 @@ from rlmusician.utils import (
                     'end_note': 'C5',
                     'lowest_note': 'C4',
                     'highest_note': 'C6',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2
                 },
                 rules={
@@ -90,7 +90,7 @@ def test_create_midi_from_piece(
                     'end_note': 'C5',
                     'lowest_note': 'C4',
                     'highest_note': 'C6',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2
                 },
                 rules={
@@ -121,7 +121,7 @@ def test_create_midi_from_piece(
                     'end_note': 'C5',
                     'lowest_note': 'C4',
                     'highest_note': 'C6',
-                    'start_pause_in_eights': 4,
+                    'start_pause_in_eighths': 4,
                     'max_skip_in_degrees': 2
                 },
                 rules={

### 2020-05-17 docs/paper/paper.tex
docs: Revise the draft of the paper

@@
 Again, like in other species, a new melodic line of the same duration is going to be created (so called counterpoint line) and these two lines should sound aesthetically pleasant when played simultaneously.
-The difference between fifth species and other species is that a counterpoint line may change its rhythmic pattern: half notes, quarter notes, eights notes, and suspended whole notes can be used.
+The difference between fifth species and other species is that a counterpoint line may change its rhythmic pattern: half notes, quarter notes, eighths notes, and suspended whole notes can be used.
 
@@
 Suppose that a diatonic scale (like C-major or A-minor) is chosen.
-Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eights of measure).
+Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eighths of measure).
 In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint.
@@
 
-Alternatively, a piece can be represented by a piano roll which is a matrix $P$ of size $n \times 8m$ where $n$ is size of available pitches range (hence, $n \le 88$, since there are 88 piano keys) and $8m$ is number of eights of measure in a piece lasting $m$ measures.
-Matrix element $P_{ij}$ is 1 if $i$-th pitch is sounding at $j$-th eight and 0 otherwise.
+Alternatively, a piece can be represented by a piano roll which is a matrix $P$ of size $n \times 8m$ where $n$ is size of available pitches range (hence, $n \le 88$, since there are 88 piano keys) and $8m$ is number of eighths of measure in a piece lasting $m$ measures.
+Matrix element $P_{ij}$ is 1 if $i$-th pitch is sounding at $j$-th eighth and 0 otherwise.
 
@@
 These values are the same for all episodes of Monte Carlo Beam Search.
-During each episode, a gap between the first note and the last note is reduced by one new note added to the left.
-This is done until the counterpoint line is finished.
+During each episode, a gap between the first note and the last note is reduced by one new note added to the left and this continues until the counterpoint line is finished.
 
@@
 
-Below figure shows a piece produced given this setup. As Table \ref{table:scores} shows, the piece is close to theoretical optimum.
+Below figure demonstrates a piece produced with this setup. As Table \ref{table:scores} shows, the piece is close to theoretical optimum.
 

### 2020-05-24 docs/paper/paper.tex
docs: Tweak minor issues

@@
 Not every action is valid, there are rules prohibiting some of them.
-Invalid actions can not be done by design.
-When the last note is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.
+Invalid actions cannot be done by design.
+When the last note is reached, the episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.
 
@@
 Conversely, in the current study, both rules of composition and rules of evaluation are present as separate entities.
-An action that contradicts a rule of composition can not be done and so only compliant pieces are evaluated.
+An action that contradicts a rule of composition cannot be done and so only compliant pieces are evaluated.
 	\item In \cite{herremans2012first, herremans2013fifth}, evaluation rules accurately follow counterpoint theory, but here it is not reproduced too literally.
-From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sounds unpleasant, because some of these prescriptions are style-specific.
+From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sound unpleasant, because some of these prescriptions are style-specific.
 	\item Output of VNS is a fixed piece.
@@
 After each episode played by the trained agent, a new output piece can be created even for a fixed cantus firmus.
-	\item Methodology of the two discussed papers has a software implementation named Optimuse, but download links are broken and so results can not be reproduced.
+	\item Methodology of the two discussed papers has a software implementation named Optimuse, but download links are broken and so results cannot be reproduced.
 Unlike this, links from the present paper lead to time-tested sites such as GitHub and PyPI.
@@
 In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint.
-Definition of cantus firmus implies that every its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus.
+Definition of cantus firmus implies that each its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus.
 Denote this length by $m$, where $m$ is usually between 8 and 24.
@@
 	\item No pitches can be rearticulated two times in a row, because it makes a line stalling;
-	\item A line can not move from an intermediate pitch by more than 9 semitones without any changes in direction (rearticulation is not a change of direction);
+	\item A line cannot move from an intermediate pitch by more than 9 semitones without any changes in direction (rearticulation is not a change of direction);
 	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;

### 2020-05-26 setup.py
chore: Update package version

@@ -17,7 +17,7 @@ with open(os.path.join(current_dir, 'README.md'), encoding='utf-8') as f:
 
 setup(
     name='rl-musician',
-    version='0.4.0',
+    version='0.4.1',
     description=description,
     long_description=long_description,
     long_description_content_type='text/markdown',
